### 2018-03-11 hw1/hw1.tex
initial commit

\documentclass{report} 
\usepackage{geometry}
\geometry{margin=1in}

\usepackage{pgfplots}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{xcolor}

\title{STAT 672: Homework \#1}
\author{Tom Wallace}

\begin{document}

\maketitle

\section*{Problem 1}

\subsection*{A}

See submitted code \texttt{hw1\_1A.py}.

The acceptance probability is the volume of an $n$-dimensional ball divided by
the volume of an $n$-dimensional cube:

\begin{equation}
P_{X \sim B^d_\infty}(||X||_2 \leq 1) =
\frac{\mathrm{Vol}(B^d_2)}{\mathrm{Vol}(B^2_\infty)}
= \frac{\frac{\pi ^ {d/2}}{\Gamma(\frac{d}{2} + 1)}}{2^d}
\end{equation}

We know that $\Gamma (n+1) = n!$ Thus, $\Gamma (\frac{d}{2} + 1) = (\frac{d}{2})!$ Stirling's 
approximation for factorials is useful here.

\begin{equation}
n! \approx \sqrt{2 \pi n}n^ne^{-n}
\end{equation}

Let us consider the case of $2d$. In this case:

\begin{equation}
\mathrm{Vol}(B^{2d}_2) = \frac{\pi ^ d}{\Gamma(d + 1)} = \frac{\pi^d}{d!}
\approx \frac{\pi^d}{\sqrt{2\pi d}d^de^{-d}}
\end{equation}

Rearranging terms, we have:

\begin{equation}
= \frac{1}{\sqrt{2\pi d}} \left( \frac{\pi e}{d} \right)^d
\end{equation}

Referring to $ \left( \frac{\pi e}{d} \right)^d$, the denominator grows faster with $d$ than does the
numerator. As a consequence, if we extend $d$ infinitely, the limit of the ratio is 0.

\begin{equation}
\lim_{d \to \infty} \left( \frac{\pi e}{d} \right)^d = 0
\end{equation}

Which implies:
\begin{equation}
\lim_{d \to \infty} \frac{1}{\sqrt{2\pi d}} \left( \frac{\pi e}{d} \right)^d =
	\frac{1}{\infty} \times 0 = 0
\end{equation}

Returning to (1), we compute the the volume of the unit cube as $d \to \infty$.

\begin{equation}
\lim_{d \to \infty} 2^{2d} = \infty
\end{equation}

So, the ratio of the unit sphere to the unit cube, i.e. the probability of
acceptance, is:

\begin{equation}
\lim_{d \to \infty} \frac{\mathrm{Vol}(B^d_2)}{\mathrm{Vol}(B^2_\infty)} =
\frac{0}{\infty} = 0
\end{equation}

This finding has severe consequences for our expected run time. As $d$
increases, the probability of accepting a random vector becomes increasingly
low, meaning that we must conduct many draws to generate a single random vector,
meaning that our runtime greatly increases. For example, with $d=10,000$, the
program might require weeks to generate a modest number of vectors.

\subsection*{B}
See submitted code \texttt{hw1\_1B.py}.

We can generate $\Theta$ using the
formula $Z / ||Z||_2$, where $Z \sim \mathcal{N}(0,
\mathbf{I}_d)$. Let us examine why. The distribution of
$Z$ is invariant to rotations about the origin. That is, for any orthogonal matrix
$\mathbf{Q}$, $\mathbf{Q}Z \sim \mathcal{N}(0, \mathbf{I}_d)$. Proof: for any
$Y \sim \mathcal{N}(\mathbf{\mu}, \mathbf{\Sigma})$, and  matrix $\mathbf{C}_{p
\times n}$ with rank $p$, $\mathbf{C}Y \sim \mathcal{N}_p (\mathbf{C\mu}, \mathbf{C\Sigma
C'})$. So, for any orthogonal matrix $\mathbf{Q}$, $\mathbf{Q}Y \sim
\mathcal{N}(\mathbf{Q\mu}, \mathbf{Q\Sigma Q'})$. In the case of $Z$,
$\mu = 0$ and $\mathbf{\Sigma} = \mathbf{I}$, and so $\mathbf{QZ} \sim
\mathcal{N}(0\mathbf{Q}, \mathbf{QIQ'})$. Obviously, $0\mathbf{Q}=0$, and by the
respective definitions of identity and orthogonal matrices,
$\mathbf{QIQ}'=\mathbf{QQ}'=\mathbf{I}$. So, $\mathbf{Q}Z \sim \mathcal{N}(0,
\mathbf{I})$. That $Z$ is invariant to rotation about the origin tells
us that its distribution is $d$-spherical, which is the shape we want to sample
from. Our remaining task then is to ensure that the norm of our sample is 1. By the
definition of the norm, $||aZ||_2 = |a|\times||Z||_2$. So, $||\frac{Z}{||Z||}||
= \frac{1}{||Z||} \times ||Z|| = 1$. In conclusion, because the distribution of
$Z$ is spherical, and $||Z / ||Z||_2||_2 = 1$, we are
confident that $Z / ||Z||_2$ is a sample from the surface of the $d$-dimensional
unit sphere.

We can generate $||X||_2 = R \sim u^{1/d}$, where $u$ is uniform on [0,1], using inversion
sampling. Let us examine why. We know the CDF of $R$: $F = P(R \leq t) = t^d$ for $0
\leq t \leq 1$. We find the inverse CDF $F^{-1}$ by solving $F(F^{-1}(u))=u$,
leading to $F^{-1}(u) = u^{1/d}$. $F^{-1}(u)$ has $F$ as its CDF. Proof:
$P(F^{-1}(u) \leq t) = P(u \leq F(t)) = F(t)$, since for random uniform [0,1]
$u$, $P(u \leq x) = x$. So, sampling via $u^{1/d}$ is equivalent to sampling from
the desired distribution with CDF $t^d$.

\subsection*{C}

The methods implemented in \textbf{A} and \textbf{B} can be empirically
compared. Each program was used to generate 100 samples with $d=10$. The
execution runtime was measured with the Linux \texttt{time} command. Method B is faster than Method
A, as shown in Table 1.

	\begin{table}[h]
		\centering
		\caption{Runtime comparison (seconds)}
		\begin{tabular}{|c c c|}
			\hline
			& \textbf{A} & \textbf{B} \\
			\textbf{real} & 2.131 & 0.343\\
			\textbf{user} & 2.150 & 0.329\\
			\textbf{sys}  & 0.172 & 0.170\\
			\hline
		\end{tabular}
	\end{table}

\section*{Problem 2}

\subsection*{A}

See submitted code \texttt{hw1\_2A.py}.

As $d$ grows larger, the frequency of the event $\{||X_+ -
Z||^2_2 \geq ||X_{-} - Z||_2^2\}$ increases, as depicted in Figure 1. This
result is somewhat counter-intuitive.
Because $Z$ is drawn from the same distribution as $X_+$, we would expect their
difference $X_+ - Z$ to result in something close to a zero-vector, which should
have a smaller Euclidean norm than that of $X_- - Z$
(since these two have different
means), resulting in \textit{low} frequency of the event. This is true in low
dimensions, but as dimensionality grows, the event occurs with \textit{high}
frequency. Why this counter-intuitive result occurs is explained in \textbf{B}.

\begin{figure}[h!]
	\centering
	\caption{Frequency of event out of 1000 draws}
	\begin{tikzpicture}
\begin{axis}[ylabel = Number of Events, 
	     xlabel = Dimensionality,
	     xtick = data,
	     symbolic x coords = {1, 2, 5, 10, 20, 50, 100, 200, 500, 1000}]
	\addplot+[ybar, fill = blue] coordinates { 
	(1, 124) 
	(2, 211) 
	(5, 393)
	(10, 617)
	(20, 846)
	(50, 994)
	(100, 1000)
	(200, 1000)
	(500, 1000)
	(1000, 1000)};
\end{axis}
\end{tikzpicture}
\end{figure}

\subsection*{B}

$$ E[||X_+ - Z||^2_2] $$
$$ = E[||X_+||_2^2] -2E[X_+^TZ] + E[||Z||^2_2] $$
$$ E[||X_+||^2_2] = ||\mu_+||^2_2 + tr(\Sigma_+) = 25 + 4d$$
$$ E[||Z||^2_2] = E[||X_+||^2_2] = 25 + 4d$$
$$
E[X_+^TZ] = E[X_+^T]E[Z] = ||\mu_+||^2 = 25
$$
$$
E[||X_+ - Z||^2_2] = 2(25+4d) - 2(25) = \boxed{8d}
$$

\medskip

$$ E[||X_- - Z||^2_2] $$
$$ = E[||X_-||_2^2] -2E[X_-^TZ] + E[||Z||^2_2] $$
$$ E[||X_-||^2_2] = ||\mu_-||^2_2 + tr(\Sigma_-) = 25 + d$$
$$ E[||Z||^2_2] = 25 + 4d$$
$$ E[X_-^TZ] = E[X_-^T]E[Z] = -25$$
$$E[||X_- - Z||^2_2] = (25+d) - 2(-25) + (25+4d) = \boxed{100+5d}$$

\medskip

This theoretical result explains our empirical observations in \textbf{A}. With
low dimensionality, $8d < 100 + 5d$, and so our event occurs with low frequency.
However, $8d$ grows faster with $d$ than does $100 + 5d$, and so in
high dimensions, our event occurs with high frequency. In essence,
concentration of measure / curse of
dimensionality magnify the effect of $X_+$'s higher variance as dimensionality
grows higher. The different rate of
growth of expected value with $d$ is visualized in Figure 2.

\begin{figure}[h!]
	\centering
	\caption{Change in expected value with $d$}
	\begin{tikzpicture}
\begin{axis}[ylabel = Expected value, 
	     xlabel = Dimensionality,
	     legend pos = north west]
	     \addplot[domain=0:300, color=blue]{8*x};
	     \addlegendentry{$||X_+ - Z||^2_2$}
	     \addplot[domain=0:300, color=orange]{100+5*x};
	     \addlegendentry{$||X_- - Z||^2_2$}
\end{axis}
\end{tikzpicture}
\end{figure}
\end{document}

### 2018-03-11 hw1/hw1_backup.tex
initial commit

\documentclass{article}

\usepackage{geometry}
\geometry{margin=1in}

\usepackage{pgfplots}

\title{STAT 672: Homework \#1}
\author{Tom Wallace}

\setlength{\parindent}{0cm}

\begin{document}

\maketitle

\section*{Problem 1}

\subsection*{A}

See attached code \texttt{hw1\_A.py} for an implementation of rejection sampling
from the unit cube.

\smallskip

The acceptance probability is the volume of an $n$-dimensional ball divided by
the volume of an $n$-dimensional cube:

\begin{equation}
P_{X \sim B^d_\infty}(||X||_2 \leq 1) =
\frac{\mathrm{Vol}(B^d_2)}{\mathrm{Vol}(B^2_\infty)}
= \frac{\frac{\pi ^ {d/2}}{\Gamma(\frac{d}{2} + 1)}}{2^d}
\end{equation}

We know that $\Gamma (n+1) = n!$ Thus, $\Gamma (\frac{d}{2} + 1) = (\frac{d}{2})!$

\smallskip

Stirling's approximation for factorials is useful here.

\begin{equation}
n! \approx \sqrt{2 \pi n}n^ne^{-n}
\end{equation}

Let us consider the case of $2d$. In this case:

\begin{equation}
\mathrm{Vol}(B^{2d}_2) = \frac{\pi ^ d}{\Gamma(d + 1)} = \frac{\pi^d}{d!}
\approx \frac{\pi^d}{\sqrt{2\pi d}d^de^{-d}}
\end{equation}

Rearranging terms, we have:

\begin{equation}
= \frac{1}{\sqrt{2\pi d}} \left( \frac{\pi e}{d} \right)^d
\end{equation}

Referring to $ \left( \frac{\pi e}{d} \right)^d$, the denominator grows faster with $d$ than does the
numerator. We
can show this more formally by computing the derivative of the numerator and of the
denominator, respectively:

\begin{equation}
\frac{d}{dx} ((\pi e)^x) = (\pi e)^x (1 + \ln \pi)
\end{equation}

\begin{equation}
\frac{d}{dx}(x^x) = x^x (1 + \ln x)
\end{equation}

As $x$ grows larger, (6) becomes increasingly larger than (5). As a consequence, if we extend $d$ infinitely, 
the limit of the ratio is 0.

\begin{equation}
\lim_{d \to \infty} \left( \frac{\pi e}{d} \right)^d = 0
\end{equation}

Which implies:
\begin{equation}
\lim_{d \to \infty} \frac{1}{\sqrt{2\pi d}} \left( \frac{\pi e}{d} \right)^d = 0
\end{equation}

In other words, as $d \to \infty$, more and more of the volume of the unit ball
is contained near its surface. This phenomenon is called \textit{concentration
of measure}.

\smallskip

Returning to (1), we compute the the volume of the unit cube as $d \to \infty$.

\begin{equation}
\lim_{d \to \infty} 2^{2d} = \infty
\end{equation}

So, the ratio of the unit sphere to the unit cube is:

\begin{equation}
\lim_{d \to \infty} \frac{\mathrm{Vol}(B^d_2)}{\mathrm{Vol}(B^2_\infty)} =
\frac{0}{\infty} = 0
\end{equation}

Since this ratio also is the probability of accepting a sample in a rejection sampling
scheme, our expected runtime to generate a single random vector from $\mathrm{B}^d_2$ with
rejection sampling grows asymptotically with $d$. For example, for $d=10000$, our
program would run for a very, very, \textit{very} long time.

\subsection*{B}
We can generate $\Theta$ as $\Theta = Z / ||Z||_2$, where $Z \sim N(0, I_d)$ due
to the Box-Muller (B-M) transformation.

\section*{Problem 2}

\subsection*{A}

Below are the results of my simulation, the code of which is attached as
\texttt{hw1\_B.py}. As $d$ grows larger, the frequency of the event $\{||X_+ -
Z||_2 \geq ||X_{-} - Z||_2\}$ increases. This is somewhat counter-intuitive.
Because $Z$ is drawn from the same distribution as $X_+$, we would expect their
difference $X_+ - Z$ to result in something close to a zero-vector, which should
be smaller than the difference between $X_-$ and $Z$ (since they have different
means and hence should have a larger difference). Yet, as dimensionality grows,
this is not the case: the event occurs with increasing frequency. This has two
explanations. One, because the mean is $\pm 5/\sqrt{d}$, as $d$ increases, the
means increasingly converge towards 0. Two, $X_+$ and $Z$ have larger variance than
$X_-$, and this variance does not decrease with $d$. Thus, as $d$ increases, the
larger variance of $X_+$ and $Z$ predominates over their increasingly similar
means, resulting in them more often being further apart (their difference has a
larger Euclidean norm) than $X_-$ and $Z$.

\begin{tikzpicture}
\begin{axis}[ylabel = Number of Events, 
	     xlabel = Dimensionality, 
	     xbar interval = 0.8,
\addplot[fill=blue] coordinates { (foo, 1) (bar, 2)};
\end{axis}
\end{tikzpicture}

\end{document}

### 2018-03-11 hw2/hw2.tex
initial commit

\documentclass{report}

\usepackage{geometry}
\geometry{margin=1in}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{graphicx}
\graphicspath{}

\DeclareMathOperator*{\argmax}{arg\,max}

\title{STAT 672: Homework 2}

\author{Tom Wallace}

\begin{document}

\maketitle

\section*{Problem 1}

\subsection*{A}

For a supervised learning problem, risk is defined as the expected value
of the loss function:
$$
R(f) = \mathbf{E}_{X, Y \sim P}[L(Y, f(x))]
$$
Bayes risk is the risk present when using the Bayes classifier $f^*(x)$:
$$f^*(x) = \argmax_Y P(Y=y|X=x)$$
Typically, this classifier is not practical because we do not know the
conditional distribution of $Y$ given $X$, but in this problem it is given.
Our resultant Bayes classifier is:
$$
f^*(x) = 
\begin{cases}
	0 & x \in [0.2, 0.8] \\
	1 & x \in \{(0, 0.2) \cup (0.8, 1)\} 
\end{cases}
$$
Suppose that we use a typical 0-1 loss function.  
$$
L(y, f(x)) = 
\begin{cases} 
	0 & y = \mathrm{sign}(f(x)) \\ 
	1 & \mathrm{otherwise}
\end{cases}
$$
Risk in our problem is equal to:
$$
P(\mathrm{sign}(Y) \neq \mathrm{sign}(f(x)))
$$
Using the law of total probability, this is equal to:
\begin{equation}
	\begin{split}
		& P(Y=0|X \in \{(0, 0.2) \cup (0.8, 1)\})P(X \in \{(0, 0.2) \cup (0.8,
		1)\}) \\
		& + P(Y=1|X \in [0.2, 0.8])P(X \in [0.2, 0.8]) 
	\end{split}
\end{equation}

$$
(0.2 \times 0.1) + (0.2 \times 0.1) + (0.2 \times 0.6) = 0.16
$$

\subsection*{B}
First, consider $d=0$. $f(x)$ must take the form of $f(x) =
C$, with $C$ being some constant. Because $Y \in \{-1, 1\}$, our candidate functions are $f(x)
= 1$ (i.e. always predict that $Y=1$, regardless of $x$) and $f(x) = -1$ (i.e.
always predict that $Y=-1$, regardless of $x$). We calculate the risk for each:

\begin{equation*}
	\begin{split}
		R(f(x)) = & P(Y \neq 1 | X\in(0, 0.2))P(X\in(0, 0.2)) \\
		  & + P(Y \neq 1 | X\in[0.2, 0.8])P(X\in[0.2, 0.8]) \\ 
		  & + P(Y \neq 1 | X\in(0.8, 1))P(X\in(0.8, 1))
	\end{split}
\end{equation*}
$$
=(0.1)(0.2) + (0.8)(0.6) + (0.1)(0.2) = 0.52
$$
By similar logic, $R(f(x)=-1) = 0.48$. This is smaller than $0.52$ and so is the
best we can do for $d=0$. Excess risk for $d=0$ is:
$$
\min_{f \in \mathcal{F}_{d=0}} R(f) - R(f^*) = 0.48 - 0.16 = 0.32
$$
Lastly, consider $d=1$. Our candidate functions are of the form:
$$
f(x) = \alpha x + C
$$
This equation defines a straight line and so we are limited to a
``threshold'' approach: for example, always predict that $Y=1$ if $x$ is greater
than some value. Because the distribution of $Y|X$ is a simple step function, we
have some obvious candidate functions. Consider 
$f(x)=-x+0.2$, i.e. predict that $Y=1$ for
$X \in (0, 0.2)$ and $Y=-1$ for $X\in(0.2, 0.8)$:

\begin{equation*}
	\begin{split}
		R(f(x)) = & P(Y \neq 1  | X\in(0, 0.2))P(X\in(0, 0.2))  \\
		  & + P(Y \neq -1  | X\in(0.2, 0.8))P(X\in(0.2, 0.8))  \\ 
		  & + P(Y \neq -1 | X\in(0.8, 1.0))P(X\in(0.8,1.0)) 
	\end{split}
\end{equation*}
$$
= (0.1)(0.2) + (0.6)(0.2) + (0.9)(0.2) = 0.32
$$
By symmetry the risk of this function is the same as
other similar candidates that set the threshold at a discontinuity in the step
function: we cannot do better. Excess risk for $d=1$ thus is:
$$
\min_{f \in \mathcal{F}_{d=1}} R(f) - R(f^*) = 0.32 - 0.16 = 0.16
$$
Lastly, consider $d=2$. Our candidate functions are of the form:
$$
f(x) = \alpha_1 x^2 + \alpha_2 x + C
$$
A good candidate is $f(x)=(x-0.2)(x-0.8)$. This function will return a positive
number (i.e., predict that $Y=1$) if $x \in (0, 0.2)$, and will return a
negative number (i.e., predict that $Y=-1$) if $x \in (0.2, 0.8)$. Risk is thus
equal to:

\begin{equation*}
	\begin{split}
		R(f(x)) = & P(Y \neq 1  | X\in(0, 0.2))P(X\in(0, 0.2))  \\
		  & + P(Y \neq -1  | X\in(0.2, 0.8))P(X\in(0.2, 0.8))  \\ 
		  & + P(Y \neq 1 | X\in(0.8, 1.0))P(X\in(0.8,1.0)) 
	\end{split}
\end{equation*}
$$
(0.2)(0.1) + (0.2)(0.1) + (0.2)(0.6) = 0.16
$$
And excess risk is:
$$
\min_{f \in \mathcal{F}_{d=2}} R(f) - R(f^*) = 0.16 - 0.16 = 0
$$
Further increasing dimensionality ($d=3, 4\ldots$) will allow
us to do as well (i.e. achieve zero excess risk) but no better, since Bayes
risk is the theoretical optimum. We have proved the following two statements 
and so can conclude that excess risk is zero if and only if $d \geq 2$:

\begin{itemize}
	\item $\min_{f \in \mathcal{F}_{d \geq 2}} R(f) - R(f^*) = 0$ \qquad (if)
	\item $\min_{f \in \mathcal{F}_{d<2}} R(f) - R(f^*) > 0$ \qquad (only
		if)
\end{itemize}

\newpage

\subsection*{C}

See \texttt{hw2.py}. I empirically tested the accuracy of my $X$ and $Y|X$
generating functions by generating 10,000 observations with them and plotting the results,
as seen in Figure 1. Although this is just a crude 10-bin histogram 
and so does not perfectly match the theoretical distribution, it appears broadly
correct; for example, it has a ``barbell'' shape.

\begin{figure}[h]
	\caption{$X$ and $Y|X$}
	\centering
	\includegraphics[scale=0.5]{hist}
\end{figure}

\subsection*{D}

See \texttt{hw2.py}. I trained three separate logistic regression models for $d=1$, $d=2$, 
and $d=5$ on randomly generated $X$ and $Y$ ($n=100$), which were generated using the specified
PDFs (random uniform for $X$ and the given conditional density for $Y|X$). 
Table 1 shows each model's predictions for a fixed set of $X$.
\begin{table}[h]
	\centering
	\caption{Model predictions of $Y|X$}
	\begin{tabular}{|r r r r|}
		\hline
		\textbf{X} & \textbf{d=1} & \textbf{d=2} & \textbf{d=5} \\
		0    & 1 & 1 & 1 \\
		0.25 & 0 & 1 & 0 \\
		0.50 & 0 & 0 & 0 \\
		0.75 & 0 & 0 & 0 \\
		1    & 0 & 1 & 1 \\
		\hline
	\end{tabular}
\end{table}

\newpage

\subsection*{E}
See \texttt{hw2.py}. My results are shown in Figure 2.

As $n$ grows larger, empirical error increases and generalization error decreases. 
This phenomenon has an intuitive explanation.

When $n$ is small, the model has only a limited amount of information to learn
on. Its classification rules overfit to the few random observations in $D$ rather than
learning the true general relationship between $X$ and $Y$. Because $n$ is small, basic
probability means that $D^\prime$ likely will contain behavior not present in
$D$. The model has no idea how to predict these. The
model hence will do a good job of predicting $D$ (low empirical error) but a
poor job of predicting $D^\prime$
(high generalization error).

When $n$ is large, $D$ contains enough information for the model to learn the
true relationship between $X$ and $Y$. It no longer will overfit to $D$-idiosyncratic
behavior and so will attain higher empirical error than was present with small
$N$. But, since the model has learned the true relationship between $X$ and $Y$,
it also will do a better job of predicting $D^\prime$ (lower generalization error).

\begin{figure}[h]
	\caption{Advanced Simulation}
	\centering
	\includegraphics[scale=1]{plot}
\end{figure}
\end{document}

### 2018-03-11 sgd/sgd.tex
initial commit

\documentclass{beamer}
\usetheme{Madrid}

\usepackage{pgfplots}

\usepackage{bm}

\usepackage{graphicx}
\graphicspath{ {img/} }

\title{Stochastic Gradient Descent}
\subtitle{STAT 672 Project}
\author{Tom Wallace}
\institute{George Mason University}
\date{Spring 2018}

\begin{document}

\frame{\titlepage}

%%%%%%%%%%%%% Introduction %%%%%%%%%%%%%%

\begin{frame}
	\frametitle{Optimization is everywhere, and sometimes is easy}
	Many statistical procedures involve minimizing or maximizing some function 
	applied to data \\~\\

	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice'':
	\begin{itemize}
		\item \small Example: in OLS, we do not need to check every possible value
			of $\hat{\beta}$ to see if it minimizes the loss
			function, we (typically) can just evaluate
			$(X'X)^{-1}X'Y$ \\~\\
	\end{itemize}
	\smallskip

	\centering
	\begin{tikzpicture}
		\begin{axis}[
			yticklabels={,,},
			xticklabels={,,},
			width=0.5\textwidth,
			height=0.4\textwidth,
			legend style = {font=\small}
		]
			\addlegendentry{$f=\sum (\hat{y} - y)^2$}
			\addplot[mark=none, color=blue]{x^2};
		\end{axis}
	\end{tikzpicture}
\end{frame}

\begin{frame}
	\frametitle{Other times, optimization is not so easy}
	Suppose that we have a typical supervised classification problem:
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f^*$ from class $\mathcal{F}$
		\item $f$ has parameter vector $\bm{\theta}$
		\item Optimization: pick values for $\bm{\theta}$ that minimize
			empirical risk according to some
			loss function $L$
		\item Assume $L$ is convex (if not, problem is \textit{much}
			harder)\\~\\
	\end{itemize}

	Our lack of assumptions requires a different approach to optimization
	\begin{itemize}
		\item \small Cannot analytically identify stationary point
		\item Need to numerically search for it
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Gradient descent is a numerical approach to optimization}
	Intuition: gradient is multidimensional version of 1st derivative
	\begin{itemize}
		\item \small $\nabla f(\mathbf{X}) = (\frac{\partial f}{x_1},
			\frac{\partial f}{x_2}...)$
		\item So if the gradient is (approximately) 0 at some point, that point is
			(approximately) a stationary point 
		\item Since we know the loss function is convex, that stationary
			point must be the global minimum\\~\\
	\end{itemize}

	Give basic idea: take guess, step, evaluate, stop once below epsilon
\end{frame}

\begin{frame}
	\frametitle{Batch gradient descent, animated}
\end{frame}

\begin{frame}
	\frametitle{Batch gradient descent is computationally expensive}
	Elaborate
\end{frame}

\begin{frame}
	\frametitle{Stochastic gradient descent (SGD) is more efficient}
	Overview of how it works \\~\\
\end{frame}

\begin{frame}
	\frametitle{SGD, visualized}
\end{frame}

\begin{frame}
	\frametitle{Choice of hyper-parameters is important}
	Reminder of what they are \\~\\
	Step size from Bottou \\~\\
	Learning parameter
\end{frame}

\begin{frame}
	\frametitle{SGD has nice properties for high-dimensional
	data}
	Theoretical explanation \\~\\
	Empirical performance \\~\\

\end{frame}

\begin{frame}
	\frametitle{Applications of SGD}
	If a Silicon Valley press release uses any of the following phrases...
	\begin{itemize}
		\item \small ``Neural networks''
		\item ``Machine learning''
		\item ``AI'' 
	\end{itemize}

	...SGD probably is involved. Example: Google's \texttt{AlphaGo} program.

	\begin{figure}[b]
	\centering
	\includegraphics[scale=0.33]{go}
	\end{figure}

\end{frame}

\end{document}

### 2018-03-11 week1/hw1.tex
initial commit

\documentclass{article}
\begin{document}

\section{Problem 1}

$f(x) = ||\mathbf{Ax-b}||^2_2$

\medskip

This is just the ordinary least squares OLS problem

$= (\mathbf{Ax-b)'(Ax-b)}$

$= (\mathbf{x'A'-b')(Ax-b)}$

$= \mathbf{x'A'Ax - x'A'b - b'Ax + b'b}$

$= \mathbf{x'A'Ax - 2x'A'b + b'b}$

\medskip

$\nabla f(x) = \mathbf{2A'Ax-2A'b}$

$\nabla^2 f(x) = \mathbf{2A'A}$

\medskip

Reminders:

$\mathbf{Av = \lambda v}$

$\mathbf{v}$ is an eigenvector if it satisfies this equation

$\lambda$ is the eigenvalue corresponding to $\mathbf{v}$

Eigenvectors are the vectors that linear transformation $\mathbf{A}$ merely elongates or shrinks (as opposed to
rotating)

Eigenvalue is the amount that they elongate or shrink

To find, solve $det(\mathbf{A-\lambda I})=0$ (characteristic equation)

Eigendecomposition: $\mathbf{A=Q \Lambda Q^{-1}}$

$\mathbf{Q}$ is a $n \times n$ matrix whose $i$'th column is eigenvector $q_i$ of $\mathbf{A}$

$\Lambda$ is a diagonal matrix for which $\Lambda_{ii}=\lambda_i$

If eigenvalues of Hessian are all positive, then $\mathbf{A}$ is positive semi-definite (multivariate equivalent of
convex)

If all negative, then $\mathbf{A}$ is negative-definite (multivariate equivalent of concave)

Matrix $\mathbf{A}$ is positive definite IFF $\mathbf{A}=\mathbf{R}'\mathbf{R}$, where $\mathbf{R}$ is a matrix with
independent columns (full rank)

\medskip

So, in our case, assuming $\mathbf{A}$ is of full rank, since it has a Hessian of $\mathbf{A}\mathbf{A}'$, it is positive
definite and so all eigenvalues are positive.

\medskip

The necessary and sufficient condition for $x^*$ to be a minimizer of $f(x)$ is that the gradient is 0 at $x^*$

\medskip

$f$ has a unique minimizer if $A$ is positive definite

\section{Problem 2}

By the law of total probability, $ f_{X}(x) = \sum_{k=1}^K f_{X|Y=k}(x|k) f_Y(k)$

\medskip

$P(Y=k|X=x) = \frac{P(X=x|Y=k)P(Y=k)}{P(X=x|Y=k)P(Y=k)+P(X=x|Y\neq k)P(Y\neq k)}$
$=\frac{f_{X|Y=k}(x)P(Y=k)}{f_X(x)}$
\end{document}

### 2018-03-11 week2/notes.tex
initial commit

\documentclass{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pgfplots}
\DeclareMathOperator*{\argmax}{argmax}
\pagenumbering{gobble}
\begin{document}
High-dimensional data exhibit \textbf{concentration of measure} and
\textbf{curse of dimensionality}.

\medskip

\textbf{Concentration of measure}: for high-dimensional objects, most of the volume is concentrated near the
surface. Consider any object $A$ in $\mathbb{R}^d$. Shrink $A$ by some small
amount $\epsilon$ to produce a new object $(1-\epsilon)A$. It is the case that
$Vol((1-\epsilon)A) = (1-\epsilon)^d Vol(A)$. It also is the case that for any
$x$, $1-x \leq e^{-x}$. So, $\frac{Vol((1-\epsilon)A)}{Vol(A)} = (1-\epsilon)^d
\leq e^{-\epsilon d}$. As $d \to \infty$, then $(1-\epsilon)^d \to 0$. This
implies that nearly all the volume of $A$ is contained in the portion of A not
included in $(1-\epsilon)A$.

%\begin{center}
%	\begin{tikzpicture}[scale= .4]
%	\begin{axis}
%		\addplot[color=orange,
%		domain = -3:3]{exp(-x)};
%		\addplot[color=blue,domain=-3:3]{1-x};
%	\end{axis}
%\end{tikzpicture}
%\end{center}

\medskip

\textbf{Curse of dimensionality}: because of concentration of measure, some nice
things we would otherwise like to do become intractable. As dimensionality grows
and concentration of measure takes over, two random elements are likely to be
very far apart. This means that nearest neighbor methods don't work well.

\medskip

\textbf{Kernel density estimation (KDE)}: As motivation, suppose that we have some $d$-dimensional data
$\mathbf{X}$ and class labels $Y=k_1, k_2...$, and want to predict $Y$ based on
$\mathbf{X}$. In other words, we're considering $P(Y=k|X=x)$, which we know by
Bayes is equal to $\frac{P(X=x|Y=k)P(Y=k)}{P(X=x)}$. We don't particularly worry
about the bottom part of that equation since it doesn't depend on $Y$ and
becaues we often have decent information available for it. We often
have good information for the prior $P(Y=k)$ or can just use the sample
proportions. So the chief challenge is estimating the the density function
$f_k(x)$ associated with $P(X=x|Y=k)$. We have two ways of doing this:
parametric (assuming some model, e.g. Normal, and estimating parameters for it
via MLE) or non-parametric (make fewer assumptions).

KDE is a non-parametric method for $\hat{f}(x)$. As the simplest possible
example, we could use a histogram, where we divide the range of observed $x$'s
into some number $h$ of bins; count the proportion $\hat{p}_i$ of observations
in bin $i$ and divide by the length of the bin $h$. In this case, $\hat{f}_h(x)
= \sum_i^{m}\frac{\hat{p}}{h}$
(the subscript is to emphasize the degree to which this estimator depends on
choice of $h$). But, this is not a great estimator because observations very
close to the border only contribute to the density estimate for one bin, when it
seems like they should contribute to both. So, we use some function to weight
how much an observation should contribute to estimation of density at some
point, depending on how far away the observation is from the point (e.g.,
observation $x=100$ should not much increase our estimated density at $x=1$, but
should contribute a lot to our estimation of density at $x=101$). This weighting
function is called a \textit{kernel}. Implemented in one dimension, we have: $\hat{f}_{kde}(x) =
\frac{1}{n}\sum_{i=1}^{n}\frac{1}{h}K(\frac{X_i - x}{h})$. In multiple
dimensions, we have: $\hat{f}_{kde}(\mathbf{x}) =  \frac{1}{n}\sum_{i=1}^n
\frac{1}{|H|^{1/2}}K(\frac{\mathbf{X_i}-\mathbf{x}}{h})$

Our main requirements are that the kernel integrate to one, and that it be
symmetric. The most common are Epanechnikov and normal (Gaussian). Our big
choice is again the size of window to consider (e.g., the standard deviation in
our Gaussian kernel). A large window means high bias and low variance
(under-fit). A small window means low bias and high variance (over-fit). A
method of evaluating candidate kernels is the \textbf{mean integrated squared
error (MISE)}, aka \textbf{L2 risk function}, which is: $E(\hat{f} - f)^2 =
\int_{-\infty}^\infty (\hat{f}(x) -f(x))^2$. The optimal kernel might be one
minimizing this loss function.

The problem is that as $d$ grows, the performance of KDE suffers, and there is
no estimator achieving better performance. The fundamental issue is that the CoD
means points become sparsely distributed in space and so $h$ has to become very
large to ensure that enough points fall into each neighborhood.

\medskip

\textbf{Naive Bayes}: as a motivating example of why KDE is useful, however, we
consider Naive Bayes. We make a big assumption that given a particular label
$Y=k$, the $p$ features $\mathbf{X}_k$ are independent of each other, implying that 
the joint density of $\mathbf{X}_k$ is just the product of the individual
densities: $f(\mathbf{X}_k)= \prod_{j=1}^p f(\mathbf{X}_{jk})$. The nice thing
is that we can estimate these individual densities via KDE (since we're doing
them one at a time the CoD doesn't apply). So, we can end up with
$P(Y_i=k|\mathbf{X}_i=\mathbf{x}) =  P(Y_i=k)P(\mathbf{X}_i=\mathbf{x}|Y_i=k) =
\hat{p}_Y(k) \hat{f}_{X|Y}(\mathbf{x}|k)$, where $\hat{p}_Y(k)$ is our prior,
$\hat{f}_{\mathbf{X}|Y}(\mathbf{x}|k)$
is our KDE-based estimate of the likelihood of vector $\mathbf{x}$, and we don't
care about the constant denominator that does not depend on $Y$. We use the
\textbf{maximum a posteriori (MAP)} decision rule, which means we chose the
$\hat{y}$ that maximizes the aforementioned function: 
$$\hat{y} = \argmax_k \hat{p}_Y(k) \hat{f}_{X|Y}(\mathbf{x}|k)$$

This is a useful technique when the dimensionality is very high (e.g. NLP, spam
detection, etc.) in part because of the ease of computing KDEs.
\end{document}

### 2018-03-11 week3/week3_notes.tex
initial commit

\documentclass{article}
\usepackage{amssymb}
\usepackage{geometry}
\geometry{margin=1.5in}
\usepackage{graphicx}
\graphicspath{{img/}}
\begin{document}

We want to find function $f$ from class $\mathcal{F}$, which we call the
\textbf{hypothesis class}.

\smallskip

We choose a \textbf{loss function} to measure how well $f$ does at the learning
problem.

\smallskip

\textbf{Risk} or \textbf{expected loss} is:

$$
R(f) = E(X,Y) \sim_P[L(Y, f(x))]
$$
or
$$
R(f) = E(X) \sim_P[L(X, f(x))]
$$

\smallskip

We denote the optimal risk as $r^*$ and the optimal function as $f^*$

\smallskip

The Bayes classifier, $f^*(x) = argmax_y P(Y=y|X=x)$, is the theoretical optimum
(at least for 0-1 loss). But, it isn't a practical scheme because we don't
actually know the conditional distribution. Rather, we use it as a benchmark.

\smallskip

We have a host of \textbf{margin-based} loss functions (margin is
$yf(\mathbf{X})$ for $\mathbf{X} \in \mathbb{R}, y \in {-1,1}$). The margin only
takes positive values when the classification is correct. A margin-based
loss function is: $L(y, f(x)) = L(yf(x))$. The simple 0-1 loss function can be
written as $I(yf(x) < 0$ (we want to minimize this). We are interested in
alternate loss functions that have two key properties:
\begin{itemize}
	\item $L$ is a convex function of the margin
	\item $L$ upper bounds $L_{0-1}$ (our above simple 0-1 loss function)
\end{itemize}

\smallskip

Some candidate functions (for classification) are below. All of them have the
same sign as Bayes loss and so are called \textbf{classification-calibrated} or
\textbf{surrogate} loss functions, in that minimizing them is equivalent to
minimizing Bayesian loss.

\begin{table}[h]
	\centering
	\begin{tabular}{|c | c | c|}
		\hline
		\textbf{Loss function} & \textbf{Form} & \textbf{Notes}\\
		Hinge & $\mathrm{max}\{0, 1-yf(x)\}$ & Tightest convex upper bound to 0-1; non-smooth\\
		Exponential &  $ \mathrm{exp}(-yf(x))$ & 1-Lipschitz (robust)\\
		Logistic & $ \mathrm{log}_2(1+\exp(-yf(x)))$ & 1-Lipschitz (robust)\\
		Squared & $(1-yf(x))^2$ & Not recommended for class. (weird errors)\\
		Squared hinge &  $\mathrm{max}\{0, 1-yf(x)\}^2$ & Only 1x differentiable
		at 1\\
		\hline
	\end{tabular}
\end{table}
\begin{centering}
\includegraphics[scale=0.35]{graph}
\end{centering}

Now, let's switch from classification to regression. Some potential loss
functions are:

\begin{table}[h]
	\centering
	\begin{tabular}{|c | c |}
		\hline
		\textbf{Loss function} & \textbf{Form} \\
		Squared & $(y-f(x))^2$  \\
		Absolute & $L_1(y, f(x)) = |y-f(x)|$\\
		$\epsilon$-insensitive & $|y-f(x)|I(|y-f(x)|\geq \epsilon)$ \\
		Huber & loooong \\
		\hline
	\end{tabular}
\end{table}

\begin{centering}
\includegraphics[scale=0.35]{graph2}
\end{centering}

We want minimize risk (expected loss) ($E_{(X,Y)\sim_P[L(Y, f(X))]}$) over all
possible $f$. But that's ambitious, so lets replace $R$ by its empirical
counterpart:

$$
R_{\mathrm{emp}} (f) = \frac{1}{n} \sum_{i=1}{n} L(Y_i, f(X_i))
$$

By LLN, we know that for any \textbf{fixed} f, $R_{\mathrm{emp}}(f) \to R(f)$ as
$n \to \infty$, but we can't be assured that we have infinite $n$, nor is $f$
fixed (we are considering lots of $f$), so this is a weak guarantee. Let $\mathcal{F}$ be the hypothesis class the minimum is taken over:

$$
\mathrm{min}_{f \in \mathcal{F}} R_{\mathrm{emp}}(f)
$$

Denote empirical risk minimizer as $\hat{f}$. Denote minimizer of risk over the
class $\mathcal{F}$ as $\bar{f}$. The difference between the minimum risk $\bar{R}$ over $\mathcal{F}$ and the
minimum risk $R^*$ is called \textbf{excess risk}. Reminder: $R(f^*)$ is
theoretical optimum. Theorem:
$$R(\hat{f}) \leq R(\bar{f}) + 2 \mathrm{sup}_{f \in
\mathcal{F}}|R_\mathrm{emp}(f)-R(f)|$$
$$= R(f^*) + \epsilon(\bar{f}) + 2 \mathrm{sup}_{f \in
\mathcal{F}}|R_\mathrm{emp}(f)-R(f)|$$
$$
= \mathrm{intrinsic error} + \mathrm{approximation error} + \mathrm{estimation
error}
$$

The latter two are bias and variance by another name. They are antagonists:
decreasing one means increasing another.

\smallskip

You can fit any series of datapoints with interpolating polynomial of degree
$n-1$. You are guaranteed to minimize empirical risk, but it will generalize
poorly to new data. Or, you can have some really general method, but it will
have high empirical risk.

\smallskip

A good way to find the optimal balance is \textbf{regularization}, covered next
class.

\end{document}

### 2018-03-11 week4/week4notes.tex
initial commit

\documentclass{minimal}
\usepackage{amsmath}
\usepackage{amssymb}
\DeclareMathOperator*{\argmin}{argmin}

\begin{document}

\textbf{Regularization}

We want to minimize empirical risk, but taken to an extreme we will overfit. So,
we need some strategy to enforce parsimony and generality. We add an extra term
that penalizes complexity. General form:
$$
\mathrm{min}_{f \in \mathcal{F}}R_{\mathrm{emp}}(f) + \lambda \Omega (f)
$$
with $\lambda$ the regularization parameter and $\Omega$ the regularizer.

\medskip

\textbf{Regularization for least squares}

OLS tends to become overfit when $D$ is large relative to $n$ (and is completely
useless if $D>n$). Ridge and lasso are techniques to regularize. SVD is useful
for them.

\medskip

\textbf{Singular value decomposition (SVD)}

Any $n \times p$ matrix $\mathbf{A}$ can be decomposed into the following:

$$
\mathbf{A} = \mathbf{UDV}'
$$

Where $\mathbf{U}$ is a $n \times p$  matrix, $\mathbf{V}$ is a $p \times p$
matrix, and $\mathbf{D}$ is a $p \times p$ diagonal matrix.

Conceptually, SVD is similar to the eigen-decomposition
$\mathbf{A}=\mathbf{Q\Lambda Q}' $, where $\mathbf{A}$'s eigenvalues are the
diagonal of $\mathbf{\Lambda}$ and $\mathbf{Q}$ is orthogonal. But the
eigen-decomposition only is valid for positive definite matrices, whereas SVD is
applicable for any matrix.

The diagonal values of $\mathbf{D}$ are the singular values of $\mathbf{A}$.
They are the square root of the eigenvalues of $\mathbf{A}'\mathbf{A}$. All
$d_1, d_2...d_p \geq 0$. If one ore more equals 0, $\mathbf{A}$ is singular and
not of full rank. The
ratio of the largest to the smallest $d$ is called the condition number. A high
condition number means an ill-conditioned matrix (i.e. one that is not
numerically stable and may have results ``blow up'' from minor changes, perhaps
from rounding off).

\medskip

\textbf{SVD and OLS}

Let $\mathbf{U}_i$ be the $i$'th column of $U$, and similar for $\mathbf{V}$.
Subscript $r$ means the effective rank.
$$
\hat{\mathbf{\beta}} = \mathbf{V}_r\mathbf{D}^{-1}\mathbf{U}_r'\mathbf{y}
$$
$\mathbf{V}_r\mathbf{D}^{-1}\mathbf{U}_r'$ is the pseudo-inverse of
$\mathbf{A}$. This method is more numerically stable than the normal
equations.

Also: the hat matrix can be computed via SVD.

$$
\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}' = \mathbf{UU}'
$$

\medskip
\textbf{Ridge regression (and SVD)}

Stepwise and similar schemes are discrete, all-or-nothing. Ridge regression instead is continuous and
shrinks the coefficients via a penalty on their size. Also known as weight decay in neural networks.

$$
\hat{\beta}_{\mathrm{ridge}} = \mathrm{argmin}_\beta \left [ \sum_{i=1}^n (y_i -
\beta_0 - \sum_{j=1}^p \beta_j x_{ij}) + \lambda \sum_{j=1}^n \beta^2_j) \right ]
$$
$$
\hat{\beta}_{\mathrm{ridge}} = (y-X\beta)'(y-X\beta) + \lambda \beta'\beta =
(X'X+I\lambda)^{-1}X'y
$$

The fit (equivalent to $X\beta$ in OLS) can be expressed via SVD as:
$$
X\beta_{\mathrm{ridge}} = \mathbf{U}
\mathrm{diag}\left (\frac{d_1}{d_1+\lambda}...\frac{d_p}{d_p + \lambda})\right )
\mathbf{U}'\mathbf{y}
$$

You have to normalize $x_{ij}$ as $x_{ij}-\bar{x}_j$ and estimate
$\beta_0=\bar{y}$ (ridge doesn't estimate intercept).

Ridge makes the problem non-singular even if $\mathbf{X}$ is not of full rank.

Ridge works well if the first few eigenvalues are dominant and the rest small,
and the regularization penalty chosen at the knee in the curve.

Ridge does not work well for sparse set-ups.

\medskip
\textbf{Lasso}

Basic idea: ridge but with L1 norm. Zero norm (i.e. return 1 if $w_j \neq 0$)
would be even better but is computationally intractable.

$$
\hat{\beta}_\mathrm{lasso} = \argmin_{\beta \in \mathbb{R}^d}
\frac{1}{2n}||\mathbf{y} - \mathbf{X}\beta||^2_2 + \lambda||\beta||_1
$$

Naturally produces sparse solutions, i.e. is implicitly performing feature
selection
\end{document}

### 2018-03-11 week6/week6.tex
initial commit

\documentclass{minimal}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}

\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}

\begin{document}

\textbf{Linear classifiers}

Linear discriminant analysis (LDA)

Logistic regression

Support vector machine (SVM)

\medskip

\textbf{LDA}

Assume classes are multivariate Gaussian with common covariance matrix $\mathbf{\Sigma}$

It can be shown that under these assumptions the decision boundary between any
two classes is linear, i.e. a hyperplane in $\mathbb{R}$.

\smallskip

$w$ is a weight vector that defines a projection into a 1-dimensional subspace.
Then, in that sub-space, we set a threshold $t$ as the boundary between the two
classes.

\includegraphics[scale=0.25]{lda}

Fisher's optimal projection: maximize between-class variance relative
to within-class variance. In other words: projected class centroids are far
apart, projected data are close to centroids.

$$
\hat{w} = \argmax_{w \in \mathbb{R}^d} \frac{w'\mathbf{B}w}{w'\mathbf{W}w}
$$

$$
\mathbf{B} = \sum_c (\bm{\mu}_c - \bar{\mathbf{x}}) (\bm{\mu}_c -
\bar{\mathbf{x}})'
$$
$$
\mathbf{W} = \sum_c \sum_{i \in c}(\mathbf{x}_i - \bm{\mu}_c) (\mathbf{x}_i -
\bm{\mu}_c)'
$$

\smallskip

Because $w$ is invariant to rescaling, we can choose $w$ such that
$w'\mathbf{W}w=1$, leading to:

$$
w^* = \argmax_w w'\mathbf{B}w 
$$
\begin{center}
s.t.
\end{center}
$$
w'\mathbf{W}w=1
$$

This is a generalized eigenvalue problem, with $w$ given by the largest
eigenvalue of $\mathbf{W}^{-1}\mathbf{B}$

\smallskip

If we assume Gaussian class densities and common covariance matrix
$\mathbf{\Sigma}$, then LDA is optimal (equivalent to Bayes).

\medskip

\textbf{Interlude: Lagrangian optimization}

Suppose that we have a standard optimization problem with $m$ inequality and/or
$p$ equality constraints:

\smallskip

Minimize $f_0(x)$

s.t. $f_i(x) \leq 0, \quad i=1...m$

$h_i(x)=0, \quad i = 1...p$

The Lagrangian is:
$$
L(x, \lambda, v) = f_0(x) + \sum_{i=1}^m \lambda_if_i(x) + \sum_{i=1}^p v_i
h_i(x)
$$

The dual function is the minimum value of the Lagrangian over $x$:
$$
g(\lambda, v) = \inf_x L(x, \lambda, v)
$$

This function is concave, even if the original problem is not convex. It also is
always less than or equal to the original objective function evaluated at its
optimal value. Putting these together, if we maximize $\lambda$ and $v$ in the dual, it is the same
as minimizing $x$ in the primal.

\smallskip

Slater's condition: if the primal and dual are feasible, then the gap is zero.

\smallskip

Complementary slackness: if a variable is positive, then the associated dual
constraint must be binding. If a constraint fails to bind, then the associated
variable must be zero.

\medskip

\textbf{Logistic}

Same as usual. Estimate using MLE. Can add a regularizer. Often outperforms LDA
because it doesn't have squared error (bad). Example:

\includegraphics[scale=0.25]{log}

\medskip

\textbf{SVM}

\smallskip

\textit{Hard margin (linearly separable data)}

Suppose we have labels $Y_i \in \{-1,1\}$ and data $X_i$.

We want to construct a decision hyperplane $\langle w, X \rangle + b= 0$. $w$ is
a weight vector that we apply to $X$. If for a particular observation $\langle w, X_i \rangle + b < 0$, we
classify as $Y_i=-1$. Similarly, if $\langle w, X_i \rangle + b > 0$, we
classify it as $Y_i = 1$

Notation note:
$$b + \langle w, x \rangle = \mathbf{w}^T\mathbf{X} + b = \sum w_ix_i + b$$

What makes a good hyperplane? Assuming linear separability, a criterion is that
it maximizes the minimum distance from the data points to the hyperplane. The
distance of a point $x_i$ to a hyperplane $w'x + b$ is:

$$
\frac{|w'x_i + b|}{||w||}
$$

So, we might construct a good hyperplane by trying to maximize the distance to
the closest point:

$$
\max_{w, b} \min\{\frac{|w'x_i + b|}{||w||}\}
$$

We also want to make sure that it classifies correctly, so we add the
below constraint.

$$
s.t. \quad Y_i(\langle w, x_i \rangle + b) \geq 1 \quad \forall i
$$

A problem, however, is that this won't have a unique solution because we can
scale the problem by some constant $\gamma$ and obtain identical results. We get
around this by stipulating that the hyperplane must be \textit{canonical}
to $X$, which means that the distance between the hyperplane and the closest
point is unity.

$$
\min_{1 \leq i \leq n} |\langle w, x_i \rangle + b| = 1
$$

Since we now know that the distance between the hyperplane and the point closest
to the hyperplane is unity, the problem becomes:

$$
\max_{w, b}\frac{1}{||w||}
$$
$$
s.t. \min_{1 \leq i \leq n} |\langle w, x_i \rangle + b| = 1
$$
$$
\quad Y_i(\langle w, x_i \rangle + b) \geq 1 \quad \forall i
$$

Maximizing the above function is equivalent to minimizing $||w||$. We will add
some other stuff to the problem to make later work easier.

$$
\min_{w, b}\frac{1}{2}||w||^2
$$
$$
s.t. \min_{1 \leq i \leq n} |\langle w, x_i \rangle + b| = 1
$$
$$
\quad Y_i(\langle w, x_i \rangle + b) \geq 1 \quad \forall i
$$

This is a quadratic function with linear constraints, aka quadratic programming.
The Lagrangian is:
$$
L(b, w, \alpha) = \frac{1}{2}||w||^2 + \sum_{i=1}^n \alpha_i [1 - Y_i(\langle
w, x_i \rangle + b)]
$$

We need the minima to compute the dual. We find them by taking the derivative
and setting equal to 0:

$$
\nabla_w = 
\nabla_w \left( \frac{1}{2}||w||^2 \right) + 
\nabla_w \left(\sum_{i=1}^n \alpha_i \right) - 
\nabla_w \left( \sum_{i=1}^n \alpha_i Y_i \langle w, x_i \rangle \right) - 
\nabla_w \left( \sum_{i=1}^n \alpha_i Y_i b \right)
= 0
$$
$$
w = \sum_{i=1}^n \alpha_i Y_i x_i
$$
$$
\frac{\partial L}{\partial b} = \frac{\partial}{\partial b}
\left( \frac{1}{2}||w||^2 \right) + \frac{\partial}{\partial b}
\left(\sum_{i=1}^n \alpha_i \right) - \frac{\partial}{\partial b}\left( \sum_{i=1}^n \alpha_i Y_i \langle
w, x_i \rangle \right) - \frac{\partial}{\partial b} \left( b \sum_{i=1}^n
\alpha_i Y_i \right)
=0
$$
$$
\sum_{i=1}^n \alpha_i Y_i = 0
$$

We plug them into the Lagrangian to obtain the dual. By
definition, $||v||^2 = \langle v, v \rangle$. So:

$$
\frac{1}{2} ||w||^2 = \frac{1}{2} \langle w, w \rangle = 
\frac{1}{2} \langle \sum_{i=1}^n \alpha_i Y_i x_i, \sum_{j=1}^n \alpha_j Y_j x_j \rangle
$$

The $\alpha$ and $Y$ are just scalars and so we can pull them out:

$$
= \frac{1}{2} \sum_{i=1}^n \sum_{j=1}^n \alpha_i \alpha_j Y_i Y_j \langle x_i, x_j \rangle
$$

Now we factor out the right hand side:

$$
\sum_{i=1}^n \alpha_i [1 - Y_i(\langle w, x_i \rangle + b)] = \sum_{i=1}^n
\alpha_i [1 - Y_i \langle w, x_i \rangle - Y_i b]
$$
$$
= \sum_{i=0}^n \alpha_i - \sum_{i=0}^n \alpha_i Y_i \langle w, x_i \rangle -
b \sum_{i=0}^n \alpha_i Y_i
$$

We know that:

$$
b \sum_{i=0}^n \alpha_i Y_i = b \times 0 = 0
$$

Now let's plug $w$ into the middle term:

$$
\sum_{i=0}^n \alpha_i Y_i \langle \sum_{j=0}^n \alpha_j Y_j x_j, x_i \rangle
$$

Again we can pull out the scalars:

$$
= \sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j Y_i Y_j \langle x_i, x_j \rangle
$$

We recombine everything to form the dual:

$$
q(\alpha) = \frac{1}{2} \sum_{i=1}^n \sum_{j=1}^n \alpha_i \alpha_j Y_i Y_j
\langle x_i, x_j \rangle + \sum_{i=1}^n \alpha_i - 
\sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j Y_i Y_j \langle x_i, x_j \rangle
$$

$$
= \sum_{i=1}^n \alpha_i - \frac{1}{2} \sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j Y_i Y_j \langle x_i, x_j \rangle
$$

Hence, the dual problem is:

$$
\max_{\alpha \in \mathbb{R}^n} 
\sum_{i=1}^n \alpha_i - \frac{1}{2} \sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j Y_i Y_j \langle x_i, x_j \rangle
$$
$$
s.t. \quad \alpha_i \geq 0 \quad \forall i
$$
$$
\quad \sum_{i=0}^n Y_i \alpha_i = 0
$$

The first constraint corresponds to our ``correctness'' constraint in the
objective. The second constraint is a ``balance'' constraint: that the distances
from points to the hyperplane must balance each other out.

Both the primal and dual are feasible, so the dual gap is 0 and solving the dual is
equivalent to solving the primal.

Looking at this dual gives us some interesting properties. Complementary slackness 
says that if a variable is positive,
then the associated dual constraint must be binding; conversely, if a constraint
fails to bind, then the associated dual variable must be zero. In the primal, we
had the constraint $Y_i(b + \langle w, x_i \rangle)\geq 1$, which is the same as
$1 - Y_i(b + \langle w, x_i \rangle) \leq 0$. If for primal constraint $1 - Y_i(b + \langle w, x_i
\rangle) = 0$, i.e. $x_i$ lays exactly on the margin, then the dual variable
$\alpha_i > 0$. In other words, only points laying on the
margin---\textit{support vectors}---determine the
separating hyperplane.

This definitely helps our robustness: 
variation of training points won't change our decision hyperplane unless the
points happen to fall on the margin.

It also can help our computation if we have
some heuristics to determine before computing $w$ what are some likely support
vectors.

In essence, specifying a hyperplane---a function of a simple form $w'x$---helps
us control $d$ and avoid the curse of dimensionality. The support vector
property helps us control $n$.

To determine $b$, we average $Y_i - \langle w, x_i \rangle$ over all $i$ with
$\alpha_i > 0$.

\includegraphics[scale=0.25]{svm1}

\medskip

\textit{Soft margin}

The above approach may not always be possible (since it requires linear
seperability) or desirable (since it may lead to overfitting). So, we have the
alternative soft margin approach. We introduce slack variables, as well as a 
cost parameter to control them. The problem becomes:

$$
\min_{w \in \mathbb{R^d}, b \in \mathbb{R}, \xi \in \mathbb{R}^n} 
\frac{1}{2} ||w||^2 + C \frac{1}{n} \sum_{i=1}^n \xi_i
$$
$$
s.t. \quad Y_i(\langle w, x_i \rangle + b) \geq 1 - \xi_i \quad \forall i
$$
$$
\quad \xi_i \geq 0 \quad \forall i
$$

We also can view this problem through the lens of ERM. Let's say we obtain an
optimal solution $w^*, b^*, \xi^*$. Then, it makes sense that:
$$
\xi_i^* = \max(0, 1 - Y_i(\langle w, x_i \rangle + b)) 
$$

We now can recharacterize the problem as:
$$
\min_{w, b} C \frac{1}{n} \sum_{i=1}^n \max(0, 1 - Y_i(\langle w, x_i \rangle +
b)) + \frac{1}{2}||w||^2
$$

This is equivalent to:
$$
\min_{f \in \mathcal{F}} \frac{1}{n} \sum_{i=1}^n L(Y_i f(X_i)) + \lambda
||w||^2
$$

Where $\mathcal{F}$ is the linear hypothesis class, $L$ is the hinge loss, and
$\lambda=\frac{1}{2C}$

\medskip

Anyways, the Lagrangian thus is:
$$
L(w, b, \xi, \alpha, \beta) = 
\frac{1}{2} ||w||^2 + 
C \frac{1}{n} \sum_{i=1}^n \xi_i +
\sum_{i=1}^n \alpha_i(1 - \xi_i - Y_i (\langle w, x_i \rangle + b)) -
\sum_{i=1}^n \beta_i \xi_i
$$

Taking the gradient or partial derivative w/r/t $w, b, \xi$ yields:
$$
w = \sum_{i=1}^n \alpha_i Y_i x_i
$$
$$
\sum_{i=1}^n \alpha_i Y_i = 0
$$
$$
\sum_{i=1}\beta_i = \frac{C}{n} - \sum_{i=1}^n \alpha_i \to \beta_i = \frac{C}{n} - \alpha_i
$$

Since $\beta_i \geq 0$, we can deduce an upper limit for $\alpha_i$
$$
0 \leq \alpha_i \leq \frac{C}{n} \quad \forall i
$$

This results in:

$$
\max_{\alpha \in \mathbb{R}^n} 
\sum_{i=1}^n \alpha_i - \frac{1}{2} \sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j Y_i Y_j \langle x_i, x_j \rangle
$$
$$
s.t. \quad 0 \leq \alpha_i \leq \frac{C}{n} \quad \forall i
$$
$$
\quad \sum_{i=0}^n Y_i \alpha_i = 0
$$

All our previous observations about support vectors still hold.

\includegraphics[scale=0.25]{soft.png}
\end{document}

### 2018-03-11 hw1/hw1_1A.py

@@ -1,0 +1,53 @@
+# Tom Wallace
+# STAT 672
+# Spring 2018
+# Homework 1, Problem 1A
+
+import numpy as np
+
+# UnitBallRejectionSampling(d, n)
+#
+# Arguments: d (int, d >= 1)
+#            n (int, n >= 1)
+#
+# Description: draw n random vectors from d-dimensional unit ball. Does so by
+# generating a candidate vector, each coordinate of which is a random
+# draw from a uniform (-1,1) distribution, and rejecting any candidate vector
+# with a Euclidean norm >= 1
+#
+def UnitBallRejectionSampling(d, n):
+	
+	n_candidates = 0
+	n_accepted = 0
+
+	print("Generating", n, "random vectors of dimensionality", d, "using rejection sampling")
+	
+	while(n_accepted < n):
+
+		#generate empty vector of length d
+		vector = np.empty(d)
+
+		#fill with (-1,1) random variables
+		for i in range(0, d):
+			vector[i] = np.random.uniform(-1,1)
+
+		#update n_candidates
+		n_candidates += 1
+
+		#take Euclidean norm
+		euclidean_norm = np.linalg.norm(vector, ord=2) 
+		
+		#accept only if Euclidean norm <= 1
+		if(euclidean_norm <= 1):
+			print(vector)
+			n_accepted += 1
+		else:
+			pass
+
+	#calculate and print observed probability of acceptance
+	prob_accept = n_accepted / n_candidates
+	print("Probability of acceptance: ", prob_accept)
+
+
+#generate 100 samples with d=10
+UnitBallRejectionSampling(10,100)

### 2018-03-11 hw1/hw1_1B.py

@@ -1,0 +1,69 @@
+# Tom Wallace
+# STAT 672
+# Spring 18
+# Homework 1, Problem 1B
+
+import numpy as np
+
+# GenerateTheta(d)
+#
+# Arguments: d (int, d >= 1)
+#
+# Description: generates vector of dimensionality d filled with uniform random
+# variables from surface of d-dimensional sphere. Computed as Z / ||Z||_2 where
+# Z is a vector populated with d N(0,1) random variables.
+#
+# Returns: numpy vector
+#
+def GenerateTheta(d):
+
+	gaussians = np.random.randn(d)
+	
+	normalization = np.linalg.norm(gaussians, 2)
+
+	return(gaussians / normalization)
+
+# GenerateRadius(d)
+#
+# Arguments: d (int, d >= 1)
+# 
+# Description: generates scalar between 0 and 1. Calculated as U^(1/d), where U
+# is a uniform(0,1) random variable and d is dimensionality.
+#
+# Returns: real
+#
+def GenerateRadius(d):
+
+	u = np.random.uniform(0,1)
+
+	r = np.power(u, 1/d)
+
+	return(r)
+
+# PolarSampling(n, d):
+#
+# Arguments: n (int, n >= 1)
+#            d (int, d >= 1) 
+#
+# Description: generates n random vectors of the d-dimensional unit ball. Does
+# so by computing the product of theta and radius.
+#
+# Returns: nothing (prints random vectors to stdout)
+#
+def PolarSampling(d, n):
+
+	i = 0
+
+	print("Generating", n, "random vectors of dimensionality", d)
+
+	while(i < n):
+		
+		theta = GenerateTheta(d)
+		r = GenerateRadius(d)
+		vector = np.dot(theta, r)
+		i += 1
+		print(vector)
+
+
+#create 100 samples with d=10
+PolarSampling(10, 100)

### 2018-03-11 hw1/hw1_2A.py

@@ -1,0 +1,99 @@
+# Tom Wallace
+# STAT 672
+# Spring 2018
+# Homework 1, Problem 2A
+
+import numpy as np
+
+# GenerateXPlusVector(d)
+#
+# Arguments: d (integer, d >= 1)
+#
+# Description: generate vector of length d populated with N(5/sqrt(d), 4) Gaussians 
+#
+# Returns: numpy vector
+#
+def GenerateXPlusVector(d):
+
+	vector = np.empty(d)
+
+	for i in range(0, d):
+
+		vector[i] = np.random.normal(5/np.sqrt(d), 4)
+	
+	return(vector)
+
+# GenerateXMinusVector(d)
+#
+# Arguments: d (integer, d >= 1)
+#
+# Description: generate vector of length d populated with N(-5/sqrt(d), 1) Gaussians 
+#
+# Returns: numpy vector
+#
+def GenerateXMinusVector(d):
+
+	vector = np.empty(d)
+
+	for i in range(0, d):
+
+		vector[i] = np.random.normal(-5/np.sqrt(d), 1)
+
+	return(vector)
+
+
+# ComputeError(x_plus, x_minus, z)
+#
+# Arguments: x_plus (numpy vector)
+#            x_minus (numpy vector)
+#            z (numpy vector)
+#
+# Description: 
+#              - compute ||X_+ - Z||_2^2. Call this norm1.
+#              - compute ||X_- - Z||_2^2. Call this norm2.
+#              - if A >= B, return 1
+#              - else, return 0
+#
+# Returns: 0 | 1
+#
+def ComputeError(x_plus, x_minus, z):
+
+	norm1 = np.linalg.norm((x_plus - z), ord=2)
+	norm2 = np.linalg.norm((x_minus - z), ord=2) 
+
+	if(norm1 > norm2):
+		return(1)
+	else:
+		return(0)
+
+
+# RunSimulation(n, d)
+#
+# Arguments: n (integer, n >= 1)
+#            d (integer, d >= 1)
+#
+# Description: conducts n iterations of generating random vectors and comparing
+# them as per above functions.
+#
+# Returns: nothing (prints to stdout)
+#
+def RunSimulation(n, d):
+
+	frequency = 0
+
+	for i in range(0, n):
+		
+		x_plus = GenerateXPlusVector(d)
+		z = GenerateXPlusVector(d)
+		x_minus = GenerateXMinusVector(d)
+
[…]

### 2018-03-11 hw2/hw2.py

@@ -1,0 +1,228 @@
+import numpy as np
+from sklearn import linear_model
+from sklearn import metrics
+import matplotlib.pyplot as plt
+from scipy.interpolate import UnivariateSpline
+
+# Tom Wallace
+# STAT 672
+# Spring 2018
+# Homework 2
+
+#####                     #####
+#####     Problem 1 C     #####
+#####                     #####
+
+# use np.random.uniform() to generate X ~ uniform(0,1)
+
+# function to generate Y given X
+def GenerateYGivenX(x):
+
+	r = np.random.uniform()
+
+	if(x < 0.2 or x > 0.8):
+
+		if(r <= 0.9):
+			y = 1
+		else:
+			y = 0
+	else:
+		if(r <= 0.2):
+			y = 1
+		else:
+			y = 0
+
+	return(y)
+
+# vectorized version of Y-generating function
+v_GenerateYGivenX = np.vectorize(GenerateYGivenX)
+
+# run many iterations of the above functions
+# plot results to verify correctness
+
+# set random seed
+np.random.seed(888)
+
+# generate 10k x
+X = np.random.rand(10000)
+
+# generate 10k y | x
+Y = v_GenerateYGivenX(X)
+
+# round X to 2 digits
+v_round = np.vectorize(np.round)
+X = v_round(X, 2)
+
+# count number of y=1 for every x
+histX = np.arange(0, 1, 0.01)
+
+histY = np.zeros(len(histX))
+
+for i in range(0, len(histX)):
+	for j in range(0, len(X)):
+		if( histX[i] == X[j] and Y[j] == 1 ):
+			histY[i] = histY[i]+1
+# bin X's into 10 bins
+
+histX2 = np.arange(0, 1, 0.1)
+histY2 = np.zeros(10)
+
+histY2[0] = sum(histY[0:9])
+histY2[1] = sum(histY[10:19])
+histY2[2] = sum(histY[20:29])
+histY2[3] = sum(histY[30:39])
+histY2[4] = sum(histY[40:49])
+histY2[5] = sum(histY[50:59])
+histY2[6] = sum(histY[60:69])
+histY2[7] = sum(histY[70:79])
+histY2[8] = sum(histY[80:89])
+histY2[9] = sum(histY[90:99])
+
+# plot
+plt.bar(x=histX2, height=histY2, width=0.155555)
+plt.title("10,000 Simulations of Y|X")
[…]

### 2018-03-11 hw2/plot.R

@@ -1,0 +1,58 @@
+df = data.frame(
+		n = c(10, 20, 50, 100, 200, 500, 1000),
+		emp_error_1d = c(0.3, 0.345, 0.402, 0.4512, 0.4297, 0.4244, 0.4454),
+		gen_error_1d = c(0.476, 0.461, 0.4948, 0.468, 0.4673, 0.45388, 0.45954),
+		emp_error_2d = c(0.142, 0.16, 0.1832, 0.169, 0.1825, 0.1758, 0.17388),
+		gen_error_2d = c(0.268, 0.276, 0.2208, 0.1868, 0.1932, 0.18252, 0.17602),
+		emp_error_5d = c(0.108, 0.144, 0.1812, 0.181, 0.1804, 0.1754, 0.17966),
+		gen_error_5d = c(0.27, 0.236, 0.2056, 0.2028, 0.1927, 0.18176, 0.18154)
+		)
+
+png("plot.png", height = 3, width=6, units = "in", res = 300)
+par(mfrow=c(1,3))
+plot(df$emp_error_1d, 
+     col = "black", 
+     lty = 1, 
+     lwd = 2,
+     type="l", 
+     ylim = c(0, 0.5),
+     ylab = "Error",
+     main = "d=1",
+     xaxt = 'n', 
+     xlab = "N")
+lines(df$gen_error_1d, col = "black", lty = 2, lwd=2)
+axis(1, at=1:7, labels = c(10, 20, 50, 100, 200, 500, 1000))
+legend("bottomright", legend = c("emp_error", "gen_error"), lty = c(1,2), col =
+       "black", lwd = 2)
+
+plot(df$emp_error_2d, 
+     col = "blue", 
+     lty = 1, 
+     lwd = 2,
+     type="l", 
+     ylim = c(0, 0.5),
+     ylab = "Error",
+     main = "d=2",
+     xaxt = 'n', 
+     xlab = "N")
+lines(df$gen_error_2d, col = "blue", lty = 2, lwd=2)
+axis(1, at=1:7, labels = c(10, 20, 50, 100, 200, 500, 1000))
+legend("bottomright", legend = c("emp_error", "gen_error"), lty = c(1,2), col =
+       "blue", lwd = 2)
+
+plot(df$emp_error_5d, 
+     col = "red", 
+     lty = 1, 
+     lwd = 2,
+     type="l", 
+     ylim = c(0, 0.5),
+     ylab = "Error",
+     main = "d=5",
[…]

### 2018-03-11 hw3/hw3.py

@@ -1,0 +1,141 @@
+import scipy as sp
+import numpy as np
+from sklearn.preprocessing import PolynomialFeatures
+
+'''
+Note: I wrote the following function to generate the desired feature matrix
+containing all quadratic and interaction terms of X. However, after doing so I
+uncovered a scikitlearn function that does the exact same thing and so just use
+that
+
+def GeneratePhiMatrix(X):
+
+	# get dimensions of X
+	nrow = X.shape[0]
+	ncol = X.shape[1]
+
+	# preallocate final array 
+	# this is *much* more memory-efficient than alternate approaches
+	mat = np.zeros((nrow, int(ncol*2 + sp.special.comb(ncol, 2))))
+
+	# first "chunk" of final array is just X
+	mat[0:nrow, 0:ncol] = X
+
+	# second "chunk" of final array is quadratic form of X
+	mat[0:nrow, ncol+1:(2*ncol+1)] = pow(X, 2)
+
+	# third "chunk" of final array is interaction terms
+
+	# create empty list 
+	lst = list()
+
+	# generate all interaction terms 
+	# apppend to list
+	for i in range(0, ncol):
+		for j in range(i+1, ncol):
+			col = X[:,i] * X[:,j]
+			lst.append(col)
+	
+	# assign columns from list to preallocated array
+	for i in range(0, len(lst)):
+		col = lst[i]
+		mat[0:nrow, ncol*2+i] = col
+
+	return(mat)
+'''
+
+# function to create test set consisting of every i'th observation
+def GenerateTest(a, i):
+
+	# get number of rows in a
+	nrow = a.shape[0]
+
+	# select every i'th row as part of test
+	train = a[range(i-1, nrow+1, i)]
+
+	return(train)
+
+# function to create train set consisting of every observation not divisible by i
+def GenerateTrain(a, i):
+
+	# get number of rows in a
+	nrow = a.shape[0]
+
+	# list of numbers \leq number of rows and not divisible by i
+	lst = list(range(1, nrow+1))
+	for item in lst:
+		if item % i == 0:
+			lst.remove(item)
+	for item in lst:
+		item = item - 1
+	
+	# index observations not divisible by i
+	train = a[lst]
+
+	return(train)
+
+# function to center matrix
+def Center(a):
+
+	# compute column means
[…]

### 2018-03-12 hw3/hw3.py
more work on svd

@@ -1,13 +1,16 @@
+#TODO: email Slawski about scaling and centering - skl?
 import scipy as sp
 import numpy as np
-from sklearn.preprocessing import PolynomialFeatures
+from sklearn import preprocessing
+from sklearn.linear_model import Ridge
 
 '''
 Note: I wrote the following function to generate the desired feature matrix
 containing all quadratic and interaction terms of X. However, after doing so I
 uncovered a scikitlearn function that does the exact same thing and so just use
 that
-
+'''
+'''
 def GeneratePhiMatrix(X):
 
 	# get dimensions of X
@@ -74,58 +77,51 @@ def GenerateTrain(a, i):
 
 	return(train)
 
-# function to center matrix
-def Center(a):
+# function to center and scale matrix
+def CenterAndScale(a):
 
 	# compute column means
 	col_means = np.mean(a, axis=0)
 
-	# subtract col means from relevant cols
-	a = a - col_means
+	# subtract col means to center matrix
+	cen = a - col_means
 
-	return(a)
+	# compute L2 norms of columns of centered matrix
+	cen_norm = np.sqrt(np.sum(pow(cen, 2), axis = 0))
 
-# function to scale matrix
-def Scale(a):
+	# divide centered matrix by column norms
+	out = cen / cen_norm
+	
+	return(out)
 
-	# square matrix
-	s = pow(a, 2)
+# function to compute S_Lambda
+def ComputeS_Lambda(S, Lambda):
 
-	# compute column sums of squared matrix
-	s_squared = np.sum(s, axis = 0)
+	# number of coefficients 
+	# same as number of singular values
+	n = len(S)
 
-	# take square root of these sums
-	s_squared_root = np.sqrt(s_squared)
+	# make s_lambda matrix
+	
+	# initialize as n x n zero matrix
+	S_Lambda = np.zeros((n, n))
 
-	# divide original matrix by these square roots
-	a = a / s_squared_root
+	# fill in along diagonal
+	vals = S / (pow(S, 2) + Lambda)
+	np.fill_diagonal(S_Lambda, vals)
 
-	return(a)
+	return(S_Lambda)
 
-
-# generate lambda
-
[…]

### 2018-03-18 hw3/hw3.py
fuck yeah ridge regression!

@@ -2,7 +2,7 @@
 import scipy as sp
 import numpy as np
 from sklearn import preprocessing
-from sklearn.linear_model import Ridge
+from sklearn import linear_model
 
 '''
 Note: I wrote the following function to generate the desired feature matrix
@@ -77,24 +77,26 @@ def GenerateTrain(a, i):
 
 	return(train)
 
-# function to center and scale matrix
-def CenterAndScale(a):
+def Center(a):
 
 	# compute column means
 	col_means = np.mean(a, axis=0)
 
 	# subtract col means to center matrix
-	cen = a - col_means
+	out = a - col_means
 
-	# compute L2 norms of columns of centered matrix
-	cen_norm = np.sqrt(np.sum(pow(cen, 2), axis = 0))
+	return(out)
 
-	# divide centered matrix by column norms
-	out = cen / cen_norm
+def Scale(a):
+
+	# compute L2 norms of columns of matrix
+	norm = np.sqrt(np.sum(pow(a, 2), axis = 0))
+
+	# divide matrix by column norms
+	out = a / norm
 	
 	return(out)
 
-# function to compute S_Lambda
 def ComputeS_Lambda(S, Lambda):
 
 	# number of coefficients 
@@ -133,36 +135,28 @@ trainY = GenerateTrain(Y, 4)
 testPhi = GenerateTest(Phi, 4)
 testY = GenerateTest(Y, 4)
 
-# center and scale training X
-trainPhi = CenterAndScale(trainPhi)
+# center and scale training X and Y
+trainPhi = Scale(Center(trainPhi))
+trainY = Center(trainY)
 
 # list lambdas
 lambdas = list(np.arange(-13, 9.5, 0.5))
 
 # take SVD of testPhi
+U, S, V = np.linalg.svd(trainPhi, full_matrices = False)
 
-'''
-svd = np.linalg.svd(trainPhi)
-U = svd[0]
-S = svd[1]
-V = svd[2]
+# compute S_Lambda
 S_Lambda = ComputeS_Lambda(S, 1.0)
-'''
 
-print(U.shape)
-print(trainY.shape)
-print(S.shape)
-print(S_Lambda.shape)
-print(V.shape)
+# ridge regression coefficients can be calculated as:
+# V * S_Lambda * U.T * Y_train
+# this is most efficiently computed from right to left
 
[…]

### 2018-03-19 hw3/hw3.py
finished ridge

@@ -1,51 +1,6 @@
-#TODO: email Slawski about scaling and centering - skl?
-import scipy as sp
 import numpy as np
 from sklearn import preprocessing
 from sklearn import linear_model
-
-'''
-Note: I wrote the following function to generate the desired feature matrix
-containing all quadratic and interaction terms of X. However, after doing so I
-uncovered a scikitlearn function that does the exact same thing and so just use
-that
-'''
-'''
-def GeneratePhiMatrix(X):
-
-	# get dimensions of X
-	nrow = X.shape[0]
-	ncol = X.shape[1]
-
-	# preallocate final array 
-	# this is *much* more memory-efficient than alternate approaches
-	mat = np.zeros((nrow, int(ncol*2 + sp.special.comb(ncol, 2))))
-
-	# first "chunk" of final array is just X
-	mat[0:nrow, 0:ncol] = X
-
-	# second "chunk" of final array is quadratic form of X
-	mat[0:nrow, ncol+1:(2*ncol+1)] = pow(X, 2)
-
-	# third "chunk" of final array is interaction terms
-
-	# create empty list 
-	lst = list()
-
-	# generate all interaction terms 
-	# apppend to list
-	for i in range(0, ncol):
-		for j in range(i+1, ncol):
-			col = X[:,i] * X[:,j]
-			lst.append(col)
-	
-	# assign columns from list to preallocated array
-	for i in range(0, len(lst)):
-		col = lst[i]
-		mat[0:nrow, ncol*2+i] = col
-
-	return(mat)
-'''
 
 # function to create test set consisting of every i'th observation
 def GenerateTest(a, i):
@@ -114,49 +69,141 @@ def ComputeS_Lambda(S, Lambda):
 
 	return(S_Lambda)
 
-# read data
-X = np.genfromtxt('hw3_X.csv', delimiter = ',')
-Y = np.genfromtxt('hw3_Y.csv', delimiter = ',')
+def RidgeRegression(U, S_Lambda, V, Y):
+	
+	# ridge regression coefficients can be calculated as:
+	# V * S_Lambda * U.T * Y_train
+	# this is most efficiently computed from right to left
[…]

### 2018-03-20 hw3/hw3.py
finished lasso

@@ -1,8 +1,14 @@
+'''
+Tom Wallace <twalla11@masonlive.gmu.edu>
+STAT 672
+Spring 2017
+HW #3
+'''
+
 import numpy as np
 from sklearn import preprocessing
 from sklearn import linear_model
 
-# function to create test set consisting of every i'th observation
 def GenerateTest(a, i):
 
 	# get number of rows in a
@@ -13,7 +19,6 @@ def GenerateTest(a, i):
 
 	return(train)
 
-# function to create train set consisting of every observation not divisible by i
 def GenerateTrain(a, i):
 
 	# get number of rows in a
@@ -78,17 +83,17 @@ def RidgeRegression(U, S_Lambda, V, Y):
 	ridge_coeffs = np.matmul(V.T, np.matmul(S_Lambda, np.reshape(np.matmul(U.T, Y), (-1, 1))))
 	return(ridge_coeffs)
 
-def ScaleRidgeCoeffs(trainPhi, ridge_coeffs):
+def ScaleCoeffs(trainPhi, coeffs):
 
 	# compute L2 norms of columns of trainPhi
 	norm = np.reshape(np.sqrt(np.sum(pow(trainPhi, 2), axis = 0)), (-1,1))
 
-	# divide ridge coefficients by norm
-	scaled_ridge_coeffs = ridge_coeffs / norm
+	# divide coefficients by norm
+	scaled_coeffs = coeffs / norm
 
-	return(scaled_ridge_coeffs)
+	return(scaled_coeffs)
 
-def ScaledRidgeIntercept(trainY, scaled_ridge_coeffs, trainPhi):
+def ScaledIntercept(trainY, scaled_coeffs, trainPhi):
 
 	# calculate mean of Y
 	y_bar = np.mean(trainY, axis=0)
@@ -96,13 +101,13 @@ def ScaledRidgeIntercept(trainY, scaled_ridge_coeffs, trainPhi):
 	# calculate column-wise means of Phi
 	Phi_bar = np.reshape(np.mean(trainPhi, axis = 0), (-1,1))
 
-	# multiply scaled ridge coefficient j with Phi mean j
+	# multiply scaled coefficient j with Phi mean j
 	# and then sum
-	w0_scaled = y_bar - np.sum(Phi_bar * scaled_ridge_coeffs, axis = 0)
+	w0_scaled = y_bar - np.sum(Phi_bar * scaled_coeffs, axis = 0)
 
[…]

### 2018-03-21 hw3/hw3.tex
finished hw3

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{geometry}

\geometry{margin=1in}
\graphicspath{}

\title{STAT 672: Homework 3}
\author{Tom Wallace}

\begin{document}

\maketitle

\section*{SVD and Ridge Regression}

Estimated coefficients in ridge regression are given by:

$$
\hat{\bm{\beta}}_{\mathrm{ridge}} = (\mathbf{X}^T\mathbf{X} + \lambda
\mathbf{I})^{-1}\mathbf{X}^T\mathbf{y}
$$

We take the singular value decomposition of the feature matrix:

$$
= ((\mathbf{UDV}^T)^T(\mathbf{UDV}^T)+\lambda \mathbf{I})^{-1} (\mathbf{UDV}^T)^T \mathbf{y}
$$
$$
= (\mathbf{VD}^T\mathbf{U}^T\mathbf{UDV}^T + \lambda \mathbf{I})^{-1}
\mathbf{VD}^T\mathbf{U}^T \mathbf{y}
$$

We know that $\mathbf{U}^T\mathbf{U}=\mathbf{I}$ and $\mathbf{D}^T\mathbf{D}=\mathbf{D}^2$ and so can
simplify this to:
$$
= (\mathbf{VD}^2\mathbf{V}^T + \lambda \mathbf{I})^{-1} \mathbf{VDU}^T\mathbf{y}
$$

We substitute in $\mathbf{VV}^T$ for $\mathbf{I}$:

$$
= (\mathbf{VD}^2\mathbf{V}^T + \lambda \mathbf{VV}^T)^{-1} \mathbf{VDU}^T\mathbf{y}
$$

And factor out $\mathbf{VV}^T$:
$$
= \mathbf{V}(\mathbf{D}^2 + \lambda)^{-1}\mathbf{V}^T \mathbf{VDU}^T\mathbf{y}
$$

And make use of the fact that $\mathbf{V}$ is orthonormal:
$$
= \mathbf{V}(\mathbf{D}^2 + \lambda)^{-1}\mathbf{DU}^T\mathbf{y}
$$

Since $\mathbf{D}$ is diagonal, we can rewrite the expression involving it and
$\lambda$:
$$
\mathbf{D}_\lambda := (\mathbf{D}^2 + \lambda)^{-1}\mathbf{D}
$$
$$
= \mathrm{diag}\left(\frac{d_1}{d_1^2 + \lambda} \ldots \frac{d_D}{d_D^2 +
\lambda} \right)
$$

Thus, computation of estimated coefficients in ridge regression via SVD is
given by:
$$
\hat{\bm{\beta}}_{\mathrm{ridge}} = \mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}
$$
$$
\blacksquare
$$

\section*{Efficiency of Computation}

There are an inefficient method and an efficient method of re-calculating ridge
regression coefficients for a new regularization parameter $\lambda$.

In the inefficient method, we redo the entire singular value decomposition every
time we update $\lambda$. SVD has a computational complexity on the order of
$O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
$d$ corresponding to the number of columns, i.e. the dimensionality of the
data). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in
\mathbb{R}^{n \times p}$, costs $O(2mnp)$ flops. In our case, $\mathbf{U}^T \in
\mathbb{R}^{d \times n}$ and $\mathbf{Y} \in \mathbb{R}^{n \times 1}$, and so
multiplying the two costs $O(2 \times n \times d \times 1) = O(2nd)$ flops and results
in a $d \times 1$ matrix. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
\times d}$ and this $d \times 1$ matrix costs $O(2d^2)$ flops (assuming we do not take advantage of the
diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ matrix. 
Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ matrix costs 
$O(2d^2)$ flops. Adding together all these steps, we have $O(nd^2 + 2nd + 2d^2 +
2d^2)$. Dropping all constant coefficients and only considering the
highest-order polynomial, we conclude that the inefficient method costs
$O(nd^2)$ flops.

A more efficient method notes that $\mathbf{D}_\lambda$ is the only part of 
$\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$ that depends on $\lambda$
($\mathbf{U}$ and $\mathbf{V}$ depend solely on the feature matrix $\mathbf{X}$, and
$\mathbf{y}$  depends only on itself) and so we do not not need to recompute the
SVD for every new value of $\lambda$. Assume that we have pre-calculated and
cached $\mathbf{U}^T\mathbf{y}$ and $\mathbf{V}$. Modifying the $d$ non-zero
values of $\mathbf{D}_\lambda$ to reflect our new value of $\lambda$ costs $O(d)$ flops. 
Computing new regression coefficients requires multiplying the new $\mathbf{D}_\lambda$ and
$\mathbf{U}^T\mathbf{y}$, which costs $O(2d^2)$ flops, and then multiplying that
result by $\mathbf{V}$, which also costs $O(2d^2)$ flops. Adding together these
two steps, we have $O(d + 2d^2 + 2d^2)$. Dropping all constant coefficients and
only considering the highest-order polynomial, we conclude that the efficient method
costs $O(d^2)$ flops.

\begin{centering}
\begin{table}[h!]
	\begin{tabular}{|l|l|}
		\hline
		& \textbf{Complexity} \\
		\textbf{Inefficient method} & $O(nd^2)$ \\
		\textbf{Efficient method} & $O(d^2)$ \\
		\hline
	\end{tabular}
\end{table}
\end{centering}

\section*{Results}

Below, I compare the performance of ridge and lasso regression on the homework
dataset for different values of $\lambda$. For ridge regression, a
regularization parameter value of 0.00390625 achieves the lowest test error
(2.81855); for lasso regression, a regularization parameter of $4.46e^{-0.6}$
(i.e. the tested value closest to 0) achieves the lowest test error (2.85948).
The ridge method achieves the better minimum test error.

\includegraphics[scale=0.5]{foo}

\end{document}

### 2018-03-21 hw3/hw3.tex
final commit

@@
 
-In the inefficient method, we redo the entire singular value decomposition every
-time we update $\lambda$. SVD has a computational complexity on the order of
+In the inefficient method, we recompute the SVD every
+time we update $\lambda$. SVD has complexity on the order of
 $O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
-$d$ corresponding to the number of columns, i.e. the dimensionality of the
-data). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
+$d$ corresponding to the number of columns). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
 A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in
-\mathbb{R}^{n \times p}$, costs $O(2mnp)$ flops. In our case, $\mathbf{U}^T \in
+\mathbb{R}^{n \times p}$, costs $2mnp$ flops. In our case, $\mathbf{U}^T \in
 \mathbb{R}^{d \times n}$ and $\mathbf{Y} \in \mathbb{R}^{n \times 1}$, and so
-multiplying the two costs $O(2 \times n \times d \times 1) = O(2nd)$ flops and results
-in a $d \times 1$ matrix. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
-\times d}$ and this $d \times 1$ matrix costs $O(2d^2)$ flops (assuming we do not take advantage of the
-diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ matrix. 
-Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ matrix costs 
-$O(2d^2)$ flops. Adding together all these steps, we have $O(nd^2 + 2nd + 2d^2 +
-2d^2)$. Dropping all constant coefficients and only considering the
+multiplying the two costs $2 \times n \times d \times 1 = 2nd$ flops and results
+in a $d \times 1$ vector. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
+\times d}$ and this $d \times 1$ vector costs $d$ flops (assuming we take advantage of the
+diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ vector. 
+Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ vector costs 
+$2d^2$ flops. Adding together all these steps, we have $nd^2 + 2nd + d +
+2d^2$ flops. Dropping all constant coefficients and only considering the
 highest-order polynomial, we conclude that the inefficient method costs
@@
 $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$ that depends on $\lambda$
-($\mathbf{U}$ and $\mathbf{V}$ depend solely on the feature matrix $\mathbf{X}$, and
-$\mathbf{y}$  depends only on itself) and so we do not not need to recompute the
+and so we do not need to recompute the
 SVD for every new value of $\lambda$. Assume that we have pre-calculated and
 cached $\mathbf{U}^T\mathbf{y}$ and $\mathbf{V}$. Modifying the $d$ non-zero
-values of $\mathbf{D}_\lambda$ to reflect our new value of $\lambda$ costs $O(d)$ flops. 
+values of $\mathbf{D}_\lambda$ to reflect our new value of $\lambda$ costs
+$2d$ flops ($d$ flops for addition of the new $\lambda$ in the denominator, 
+and $d$ flops to
+divide the numerator by the new denominator). 
 Computing new regression coefficients requires multiplying the new $\mathbf{D}_\lambda$ and
-$\mathbf{U}^T\mathbf{y}$, which costs $O(2d^2)$ flops, and then multiplying that
-result by $\mathbf{V}$, which also costs $O(2d^2)$ flops. Adding together these
-two steps, we have $O(d + 2d^2 + 2d^2)$. Dropping all constant coefficients and
+$\mathbf{U}^T\mathbf{y}$, which costs $2d^2$ flops, and then multiplying that
+result by $\mathbf{V}$, which also costs $2d^2$ flops. Adding together these
+two steps, we have $2d + 2d^2 + 2d^2$ flops. Dropping all constant coefficients and
 only considering the highest-order polynomial, we conclude that the efficient method
@@
 
-\begin{centering}
-\begin{table}[h!]
-	\begin{tabular}{|l|l|}
-		\hline
-		& \textbf{Complexity} \\
-		\textbf{Inefficient method} & $O(nd^2)$ \\
-		\textbf{Efficient method} & $O(d^2)$ \\
-		\hline
-	\end{tabular}
-\end{table}
-\end{centering}
+In practical terms, since we are fitting regression coefficients on a training
+dataset with $n=6000$, the efficient method is 6,000 times faster.
 
@@
 
+In the below graph, I have omitted some tested values of $\lambda$ to make
+things more visually clear.
+
 \includegraphics[scale=0.5]{foo}

### 2018-03-21 hw3/hw3.py

@@ -4,11 +4,11 @@ STAT 672
 Spring 2017
 HW #3
 '''
-
 import numpy as np
 from sklearn import preprocessing
 from sklearn import linear_model
 
+# Function to generate test set consisting of every i'th observation
 def GenerateTest(a, i):
 
 	# get number of rows in a
@@ -19,6 +19,7 @@ def GenerateTest(a, i):
 
 	return(train)
 
+# Function to generate training set consisting of every non-i'th observation
 def GenerateTrain(a, i):
 
 	# get number of rows in a
@@ -37,6 +38,7 @@ def GenerateTrain(a, i):
 
 	return(train)
 
+# Function to center matrix by subtracting column means
 def Center(a):
 
 	# compute column means
@@ -47,6 +49,7 @@ def Center(a):
 
 	return(out)
 
+# Function to scale matrix by dividing by column norms
 def Scale(a):
 
 	# compute L2 norms of columns of matrix
@@ -57,6 +60,10 @@ def Scale(a):
 	
 	return(out)
 
+# Function to compute S_Lambda
+# Where S refers to USV' resulting from a SVD
+# And Lambda refers to the regularization parameter of ridge regression
+# S_Lambda is a diagonal matrix with non-zero values d_i / d_ii^2 + Lambda
 def ComputeS_Lambda(S, Lambda):
 
 	# number of coefficients 
@@ -74,6 +81,7 @@ def ComputeS_Lambda(S, Lambda):
 
 	return(S_Lambda)
 
+# Function to conduct ridge regression via SVD
 def RidgeRegression(U, S_Lambda, V, Y):
 	
 	# ridge regression coefficients can be calculated as:
@@ -83,6 +91,7 @@ def RidgeRegression(U, S_Lambda, V, Y):
 	ridge_coeffs = np.matmul(V.T, np.matmul(S_Lambda, np.reshape(np.matmul(U.T, Y), (-1, 1))))
 	return(ridge_coeffs)
 
+# Function to re-scale coefficients calculated on scaled data
 def ScaleCoeffs(trainPhi, coeffs):
 
 	# compute L2 norms of columns of trainPhi
@@ -93,6 +102,7 @@ def ScaleCoeffs(trainPhi, coeffs):
 
 	return(scaled_coeffs)
 
+# Calculate re-scaled intercept 
 def ScaledIntercept(trainY, scaled_coeffs, trainPhi):
 
[…]

### 2018-03-21 hw3/hw3.tex
finished final presentation

@@
 
-We know that $\mathbf{U}^T\mathbf{U}=\mathbf{I}$ and $\mathbf{D}^T\mathbf{D}=\mathbf{D}^2$ and so can
-simplify this to:
+$\mathbf{U}$ is orthogonal ($\mathbf{U}^T\mathbf{U}=\mathbf{I}$) and
+$\mathbf{D}$ is diagonal ($\mathbf{D}^T = \mathbf{D}$):
 $$
@@
 
-We substitute in $\mathbf{VV}^T$ for $\mathbf{I}$:
+Because $\mathbf{V}$ is orthonormal, i.e. $\mathbf{V}^T = \mathbf{V}^{-1}$, we substitute in $\mathbf{VV}^T$ for $\mathbf{I}$:
 
@@
 
-And make use of the fact that $\mathbf{V}$ is orthonormal:
+And again make use of the fact that $\mathbf{V}$ is orthonormal, this time to
+cancel some terms:
 $$
@@
 $O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
-$d$ corresponding to the number of columns). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
+$d$ corresponding to the number of columns). We then multiply
+$\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$. We exploit the associative
+property of matrices and proceed right to left for maximum efficiency, 
+i.e. $\mathbf{V}(\mathbf{D}_\lambda(\mathbf{U}^T\mathbf{y}))$.
 A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in

### 2018-03-21 sgd/sgd.tex
finished final presentation

@@
 \usetheme{Madrid}
+\usecolortheme{spruce}
 
@@
 
+\usepackage{color}
+
 \usepackage{graphicx}
-\graphicspath{ {img/} }
+
+\graphicspath{{./img/}}
 
@@
 	\begin{itemize}
-		\item \small Example: in OLS, we do not need to check every possible value
-			of $\hat{\beta}$ to see if it minimizes the loss
+		\item \small Example: in OLS, we do not need to try
+			different values
+			of $\hat{\beta}$ to see which minimizes the loss
 			function, we (typically) can just evaluate
@@
 		\item Feature vector $\mathbf{X}_i$, label $Y_i$
-		\item Want to find best prediction function $f^*$ from class $\mathcal{F}$
-		\item $f$ has parameter vector $\bm{\theta}$
-		\item Optimization: pick values for $\bm{\theta}$ that minimize
-			empirical risk according to some
-			loss function $L$
-		\item Assume $L$ is convex (if not, problem is \textit{much}
-			harder)\\~\\
+		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
+		\item Optimization: pick weights $\bm{w}$ that minimize
+			empirical risk according to some convex
+			loss function $L(f_w(x), y)$ \\~\\
 	\end{itemize}
@@
 \begin{frame}
-	\frametitle{Gradient descent is a numerical approach to optimization}
-	Intuition: gradient is multidimensional version of 1st derivative
-	\begin{itemize}
-		\item \small $\nabla f(\mathbf{X}) = (\frac{\partial f}{x_1},
-			\frac{\partial f}{x_2}...)$
-		\item So if the gradient is (approximately) 0 at some point, that point is
-			(approximately) a stationary point 
-		\item Since we know the loss function is convex, that stationary
-			point must be the global minimum\\~\\
-	\end{itemize}
-
-	Give basic idea: take guess, step, evaluate, stop once below epsilon
+	\frametitle{Gradient descent is an iterative search procedure}
+	\begin{figure}[h]
+	\centering
+	\includegraphics[scale=0.33]{Sketch}
+	\end{figure}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Batch gradient descent, animated}
+	\frametitle{A more formal explanation}
+
+	$\mathbf{w}_{t+1} = \mathbf{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
+	\nabla_w L(f_w(\mathbf{X}_i), y_i)$ \\~\\
+
+	Stop if $\mathbf{w}_t - \ldots \leq \epsilon$ \\~\\
+
+	Step size $\gamma$ can vary over time \\~\\
+
+	Theoretical guarantees on speed of convergence ($\rho :=$ size of error):
+	\begin{itemize}
+		\item Version presented here: $ - \log \rho \sim t$
+		\item More optimized version: $ - \log \log \rho \sim t$ \\~\\
+	\end{itemize}
+
+	But, big difference between:
+	\begin{itemize}
+		\item Speed $:=$ number of iterations
+		\item Speed $:=$ time (clock on wall)
+	\end{itemize}
 \end{frame}
@@
 	\frametitle{Batch gradient descent is computationally expensive}
-	Elaborate
+	In ``plain'' (batch) gradient descent, for \textcolor{red}{every step}, 
+	we have to evaluate the gradient at \textcolor{blue} {every observation}
+
+	$$
+	\textcolor{red}{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \frac{1}{n}
+	\textcolor{blue}{\sum_{i=1}^n}
+	\nabla_w L(f_w(\mathbf{X}_i), y_i)
+	$$\\~\\
+
+	This becomes computationally intractable as $n$ grows large \\~\\
+
+	Knowing that the quality of our approximation gets linearly or
+	quadratically better with $t$ is not comforting if each $t$ takes days to run
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Stochastic gradient descent (SGD) is more efficient}
-	Overview of how it works \\~\\
+	\frametitle{Stochastic gradient descent (SGD) takes less time}
+
+	For each step, gradient is computed for a \textbf{single} randomly
+	chosen observation $i$:
+
+	$$
+	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w L(f_w(\mathbf{X}_i), y_i)
+	$$ \\~\\
+
+	This simplification makes approximation much ``noisier'', and hence SGD
+	requires more iterations \\~\\
+
+	But, each iteration is faster and so SGD can reach a predefined
+	level of risk or error in less time
 \end{frame}
@@
 \begin{frame}
-	\frametitle{SGD, visualized}
-\end{frame}
+	\frametitle{SGD is particularly useful when $n$ is large and computation
+	time is important}
+	\begin{table}[t]
+		\begin{tabular}{|l|c|c|}
+			\hline
+			& \textbf{GD} & \textbf{SGD} \\
+			\textbf{Time to accuracy} $\mathbf{\rho}$ & $n \log
+			\frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
+			\hline
+		\end{tabular}
+	\end{table}
 
-\begin{frame}
-	\frametitle{Choice of hyper-parameters is important}
-	Reminder of what they are \\~\\
-	Step size from Bottou \\~\\
-	Learning parameter
-\end{frame}
-
-\begin{frame}
-	\frametitle{SGD has nice properties for high-dimensional
-	data}
-	Theoretical explanation \\~\\
-	Empirical performance \\~\\
+	\begin{figure}
+	\centering
+	\includegraphics[scale=0.55]{comparison}
+	\end{figure}
 
@@
 \begin{frame}
-	\frametitle{Applications of SGD}
+	\frametitle{SGD is widely used in industry}
 	If a Silicon Valley press release uses any of the following phrases...
@@
 
-	...SGD probably is involved. Example: Google's \texttt{AlphaGo} program.
+	...SGD probably is involved. Example: Google's \textbf{AlphaGo} program.
 
@@
 	\centering
-	\includegraphics[scale=0.33]{go}
+	\includegraphics[scale=0.25]{go}
 	\end{figure}

### 2018-03-21 sgd/graph.R

@@ -1,0 +1,22 @@
+wd = getwd()
+setwd("~/code/stat672new/sgd/img/")
+png("comparison.png", 
+    height = 300,
+    width = 300)
+curve(x * log(1 / 0.05),
+      0, 
+      100,
+      col = "red",
+      main = "Time to (fixed) accuracy",
+      lwd = 2,
+      xlab = "n",
+      ylab = "Time")
+abline(h=1/0.05,
+       col = "blue",
+       lwd = 2)
+legend("topleft",
+       lwd = 2,
+       col = c("red", "blue"),
+       legend = c("GD", "SGD"))
+dev.off()
+setwd(wd)

### 2018-03-21 sgd/sgd.tex
finished final presentation

@@
 
-	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice'':
+	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice''
 	\begin{itemize}
@@
 	\frametitle{Other times, optimization is not so easy}
-	Suppose that we have a typical supervised classification problem:
+	Suppose that we have a typical supervised classification problem
 	\begin{itemize}
@@
 
-	Stop if $\mathbf{w}_t - \ldots \leq \epsilon$ \\~\\
+	Stop if gradient $\leq \epsilon$ \\~\\
 

### 2018-03-31 hw4/hw4.py
initial hw4

@@ -1,0 +1,1 @@
+xTrain = 

### 2018-03-31 hw4/hw4.py
finished code

@@ -1,1 +1,95 @@
-xTrain = 
+import numpy as np
+from sklearn.svm import SVC
+from sklearn.metrics import confusion_matrix
+
+def Center(a):
+
+	# compute column means
+	col_means = np.mean(a, axis=0)
+
+	# subtract col means to center matrix
+	out = a - col_means
+
+	return(out)
+
+def Scale(a):
+
+	# compute L2 norms of columns of matrix
+	norm = np.sqrt(np.sum(pow(a, 2), axis = 0))
+
+	# divide matrix by column norms
+	out = a / norm
+	
+	return(out)
+
+def RunAnalysis():
+
+	# read data
+	xTrain = np.genfromtxt('spam_Xtrain.csv', delimiter = ',')
+	xTest = np.genfromtxt('spam_Xtest.csv', delimiter = ',')
+	yTrain = np.genfromtxt('spam_Ytrain.csv', delimiter = ',')
+	yTest = np.genfromtxt('spam_Ytest.csv', delimiter = ',')
+
+	# read feature labels
+	with open('featurenames', 'r') as f:
+		labels = f.readlines()
+
+	# center and scale data
+	xTrain_cs = Scale(Center(xTrain))
+	xTest_cs = Scale(Center(xTest))
+
+	# regularization parameters to test
+	C = pow(10, np.arange(-1, 6.5, 0.5))
+
+	# fit SVM with different reg. parameter
+	for c in C:
+
+		print("SVM, regularization parameter =", c)
+		print("")
+
+		# initialize SVC
+		svm = SVC(c, kernel = "linear")
+
+		# fit on training data
+		svm.fit(xTrain_cs, yTrain)
+
+		# number of support vectors
+		print("Number of support vectors")
+		print(np.sum(svm.n_support_))
+		print("")
+
+		# training error
+		print("Training error")
+		print(1 - svm.score(xTrain_cs, yTrain))
+		print("")
+
+		# test error
+		print("Test error")
+		print(1 - svm.score(xTest_cs, yTest))
+		print("")
+
+		# calculate fp and tp
+		y_pred = svm.predict(xTest_cs)
+		y_true = yTest
+		tn, fp, fn, tp = confusion_matrix(y_true, y_pred).ravel()
+		n_test = np.shape(y_true)[0]
+
+		# false positives
+		print("False positive rate, test set")
+		print(fp / n_test)
+		print("")
+
+		# true positives
+		print("True positive rate, test set")
+		print(tp / n_test)
+		print("")
+
[…]

### 2018-04-01 hw4/hw4.tex
finished hw4

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{geometry}
\geometry{margin=1.1in}

\title{STAT 672: Homework 4}
\author{Tom Wallace}

\begin{document}

\maketitle

\section*{A}

See \texttt{hw4.py} for code.

It is recommended to scale data when using SVM because if we do not, we
essentially penalize each feature differently. To quote Hsu, Chang, and Lin
(2008), "the main advantage [of scaling] is to avoid attributes in greater
numeric ranges dominating those in smaller numeric ranges" \cite{hsu}.

This can be shown analytically. Consider a typical soft-margin SVM
classification problem.
$$
\min_{w_0, w} C \frac{1}{n} \sum_{i=1}^n \max(0, 1 - Y_i(w_0 +
\langle w, \Phi(X_i) \rangle)) + \frac{1}{2}||w||^2_2
$$

Let $\lambda = \frac{1}{2C}$ and assume for convenience with no loss of
generality that centering has occurred and so no intercepts are needed.

$$
\min_{w} \frac{1}{n} \sum_{i=1}^n \max(0, 1 - Y_i(
\langle w, \Phi(X_i) \rangle)) + \lambda||w||^2_2
$$

Consider a diagonal matrix $S$ whose non-zero entries consist of the
respective L2 norms of the columns of $\Phi$:

$$
S := \mathrm{diag}(||\Phi_{\bullet 1}||_2 \ldots ||\Phi_{\bullet D}||_2)
$$

Restate the SVM using $S$:

$$
\min_{w} \frac{1}{n} \sum_{i=1}^n \max(0, 1 - Y_i(
\langle wS, S^{-1}\Phi(X_i) \rangle)) + \lambda||w||^2_2
$$

Define $\theta = wS$:
$$
\min_{\theta} \frac{1}{n} \sum_{i=1}^n \max(0, 1 - Y_i(
\langle \theta, S^{-1}\Phi(X_i) \rangle)) + \lambda||\theta S^{-1}||^2_2
$$

We now can see that feature with a small numerical scale are penalized more
heavily (conversely, features with a large numerical scale will have greater
emphasis in the final classifier).

$$
= 
\min_{\theta} \frac{1}{n} \sum_{i=1}^n \max(0, 1 - Y_i(
\langle \theta, S^{-1}\Phi(X_i) \rangle)) + \lambda \sum_{j=1}^D
\frac{w_j^2}{||\Phi_{\bullet j}||^2_2}
$$

\section*{B}

See \texttt{hw4.py} and \texttt{output} for code and output.

\section*{C}

The number of support vectors decreases as C increases, as depicted in Table
1.

Explanation: C is the penalty on slack variables $\zeta$, which capture the
degree to which the margin constraint is violated. A small value of C
means a small penalty on margin violations, and will result in a relatively wide margin 
and many points laying on or inside of the margin (and hence a high number of support vectors). 
A large value of C means a
large penalty on margin violations, and so the separating hyperplane will have a
much tighter margin, meaning fewer
points that lie on or in the margin and hence a smaller number of support
vectors.

The above explanation also covers why run time for training increases with C,
eventually resulting in non-convergence.
When C is small, the separating
hyperplane is easy to find since margin violations are only
lightly punished, and so training run times are low. As C becomes larger, the
SVM tries more and more to find a separating hyperplane that doesn't result in
margin violations, and requires more calculation (run time) to find. 
Once C attains a sufficiently large value, the SVM may fail to converge because 
margin violations are unacceptable (due to the high value of C) but the data are 
not linearly separable and so some margin violation must be accepted, resulting
in a catch-22 that the optimization routine cannot overcome.

\begin{table}[h!]
	\caption{Influence of C on number of support vectors}
	\centering
	\begin{tabular}{|r r|}
		\hline
		\textbf{C} & \textbf{n(SV)} \\
		$10^{-1}$ & 1818 \\
		$10^{-0.5}$ & 1777 \\
		$10^{0}$ & 1443 \\
		$10^{0.5}$ & 1083 \\
		$10^{1}$ & 847 \\
		$10^{1.5}$ & 676 \\
		$10^{2}$ & 576 \\
		$10^{2.5}$ & 511 \\
		$10^{3}$ & 478 \\
		$10^{3.5}$ & 456 \\
		$10^{4}$ & 448 \\
		$10^{4.5}$ & 448 \\
		$10^{5}$ & 439 \\
		$10^{5.5}$ & 438 \\
		$10^{6}$ & 441 \\
		\hline
	\end{tabular}
\end{table}

\section*{D}

$w^*$ can be interpreted, but one must be cautious. A positive $w$
means the corresponding feature is associated with the positive class, while a
negative $w$ means the feature is associated with the negative class. The size
of $w$, within the context of a particular model, indicates the magnitude of
this association (e.g., if one $w$ is much larger in absolute value than other $w$, it is
strongly associated with a class, and probably is very useful for prediction). 
However, the caveat ``within a particular model'' is
crucial. Different regularization parameters will produce different coefficients
for identical datasets. We cannot make the sort of specific
interpetation that is possible with some parametric models, e.g. in linear
regression ``every one inch increase in height $X_1$ is associated with a $\beta_1$
increase in weight (lbs)."

Table 2 depicts the five largest and five smallest weights obtained with
$C=10^{4.5}$. We can interpret these weights. Emails that TALK IN ALL CAPS and
mention money (U.S. dollars) are especially likely to be categorized as spam.
Emails in which the sender mentions the name of someone that we presumably both
know (George) and refers to meeting
(presumably in-person) are indicative of a ``real'' relationship and hence 
are especially likely to be categorized as not-spam.

\begin{table}[h!]
	\caption{Largest and smallest weights}
	\centering
	\begin{tabular}{|l r | l r|}
		\hline
		\textbf{Feature} & \textbf{Weight} & \textbf{Feature} &
		\textbf{Weight} \\
		3d & 177.47 & george & -342.13 \\
		capital\_run\_length\_average & 144.75 & hp & -162.41 \\
		\$ & 45.77 & 415 & -161.59 \\
		857 & 33.01 & cs & -140.11 \\
		000 & 30.43 & meeting & -56.00 \\
		\hline
	\end{tabular}
\end{table}

\section*{E}

As shown in Table 3, relatively small values of C are associated with false
positive rates about 1\%. If we feel very strongly about staying below 1\%,
C=$10^{-0.5}$ has a FP rate of about 0.05\%. If we are willing to go slightly
above 1\%, C=$10^{0}$ has a FP rate of about 2\% and has over double the TP
rate.

\begin{table}[h!]
	\caption{True and false positive rates for selected C}
	\centering
	\begin{tabular}{|r r r|}
		\hline
		\textbf{C} & \textbf{FP rate} & \textbf{TP rate} \\
		$10^{-1}$ & 0.00 & 0.0004 \\
		$10^{-0.5}$ & 0.005 & 0.121 \\
		$10^{0}$ & 0.020 & 0.256 \\
		$10^{0.5}$ & 0.026 & 0.309 \\
		$10^{1}$ & 0.033 & 0.0336 \\
		\hline
	\end{tabular}
\end{table}

\section*{F}

We can modify the soft-margin SVM to incorporate different costs for different
forms of misclassification. Our typical set-up assigns the same cost C to margin
violations (e.g., misclassification) for both classes:

$$
\min_{w \in \mathbb{R}^d, w_0 \in \mathbb{R}, \xi \in \mathbb{R}^n} 
\frac{1}{2} ||w||^2 + C \frac{1}{n} \sum_{i=1}^n \xi_i
$$
$$
s.t. \quad Y_i(\langle w, x_i \rangle + w_0) \geq 1 - \xi_i \quad \forall i
$$
$$
\quad \xi_i \geq 0 \quad \forall i
$$

We can instead choose to assign different costs $C_+$ and $C_-$ for margin
violations by the positive and negative classes, respectively.

$$
\min_{w \in \mathbb{R^d}, w_0 \in \mathbb{R}, \xi \in \mathbb{R}^n} 
\frac{1}{2} ||w||^2 + C_+ \frac{1}{n_+} \sum_{i | y_i = 1} \xi_i + 
C_- \frac{1}{n_-} \sum_{i | y_i = -1} \xi_i
$$

In our case, the negative class corresponds to not-spam while the positive class
corresponds to spam. We feel that it is highly undesirable to miss important
email. We can operationalize this by more heavily penalizing margin violations
by the not-spam class.

$$
C_- >> C_+
$$

\begin{thebibliography}{2}
	\bibitem{hsu}
		Chih-Wei Hsu, Chih-Chung Chang, and Chih-Jen Lin.
		\textit{A Practical Guide to Support Vector Classification}.
		2016.

\end{thebibliography}
\end{document}

### 2018-04-02 hw4/hw4.tex
kernel space notes

@@
 mention money (U.S. dollars) are especially likely to be categorized as spam.
-Emails in which the sender mentions the name of someone that we presumably both
-know (George) and refers to meeting
+Emails that mention a person that both parties presumably know (George) and that
+refer to meeting
 (presumably in-person) are indicative of a ``real'' relationship and hence 

### 2018-04-02 week7-8/week7.tex
kernel space notes

\documentclass{article}

\begin{document}
A basic way to measure accuracy is the \textbf{confusion matrix}:

\begin{tabular}{r r r r}
$y$ & $\hat{y}$ & & \\
1 & 1 & & True positive \\
1 & -1 & & False negative \\
-1 & 1 & & False positive \\
-1 & -1 & & True negative \\
\end{tabular}

\medskip
Synonyms for TP: \textbf{recall}, \textbf{sensitivity}

\medskip
Synonyms for TN: \textbf{specificity}

\medskip
Positive predictive value = precision = $\frac{TP}{TP+FP}$

\medskip
Negative predictive value = $\frac{TN}{TN+FN}$

\medskip
Misclassification rate = $\frac{FP + FN}{n}$

\medskip
But misclassification rate is bad for unbalanced classes (e.g. not helpful to
always predict that a patient does not have a rare form of cancer). Better
measure: \textbf{F-score}
$$
F = \frac{2 (PPV + TPR)}{PPV + TPR}
$$
\medskip
Also better: ROC curve

\medskip
\underline{Proof of why using validation sets will produce the optimal
model}

\medskip

\textbf{Notation}

\smallskip

\begin{tabular}{l l}
$\{f_\mu\}_{\mu \in \mathcal{M}}$ & Models under consideration \\
$|\mathcal{M}|$ & Cardinality of model class (e.g. how many parameters) \\
$\hat{R}_m()$ & misclassification error of model on validation set of size $m$ \\
$R()$ & expected misclassification error of model \\
$f_{\hat{\mu}}$ & model achieving lowest validation error \\
\end{tabular}

\medskip

\textbf{What we want to show}: with probability of at least $1-\delta$,
$$
R(f_{\hat{\mu}}) \leq \min_{\mu \in \mathcal{M}} R(f_\mu) +
\sqrt{\frac{2}{m}\log\left(\frac{2|\mathcal{M}|}{\delta} \right)}
$$

i.e. as $m$ (the size of validation set) grows, the expected error of the model
achieving the lowest validation error attains that of the theoretically optimum
model.

\medskip

\textbf{How we show it}:

\medskip

Start with the difference in terms of expected risk between the model achieving lowest error on the
validation set (henceforth ``empirically best model'')and the model achieving
lowest expected error (henceforth ``theoretically best model'').

$$
R(f_{\hat{\mu}}) - \min_{\mu \in \mathcal{M}} R(f_\mu)
$$

Add in canceling terms reflecting the empirical error on validation set of size
$m$ of the empirically best model.

$$
= R(f_{\hat{\mu}}) -  \hat{R}_m(f_{\hat{\mu}}) + \hat{R}_m(f_{\hat{\mu}})- \min_{\mu \in \mathcal{M}} R(f_\mu)
$$

Because we have defined $f_{\hat{\mu}}$ to be the empirically best model, the
difference between its empirical risk and the expected risk of the theoretically
best model must be less than or equal to the largest possible empirical vs.
theoretical risk gap (i.e., the model $\mu$ producing the greatest difference
between its empirical and expected risk).

$$
\leq R(f_{\hat{\mu}}) -  \hat{R}_m(f_{\hat{\mu}}) + \max_{\mu \in
\mathcal{M}}(\hat{R}_m(f_\mu) -  R(f_\mu))
$$

And again by definition, the empirical vs. theoretical risk gap of the empirically best model
must be less than the empirical vs. theoretical risk gap of the model with the
largest such gap.

$$
\leq 2 \max_{\mu \in \mathcal{M}}|\hat{R}_m(f_\mu) -  R(f_\mu)|
$$

Now, consider the probability of that largest possible empirical vs. expected error gap 
being greater than some arbitrary amount:

$$
P( \max_{\mu \in \mathcal{M}}|\hat{R}_m(f_\mu) -  R(f_\mu)| > \epsilon)
$$

There may be multiple possible models producing that event:

$$
= P\left(\bigcup_{\mu \in \mathcal{M}} \{ |\hat{R}_m(f_\mu) -  R(f_\mu)| >
\epsilon)\}\right)
$$

Boole's inequality states that $P\left( \cup_i A_i \right) \leq \sum_i
P(A_i)$. Thus, the previous inequality must be less than or equal to its summed
version

$$
\leq \sum_{\mu \in \mathcal{M}} P(|\hat{R}_m(f_\mu) -  R(f_\mu)| > \epsilon)
$$

Hoeffding's inequality is $P(|\bar{S}_m - E(\bar{S}_m| \geq \epsilon) \leq 2
\exp (-2m\epsilon^2 / (b-a)^2)$. We are using Bernoulli variables (error
calculating using indicator function for whether prediction is correct or not)
so $a-b=1$. Thus:

$$
P(|\hat{R}_m(f_\mu) -  R(f_\mu)| > \epsilon) \leq 2 \exp (-2m\epsilon^2)
$$

We plug this back into the previous summation:

$$
\sum_{\mu \in \mathcal{M}} P(|\hat{R}_m(f_\mu) -  R(f_\mu)| > \epsilon) \leq  \sum_{\mu \in \mathcal{M}} 2 \exp (-2m\epsilon^2)
$$
$$
=2|\mathcal{M}|\exp(-2m\epsilon^2)
$$

Use  
$$\epsilon = \sqrt{\frac{\log \left( \frac{2|\mathcal{M}|}{\delta}\right)}{2m}} $$

Then, we plug this value into our previous expression
$$
2|\mathcal{M}|\exp(-2m\epsilon^2)
$$
$$
= 2 |\mathcal{M}|\exp \left( \frac{-2m \log \left( \frac{2|\mathcal{M}|}{\delta}\right)}{2m}
\right)
$$
$$
= \delta
$$

Going all the way back to the start, we now can say

\bigskip

\textbf{Mallow's CP}

Is a thing, but don't feel compelled to take detailed notes on it

\end{document}

### 2018-04-02 week7-8/week8.tex
kernel space notes

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{geometry}
\geometry{margin=1.5in}

\usepackage{graphicx}

\begin{document}

\textbf{What is a kernel?}

To classify, we need some measure of similarity. Let's define some function $k$
that takes two arguments $x$ and $x'$ and returns a real number expressing their
similarity:

$$
k: \mathcal{X} \times \mathcal{X} \rightarrow \mathbb{R}
$$

As an example, take the dot product.

$$
k(\mathbf{x}, \mathbf{x}') = \langle \mathbf{x}, \mathbf{x}' \rangle = \sum x_ix'_i
$$

\textit{Geometrically}, the dot product expresses the cosine of the angle
between $\mathbf{x}$ and $\mathbf{x}'$. The dot product also expresses the 
\textit{length} of a vector, i.e.
$||\mathbf{x}|| = \sqrt{\langle \mathbf{x}, \mathbf{x} \rangle}$. Lastly, the
\textit{distance} between two vectors is computed as the length of the difference vector.
So, being able to compute dot products amounts to being able to carry out all
mathematical operations that can be characterized in terms of angles, lengths,
and distances (cough cough, linear algebra and analytic geometry).

BUT, we cannot take for granted that our data actually lives in a
dot product space. For example, it might be text strings. So, we need some
transformation that takes possibly non-vectorial data $x$ and maps it to the dot
product space $\mathcal{H}$ (which need not be $\mathbb{R}^n$).

$$
\Phi: \mathcal{X} \rightarrow \mathcal{H}
$$
$$
x \mapsto \mathbf{x} := \Phi(x)
$$

We can cleverly pick our kernel function so that we do not actually need to
carry out this mapping.

\bigskip

\textbf{The ``kernel trick''}

Suppose that we have the feature map $\Phi$ that takes two-dimensional inputs and
maps them to a three-dimensional space.
$$
\Phi: \mathbb{R}^2 \rightarrow \mathbb{R}^3
$$
Specifically:
\begin{equation}
\Phi: \left[x_1, x_2 \right] \rightarrow \left[x_1^2, x_2^2, \sqrt{2}x_1x_2
\right]
\end{equation}

Now suppose that we will perform standard soft-margin SVM on the transformed
data. In its dual form:

$$
\max_{\alpha \in \mathbb{R}^n} 
\sum_{i=1}^n \alpha_i - \frac{1}{2} \sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j
Y_i Y_j \langle \Phi(X_i), \Phi(X_j) \rangle
$$

Let's examine that inner product bit on the right in more detail.

$$
\langle \Phi(X_i), \Phi(X_j) \rangle = \langle (X_{i1}^2, X_{i2}^2,
\sqrt{2}X_{i1}X_{j2}),  (X_{j1}^2, X_{j2}^2,
\sqrt{2}X_{i1}X_{j2})\rangle
$$
$$
= X_{i1}^2X_{j1}^2 + 2X_{i1}X_{i_2}X_{j1}X_{j2} + X_{i2}^2X_{j2}^2
$$
$$
= (X_{i1}X_{j1} + X_{j1}X_{j2})^2
$$
$$
= \langle X_i, X_j \rangle^2
$$
$$
:= k(x, x')
$$

(i.e. we are defining our kernel function to be the inner product, squared). Now we rewrite our SVM:

$$
\max_{\alpha \in \mathbb{R}^n} 
\sum_{i=1}^n \alpha_i - \frac{1}{2} \sum_{i=0}^n \sum_{j=0}^n \alpha_i \alpha_j
Y_i Y_j k(X_i, X_j)
$$

Note that we are operating directly on the untransformed $X$, i.e. without
having to conduct the transformation $\Phi$. We have defined $k$ such that $k(x,
x') = \langle \Phi(x), \Phi(x') \rangle$. This is the \textbf{kernel trick}.
It works for any kernel function $k$ that is symmetric and positive
semi-definite. Some common such kernels are:
\begin{itemize}
\item Linear: $k(x, x') = \langle x, x' \rangle$
\item Polynomial: $k(x, x') = \langle x, x' \rangle^\gamma$
\item Gaussian (RBF): $k(x, x') = \exp(-\frac{||x - x'||^2_2}{2\sigma^2})$. This
is the de facto standard for machine learning.
\end{itemize}

\bigskip

\textbf{Why do we care?}

The kernel trick, and kernel methods in general, have some nice value
propositions.

The first is \textbf{applying linear methods to non-linear problems}. Choosing a
kernel that lets us map between dimensions can allow us to apply linear methods
when we otherwise wouldn't be able to. Consider the below classification
problem. Clearly, the best decision boundary is an ellipse, which is non-linear,
so we cannot apply something like SVM. But if we map $\mathbb{R}^2$ to
$\mathbb{R}^3$ using (1), we can train a SVM to separate the classes.

\includegraphics[scale=0.5]{kernel}

Similarly, we can use kernels to apply machine learning methods to non-vectorial
data such as text strings or k-grams or labelled graphs.

The second is \textbf{computational efficiency}. Because of the kernel trick
(and some other stuff covered later), we can explore enormous features spaces
for classification purposes without actually computing such spaces. Suppose we
have $d=100$ and want to consider up to 3rd degree interactions. This already
leads to $d=171,700$---not trivial to compute! Yet through the kernel trick, we
do not actually need to conduct the laborious mapping $\Phi$, we can operate
directly on the input data through the kernel.

\newpage

\textbf{Reproducing kernel Hilbert spaces and the representer theorem}

HOLY SHIT THIS IS ABSTRACT

As previously covered, a Hilbert space basically is one where we have the dot
product.

Reproducing property:

$$
\langle k(x, \bullet), k(x', \bullet) \rangle_{\mathcal{H}_k} := k(x, x')
$$

\end{document}

### 2018-04-08 sgd/sgd.tex
updated prez

@@
 \begin{frame}
-	\frametitle{Optimization is everywhere, and sometimes is easy}
-	Many statistical procedures involve minimizing or maximizing some function 
-	applied to data \\~\\
-
-	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice''
-	\begin{itemize}
-		\item \small Example: in OLS, we do not need to try
-			different values
-			of $\hat{\beta}$ to see which minimizes the loss
-			function, we (typically) can just evaluate
-			$(X'X)^{-1}X'Y$ \\~\\
-	\end{itemize}
-	\smallskip
-
-	\centering
-	\begin{tikzpicture}
-		\begin{axis}[
-			yticklabels={,,},
-			xticklabels={,,},
-			width=0.5\textwidth,
-			height=0.4\textwidth,
-			legend style = {font=\small}
-		]
-			\addlegendentry{$f=\sum (\hat{y} - y)^2$}
-			\addplot[mark=none, color=blue]{x^2};
-		\end{axis}
-	\end{tikzpicture}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Other times, optimization is not so easy}
+	\frametitle{Optimization is fundamental to statistical modeling}
 	Suppose that we have a typical supervised classification problem
@@
 		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
-		\item Optimization: pick weights $\bm{w}$ that minimize
+		\item Optimization: pick weights $\hat{\bm{w}}$ that minimize
 			empirical risk according to some convex
@@
 
-	Our lack of assumptions requires a different approach to optimization
+	Our lack of assumptions takes away some familiar tools for finding $\hat{\mathbf{w}}$
 	\begin{itemize}
-		\item \small Cannot analytically identify stationary point
-		\item Need to numerically search for it
+		\item \small Cannot analytically identify $\hat{\mathbf{w}}$
+			(e.g. in OLS $=(X'X)^{-1}X'Y$))
+		\item \small Newton-Raphson requires 2nd derivative of $L$,
+			which might be a pain (unlike in, for example, GLM)\\~\\
 	\end{itemize}
+
+	But we still know that if $L$ is convex, there is a unique global
+	minimum, and the gradient at that point must be 0
 \end{frame}
@@
 
-	But, big difference between:
-	\begin{itemize}
-		\item Speed $:=$ number of iterations
-		\item Speed $:=$ time (clock on wall)
-	\end{itemize}
 \end{frame}
@@
 
-	This becomes computationally intractable as $n$ grows large \\~\\
-
-	Knowing that the quality of our approximation gets linearly or
-	quadratically better with $t$ is not comforting if each $t$ takes days to run
+	This becomes computationally intractable as $n$ grows large 
+	\begin{itemize}
+		\item \small If we have $n$ = 10 million, we have to evaluate the
+			gradient 10 million times for every step (and the
+			algorithm may take many steps to converge)\\~\\
+		\item \small May take hours or days to converge \\~\\
+	\end{itemize}
 \end{frame}
@@
 
-	For each step, gradient is computed for a \textbf{single} randomly
+	For each step, gradient is computed for a \textcolor{red}{single} randomly
 	chosen observation $i$:
@@
 	$$
-	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w L(f_w(\mathbf{X}_i), y_i)
+	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w
+	L(f_w(\mathbf{X}_{\textcolor{red}{i}}), y_{\textcolor{red}{i})}
 	$$ \\~\\
@@
 \begin{frame}
-	\frametitle{SGD is particularly useful when $n$ is large and computation
+\frametitle{SGD is noisier than batch GD}
+	\centering
+	\includegraphics[scale=0.3]{noise}
+
+	\tiny Source: towardsdatascience.com
+\end{frame}
+
+\begin{frame}
+	\frametitle{SGD is useful when $n$ is large \& compute
 	time is important}
@@
 
-	...SGD probably is involved. Example: Google's \textbf{AlphaGo} program.
+	...SGD probably is involved. \\~\\
 
+	Example: Google's \textbf{AlphaGo} program.
+\end{frame}
+
+\begin{frame}
 	\begin{figure}[b]
 	\centering
-	\includegraphics[scale=0.25]{go}
+	\includegraphics[scale=0.4]{go}
 	\end{figure}
+\end{frame}
 
+\begin{frame}
+	\frametitle{Google slides from ICML 2016}
+	\centering
+	\includegraphics[scale=0.3]{screenshot}
 \end{frame}

### 2018-04-11 paper/paper.tex
work

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

\DeclareMathOperator*{\argmax}{argmax}

\usepackage{biblatex}
\addbibresource{paper.bib}

\author{Tom Wallace}

\title{STAT 672 Final Project: Stochastic Gradient Descent}

\begin{document}

\maketitle

\section{Introduction}

\subsection{Organization}

This paper is divided into four sections. The remainder of this
\textbf{Introduction} section gives motivation for stochastic gradient descent (SGD):
why do we care about it? The \textbf{Method and Theory} section presents the
mathematics of SGD and some of its notable properties. The
\textbf{Applications} highlights the real-world settings and uses of SGD,
including a case study data analysis. The \textbf{Conclusion} section summarizes
overall findings and notes some areas for further research in SGD.

\subsection{Motivation}

Optimization is fundamental to statistical modeling. The chief task of
statistical modeling is to characterize the relationship between 
explanatory variables and an outcome variable, and the chief method for doing so
is to estimate values for coefficients that best relate
each explanatory variable to the outcome variable. The term ``best''
implies picking coefficient values that
maximize some measure of goodness (e.g. likelihood) or minimize some measure of
badness (e.g. loss function). 
Mathematical optimization is the typical
route to achieving such minimization or maximization. Two important
considerations for optimization are parametric assumptions and computational
complexity. SGD, an optimization technique, is
particularly motivated by these considerations.

\subsubsection{Parametric vs. non-parametric}

Assuming that the outcome variable follows a
particular statistical distribution aids the computation of optimal coefficients. 
For example, assumptions in ordinary least squares (OLS)
regression---assumptions that readers almost certainly are familiar with and so will not be
repeated here---allow a closed form solution. Suppose we have $n$
observations. Each observation consists of an outcome variable $y_i$ and some
explanatory variables $x_{i1}, x_{i2}\ldots x_{iD}$. We thus have a vector of
outcomes $\bm{Y}_{n \times 1}$ and a feature matrix $\bm{X}_{n \times D}$. All
variables are real numbers: $\bm{X} \in \mathbb{R}^{n \times D}$, $\bm{Y} \in
\mathbb{R}^{n \times 1}$. The goal
is estimate weights $w_1 \ldots w_j \ldots w_D$, or $\bm{w}_{1 \times D}$, that relate the explanatory variables
to the outcome.\footnote{Readers may be more familiar with $\bm{\beta}$ as the
notation for coefficients. $\bm{w}$ is used here to ensure common notation with
other classes of models.} The normal equations give:

$$
\hat{\bm{w}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}
$$

Even if a parametric model does not have a closed-form solution, the parametric
assumption allows some useful optimization techniques. Consider logistic
regression. The maximum likelihood estimator (MLE) approach leads to
a system of $D$ equations. This system of
equations typically is numerically solved using the iterative Newton-Raphson
algorithm:

$$
\hat{\bm{w}}_{n+1} = \hat{\bm{w}}_{n} -
\bm{H}^{-1}(\hat{\bm{w}}_n)\bm{J}(\hat{\bm{w}}_n)
$$

$\bm{J}$ is the Jacobian (the
first derivative of the log-likelihood function $l$ with respect to each $w_j$)
and $\bm{H}$ is the Hessian (the second derivative of $l$ with respect to $w_j,
w_{j'}$). The practicality of Newton-Raphson thus depends on whether it is convenient to
find $\bm{J}$ and $\bm{H}$. This is true for logistic regression
because parametric and independent-and-identically-distributed (IID) assumptions
mean $l$ is a simple sum of the log probability distribution
function (PDF, in this case binomial) for each observation. We ``know'' (assume)
the form of this PDF and so are
confident that the second derivative exists and is not too onerous to calculate. It may not be true, and hence
Newton-Raphson will not be practical, if the Jacobian or Hessian of the function we are trying to
maximize or minimize (perhaps not $l$) are non-existent or cumbersome.

Non-parametric modeling often is such a case. MAYBE REVISE THIS. Consider a typical supervised
classification set-up: feature matrix $\bm{X} \in \mathbb{R}^{n \times D}$ and
labels $\bm{Y} \in \{-1, 1\}$, with no assumptions regarding distribution. The goal is to pick a function
$f_w(\bm{X})$ from hypothesis class $\mathcal{F}$ that assigns weights to
features so as to predict label. The optimal weights are those that minimize empirical risk
according to some loss function $L$ that characterizes how well (or not)
$f_w(\bm{X})$ classifies in terms of discrepancy between $f_w(\bm{X}_i)$ and
$Y_i$ for a particular $i$.

$$
\hat{\bm{w}} = \argmax_{\bm{w}} \frac{1}{n} \sum_{i=1}^n L(f_w(\bm{X}_i), Y_i)
$$

Our lack of parametric assumptions means we do not have a closed-form solution a
la OLS. We also have no guarantees that the loss function $L$ has existent or easy-to-find derivatives, 
and hence no guarantee that Newton-Raphson is practical.

The need to conduct optimization in such non-parametric settings is a chief
motivation for gradient descent (GD), of which SGD is a variant. GD
does not require any parametric assumptions. It also only requires finding the
first derivative (Jacobian) of the function. For these reasons, (S)GD is
 well-suited for non-parametric settings, included supervised and
unsupervised statistical learning tasks.

\subsubsection{Computational complexity}

How an optimization technique scales with $n$ and $D$ is another important
consideration. It is little comfort if a method reaches the correct solution but
requires an excessive amount of time to do so. ``Plain'' or
``batch'' GD requires evaluating the gradient for every single observation,
every single iteration, until the algorithm converges. For example, for a
dataset of $n=10^6$ that required 25 iterations to converge, batch GD would require 
evaluating the gradient $25 \times 10^6$ times. This linear scaling with
$n$ can cause untenably long computation time.

SGD alleviates these computational difficulties by requiring the gradient to be
evaluated for only a single randomly chosen observation per iteration. This
approach means convergence is ``noisier'' and hence requires more iterations to
converge, but each iteration is less complex to compute and so can be done
faster. SGD thus scales much more favorably with $n$ than GD, and so is
particularly useful for large-$n$ applications such as typical machine learning
or big data tasks.

\section{Method and Theory}

\subsection{Setup and Notation}

\subsection{Derivation}

\cite{bottou2010large}
\cite{bottou2012stochastic}
\cite{boyd2004convex}
\cite{zinkevich2010parallelized}
\cite{dal2015calibrating}
\subsection{Key Properties}

\section{Applications}

\subsection{SGD and Statistical Learning}

\subsection{Case Study: }

\section{Conclusion}

\printbibliography

\end{document}

### 2018-04-11 paper/paper.tex
worked on paper

@@
 
-\usepackage{biblatex}
+\usepackage[citestyle=authoryear]{biblatex}
 \addbibresource{paper.bib}
@@
 This paper is divided into four sections. The remainder of this
-\textbf{Introduction} section gives motivation for stochastic gradient descent (SGD):
-why do we care about it? The \textbf{Method and Theory} section presents the
+\textbf{Introduction} section gives intuitive motivation for stochastic gradient
+descent (SGD). The \textbf{Method and Theory} section more rigorously presents the
 mathematics of SGD and some of its notable properties. The
-\textbf{Applications} highlights the real-world settings and uses of SGD,
+\textbf{Applications} sections highlights the real-world settings and uses of SGD,
 including a case study data analysis. The \textbf{Conclusion} section summarizes
-overall findings and notes some areas for further research in SGD.
+overall findings.
 
@@
 w_{j'}$). The practicality of Newton-Raphson thus depends on whether it is convenient to
-find $\bm{J}$ and $\bm{H}$. This is true for logistic regression
+find $\bm{J}$ and $\bm{H}$. It is convenient for logistic regression
 because parametric and independent-and-identically-distributed (IID) assumptions
@@
 
-Non-parametric modeling often is such a case. MAYBE REVISE THIS. Consider a typical supervised
-classification set-up: feature matrix $\bm{X} \in \mathbb{R}^{n \times D}$ and
-labels $\bm{Y} \in \{-1, 1\}$, with no assumptions regarding distribution. The goal is to pick a function
-$f_w(\bm{X})$ from hypothesis class $\mathcal{F}$ that assigns weights to
-features so as to predict label. The optimal weights are those that minimize empirical risk
-according to some loss function $L$ that characterizes how well (or not)
-$f_w(\bm{X})$ classifies in terms of discrepancy between $f_w(\bm{X}_i)$ and
-$Y_i$ for a particular $i$.
-
-$$
-\hat{\bm{w}} = \argmax_{\bm{w}} \frac{1}{n} \sum_{i=1}^n L(f_w(\bm{X}_i), Y_i)
-$$
-
-Our lack of parametric assumptions means we do not have a closed-form solution a
-la OLS. We also have no guarantees that the loss function $L$ has existent or easy-to-find derivatives, 
-and hence no guarantee that Newton-Raphson is practical.
-
-The need to conduct optimization in such non-parametric settings is a chief
-motivation for gradient descent (GD), of which SGD is a variant. GD
-does not require any parametric assumptions. It also only requires finding the
-first derivative (Jacobian) of the function. For these reasons, (S)GD is
- well-suited for non-parametric settings, included supervised and
-unsupervised statistical learning tasks.
+The need to conduct optimization in non-parametric settings is a chief
+motivation for gradient descent (GD), of which SGD is a variant. In
+non-parametric settings---most notably supervised and unsupervised statistical
+learning, in which we again seek to find optimal $\hat{\bm{w}}$ to relate input
+variables to output variables for the purposes of classification or regression---there 
+typically is no closed form solution for $\hat{\bm{w}}$. It
+also may not be convenient to find and evaluate the Hessian, making
+Newton-Raphson undesirable. SGD does not require any parametric assumptions. 
+In its most basic form, SGD only requires finding the gradient (though some extensions
+do need the Hessian or an approximation to it). 
+SGD thus is well-suited for non-parametric settings.
 
@@
 dataset of $n=10^6$ that required 25 iterations to converge, batch GD would require 
-evaluating the gradient $25 \times 10^6$ times. This linear scaling with
+evaluating the gradient $25 \times 10^6$ times. This scaling with
 $n$ can cause untenably long computation time.
@@
 faster. SGD thus scales much more favorably with $n$ than GD, and so is
-particularly useful for large-$n$ applications such as typical machine learning
-or big data tasks.
+particularly useful for large-$n$ applications such as machine learning
+and big data problems.
 
@@
 
-\subsection{Derivation}
+Consider a typical soft-margin support vector machine (SVM) problem.
+
+\subsection{Basic Form}
 
@@
 \cite{boyd2004convex}
-\cite{zinkevich2010parallelized}
 \cite{dal2015calibrating}
+
 \subsection{Key Properties}
+
+\subsection{Extensions}
+
+The basic SGD algorithm has been extended in different ways. The popularity of
+the algorithm disallows a comprehensive or detailed treatment of all
+development. This sub-section covers some of the more interesting developments.
+
+\subsubsection{Step Size (Learning Rate)}
+
+\cite{shalev2011pegasos}
+
+\subsubsection{Momentum}
+
+\subsubsection{Averaging}
+
+\subsubsection{Predictive Variance Reduction}
+
+\subsubsection{Parallelization}
+
+SGD is commonly used in large-$n$, computationally demanding applications. Thus,
+even though SGD is a computational improvement over batch GD, there has been
+interest in whether SGD can be made even faster by parallelizing it.
+\cite{zinkevich2010parallelized} present novel algorithms for doing so. The
+actual algorithms are strikingly simple; their proof is highly technical and
+omitted here.
+
+Consider the SVM problem posed in Section 2.
 
@@
 \subsection{SGD and Statistical Learning}
-
-\subsection{Case Study: }
 

### 2018-04-12 hw5/hw5_1.py
initial

@@ -1,0 +1,33 @@
+import numpy as np
+from sklearn.linear_model import LogisticRegression
+
+# Part 1
+def RunSimulation(n, D):
+	
+	# n x D feature matrix
+	# X_ij = N(0,1)
+	X = np.random.randn(n, D)
+	
+	# Y = {1, -1, 1, -1...}
+	Y = np.ones(n)
+	Y[1::2] = Y[1::2] * -1
+	Y = np.reshape(Y, (n, 1))
+
+	# c_j = \frac{1}{n} \sum_{i=1}^n X_{ij}Y_i
+	C = np.sum(X * Y, 0) / n
+
+	# find indices with largest magnitude
+	C = abs(C)
+	ind = np.argpartition(C, -5)[-5:]
+
+	# new feature matrix X_s
+	X_s = X[:,ind]
+
+	# logistic regression
+	Y = np.reshape(Y, (n,))
+	L = LogisticRegression()
+	L.fit(X_s, Y)
+	print(L.coef_)
+	print(L.intercept_)
+
+RunSimulation(100, 10000)

### 2018-04-12 hw5/hw5_1.py
finished code for part 1

@@ -1,7 +1,12 @@
+# Tom Wallace
+# STAT 672
+# Spring 2018
+# Homework 5
+# Part I
+
 import numpy as np
 from sklearn.linear_model import LogisticRegression
 
-# Part 1
 def RunSimulation(n, D):
 	
 	# n x D feature matrix
@@ -23,11 +28,44 @@ def RunSimulation(n, D):
 	# new feature matrix X_s
 	X_s = X[:,ind]
 
-	# logistic regression
-	Y = np.reshape(Y, (n,))
-	L = LogisticRegression()
-	L.fit(X_s, Y)
-	print(L.coef_)
-	print(L.intercept_)
+	# join X_s and Y into one matrix
+	mat = np.hstack((X_s, Y))
+
+	# 10-fold cross validation
+	# break data into 10 folds
+	folds = np.split(mat, 10)
+
+	lst1 = [1,2,3,4,5,6,7,8,9,10]
+	lst2 = []
+	# for every fold...
+	for fold in folds:
+
+		# this fold is for testing
+		test_X = fold[:, 0:5]
+		test_Y = fold[:, 5]
+
+		# skl logistic regression is fussy about Y's shape
+		test_Y = np.reshape(test_Y, (int(n/10),))
+
+		# everything but this fold is for training
+		train = [x for x in folds if not np.array_equal(x, fold)]
+		train = np.concatenate(train)
+		train_X = train[:, 0:5]
+		train_Y = np.reshape(train[:, 5], (int(9 * (n/10)),))
+
+		# train logistic regression 
+		L = LogisticRegression()
+		L.fit(train_X, train_Y)
+
+		# validate
+		err = 1-L.score(test_X, test_Y)
+
+		lst2.append(err)
+	
+	print("fold,err")
+	for i in range(0,10):
+		print(lst1[i], lst2[i], sep = ",")
+	
+	print("Average err:", sum(lst2)/10)
 
 RunSimulation(100, 10000)

### 2018-04-13 hw5/hw5_1.py
rover 2.0

@@ -35,8 +35,7 @@ def RunSimulation(n, D):
 	# break data into 10 folds
 	folds = np.split(mat, 10)
 
-	lst1 = [1,2,3,4,5,6,7,8,9,10]
-	lst2 = []
+	lst = []
 	# for every fold...
 	for fold in folds:
 
@@ -60,12 +59,8 @@ def RunSimulation(n, D):
 		# validate
 		err = 1-L.score(test_X, test_Y)
 
-		lst2.append(err)
+		lst.append(err)
 	
-	print("fold,err")
-	for i in range(0,10):
-		print(lst1[i], lst2[i], sep = ",")
-	
-	print("Average err:", sum(lst2)/10)
+	print("10-fold CV error:", sum(lst)/10)
 
-RunSimulation(100, 10000)
+RunSimulation(10000, 100)

### 2018-04-13 hw5/hw5.tex
rover 2.0

\documentclass{article}

\usepackage{bm}

\usepackage{geometry}
\geometry{margin=1.2in}

\title{STAT 672 HW 5}

\author{Tom Wallace}

\begin{document}

\maketitle

\section*{I-D}

10-fold cross validation error is 0.19

This error rate is lower than we would expect based on the data-generating
process described in \textbf{1A}. 
There is no true relationship between $\bm{X}_i$ and $Y_i$ and so in theory the logistic
regression model should be no more effective than random guessing. We thus would
expect a cross-validation error of about 0.5. The better-than-expected performance 
is attributable to two factors.

One, $D >> n$. Simply by random chance some features will be highly correlated with the
label. Our regularization process selected these features to be part of the
model. If $D$ is not so large relative to $n$, model performance drops to the
expected level. For example, I ran the simulation with $n=10000$ and $D=100$,
and obtained a cross-validation error of 0.49.

Two, we violated the central premise of cross-validation---a strict segregation
between training and validation sets---by conducting some model training 
on the full data-set. The regularization process in \textbf{1B} selected a
subset of features $\bm{X_S}$ with a particularly strong relationship with label
$\bm{Y}$; we only conducted cross-validation \textit{after} this feature
selection was performed on the \textit{whole} dataset. In this sense we
``cheated'' and allowed the training process to incorporate information
from the validation set. A correct implementation of cross-validation would
conduct the regularization process of \textbf{1B} only on the training set, and
doing so would cause the better-than-expected performance to disappear.

\section*{II-A}

The best models and the associated cross-validation error is given below. The
numbers in the left-hand column are the indices of features included in the
model.

\begin{table}[h!]
	\centering
	\begin{tabular}{|l r|}
		\hline
		\textbf{Features in Model} & \textbf{5-Fold CV Error} \\
		0  3  4  5  6  9 11 14 & 0.225 \\
		0  3  4  5  6  9 11 15 & 0.225 \\
		1  3  4  5  6  8  9 13 14 & 0.225 \\
		1  3  4  5  6  8  9 13 & 0.225 \\
		1  3  8  9 10 11 13 14 & 0.225 \\
		1  3  8  9 10 11 13 & 0.225 \\
		2  3  9 10 14 & 0.225 \\
		2  3  9 10 & 0.225 \\
		3  4  5  6  9 13 14 & 0.225 \\
		3  4  5  6  9 13 & 0.225 \\
		3  6  7  8  9 10 14 & 0.225 \\
		3  6  7  8  9 10 & 0.225 \\
		\hline
	\end{tabular}
\end{table}

\section*{II-B}

The final classifier\footnote{As shown in the preceding table, there are
multiple models that achieve the lowest cross-validation error, so really this
sentence should be final classifier\textit{s}. I tried several, and all led to
the same phenomenon.} trained on the whole training data attains an error rate of
0.502 against the testing data (\texttt{Xtest, Ytest}). This is much higher
than the cross-validation error of 0.225 obtained in \textbf{II-B}, and in fact
is about what one would get from random guessing. There are several reasons for
this underperformance.

One, $n_{\mathrm{train}} << n_{\mathrm{test}}$. Because the model is trained on
such a small dataset, it did not learn the general relationship between features
and label. Note that the mathematical proof of why validation works (slides 28-29 
of class 6) implies that the probability of producing the optimal model
increases with the size of the validation set. It thus is
unsurprising that our small validation set failed to produce a good model.

Two, our classifier lacks a regularization parameter, and so it overfits to the
training data. This results in achieving low empirical error (i.e. good
performance on the training set) but high generalization error (i.e. poor
performance on the training set).

\section*{II-C}

Randomly permuting the labels in \texttt{Ytrain} could expose the aforementioned
issues without the need for a test set. Randomly permuting the labels would mean
there is no true relationship between feature and labels, and so a model attempting
to predict label based on features should be no better than random guessing
(i.e. should achieve cross-validation error of about 0.5). 
If the model achieves much lower error than this,
there is something wrong with the feature selection and/or classifier.

\section*{II-D}

\end{document}

### 2018-04-13 hw5/hw5_2.py

@@ -1,0 +1,122 @@
+import numpy as np
+import itertools as it
+from sklearn.linear_model import LinearRegression
+
+def powerset(iterable):
+	s = list(iterable)
+	return it.chain.from_iterable(it.combinations(s, r) for r in range(len(s)+1))
+
+def RunSimulation():
+	
+	# read data
+	X_train = np.genfromtxt("Xtrain.csv", delimiter = ",")
+	X_test = np.genfromtxt("Xtest.csv", delimiter = ",")
+	Y_train = np.genfromtxt("Ytrain.csv", delimiter = ",")
+	Y_test = np.genfromtxt("Ytest.csv", delimiter = ",")
+
+	# get D and n_train
+	n_train = np.shape(X_train)[0]
+	D = np.shape(X_train)[1]
+
+	# combine training data into single matrix
+	Y_train = np.reshape(Y_train, (n_train, 1))
+	mat = np.hstack((X_train, Y_train))
+
+	# list of all possible feature combinations
+	fl = list(powerset(np.arange(0, 15, 1)))
+
+	# dont need empty set
+	fl.pop(0)
+
+	# skl doesn't have a linear least squares classifier
+	# but it does have a ridge classifier
+	# and ridge with regularization parameter = 0 is equivalent to linear least squares
+	L = LinearRegression()
+
+	# try every feature combination
+	for f in fl:
+
+		# get various n
+		n_fold_train = int(4*(n_train / 5))
+		n_fold_test = int(n_train / 5)
+
+		# get number of features
+		D_f = np.shape(f)[0]
+
+		# indices of selected features, and y
+		f = np.append(f, 15)
+
+		# sub-select those indices
+		mat_f = mat[:, f]
+
+		# break data into 5 folds
+		folds = np.split(mat_f, 5)
+
+		# empty list to which we will append CV errors for this model
+		err_lst = []
+
+		# 5-fold cross validation
+		for fold in folds:
+			
+			# this fold is for testing
+			test_X = fold[:, 0:D_f]
+			test_Y = fold[:, D_f]
+			
+			# skl is fussy about Y's shape
+			test_Y = np.reshape(test_Y, (n_fold_test, ))
+			
+			# everything but this fold is for training
+			train = [x for x in folds if not np.array_equal(x, fold)]
+			train = np.concatenate(train)
[…]

### 2018-04-13 hw5/hw5_2.py
final

@@ -118,5 +118,98 @@ def TestModel():
 
 	print(err)
 
+def RunPermutationSimulation():
+
+	# read data
+	X_train = np.genfromtxt("Xtrain.csv", delimiter = ",")
+	X_test = np.genfromtxt("Xtest.csv", delimiter = ",")
+	Y_train = np.genfromtxt("Ytrain.csv", delimiter = ",")
+	Y_test = np.genfromtxt("Ytest.csv", delimiter = ",")
+
+	# get D and n_train
+	n_train = np.shape(X_train)[0]
+	D = np.shape(X_train)[1]
+
+	# combine training data into single matrix
+	Y_train = np.reshape(Y_train, (n_train, 1))
+	mat = np.hstack((X_train, Y_train))
+
+	# list of all possible feature combinations
+	fl = list(powerset(np.arange(0, 15, 1)))
+
+	# dont need empty set
+	fl.pop(0)
+
+	# skl doesn't have a linear least squares classifier
+	# but it does have a ridge classifier
+	# and ridge with regularization parameter = 0 is equivalent to linear least squares
+	L = LinearRegression()
+
+	# 20 random permutations of data
+	for i in range(0, 20):
+
+		# list where we'll append cv error from each fit
+		cv_fits = []
+
+		# try every fit
+		for f in fl:
+
+			# get various n
+			n_fold_train = int(4*(n_train / 5))
+			n_fold_test = int(n_train / 5)
+
+			# get number of features
+			D_f = np.shape(f)[0]
+
+			# indices of selected features, and y
+			f = np.append(f, 15)
+
+			# sub-select those indices
+			mat_f = mat[:, f]
+
+			# break data into 5 folds
+			folds = np.split(mat_f, 5)
+
+			# empty list to which we will append CV errors for this model
+			err_lst = []
+
+			# 5-fold cross validation
+			for fold in folds:
+				
+				# this fold is for testing
+				test_X = fold[:, 0:D_f]
+				test_Y = fold[:, D_f]
+				
+				# skl is fussy about Y's shape
+				test_Y = np.reshape(test_Y, (n_fold_test, ))
+				
+				# everything but this fold is for training
+				train = [x for x in folds if not np.array_equal(x, fold)]
+				train = np.concatenate(train)
+				train_X = train[:, 0:D_f]
+				train_Y = np.reshape(train[:, D_f], (n_fold_train, ))
+
+				# fit model
+				L.fit(train_X, train_Y)
+
[…]

### 2018-04-15 paper/paper.tex
initial
rover 2.0

@@
 \DeclareMathOperator*{\argmax}{argmax}
+\DeclareMathOperator*{\argmin}{argmin}
 
@@
 regression---assumptions that readers almost certainly are familiar with and so will not be
-repeated here---allow a closed form solution. Suppose we have $n$
-observations. Each observation consists of an outcome variable $y_i$ and some
-explanatory variables $x_{i1}, x_{i2}\ldots x_{iD}$. We thus have a vector of
-outcomes $\bm{Y}_{n \times 1}$ and a feature matrix $\bm{X}_{n \times D}$. All
-variables are real numbers: $\bm{X} \in \mathbb{R}^{n \times D}$, $\bm{Y} \in
-\mathbb{R}^{n \times 1}$. The goal
-is estimate weights $w_1 \ldots w_j \ldots w_D$, or $\bm{w}_{1 \times D}$, that relate the explanatory variables
-to the outcome.\footnote{Readers may be more familiar with $\bm{\beta}$ as the
-notation for coefficients. $\bm{w}$ is used here to ensure common notation with
-other classes of models.} The normal equations give:
-
-$$
-\hat{\bm{w}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}
-$$
+repeated here---allow a closed form solution. The optimization problem of
+choosing coefficient $\hat{\bm{\beta}}$ that minimize squared error is solved
+by $\hat{\bm{\beta}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ (where $\bm{X}$ is the
+feature matrix and $\bm{Y}$ the outcome variable).
 
@@
 assumption allows some useful optimization techniques. Consider logistic
-regression. The maximum likelihood estimator (MLE) approach leads to
-a system of $D$ equations. This system of
+regression. The maximum likelihood estimator (MLE) approach for estimating
+coefficients leads to a system of $D$ equations. This system of
 equations typically is numerically solved using the iterative Newton-Raphson
@@
 $$
-\hat{\bm{w}}_{n+1} = \hat{\bm{w}}_{n} -
-\bm{H}^{-1}(\hat{\bm{w}}_n)\bm{J}(\hat{\bm{w}}_n)
+\hat{\bm{\beta}}_{n+1} = \hat{\bm{\beta}}_{n} -
+\bm{H}^{-1}(\hat{\bm{\beta}}_n)\bm{J}(\hat{\bm{\beta}}_n)
 $$
@@
 the form of this PDF and so are
-confident that the second derivative exists and is not too onerous to calculate. It may not be true, and hence
-Newton-Raphson will not be practical, if the Jacobian or Hessian of the function we are trying to
-maximize or minimize (perhaps not $l$) are non-existent or cumbersome.
+confident that the second derivative exists and is not too onerous to calculate.
+In non-parametric settings, we often cannot be so certain and face the possibility of
+$\bm{H}$ being non-existent or cumbersome.
 
@@
 non-parametric settings---most notably supervised and unsupervised statistical
-learning, in which we again seek to find optimal $\hat{\bm{w}}$ to relate input
+learning, in which we again seek to find optimal coefficients to relate input
 variables to output variables for the purposes of classification or regression---there 
-typically is no closed form solution for $\hat{\bm{w}}$. It
+typically is no closed form solution for the coefficients. It
 also may not be convenient to find and evaluate the Hessian, making
@@
 
-\subsubsection{Computational complexity}
+\subsubsection{Computational Complexity}
 
-How an optimization technique scales with $n$ and $D$ is another important
+How an optimization technique scales with sample size $n$ is another important
 consideration. It is little comfort if a method reaches the correct solution but
@@
 
-\subsection{Setup and Notation}
+\subsection{Basic setup}
 
-Consider a typical soft-margin support vector machine (SVM) problem.
+Loosely following the set-up of \cite{bottou2010large}, consider a typical supervised classification problem. We have
+feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
+features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
+predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
+We have a hypothesis class $\mathcal{F}$ consisting of
+various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
+vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
+expresses the cost of mis-classification. Assume for now that $L$ is convex. 
+We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
+that which minimizes empirical risk over all observations: $\frac{1}{n}
+\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
+Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
+have:
 
-\subsection{Basic Form}
+\begin{equation}
+	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))
+\end{equation}
 
-\cite{bottou2010large}
-\cite{bottou2012stochastic}
-\cite{boyd2004convex}
-\cite{dal2015calibrating}
+Because $L$ is convex, basic calculus tells us that there is one critical point,
+it is the global minimum, and it is located where the gradient of the loss
+function with respect to $\bm{w}$, $\nabla_{\bm{w}}$, is zero. GD is an iterative
+algorithm to numerically approximate this point:
+
+\begin{equation}
+	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
+	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+\end{equation}
+
+$t$ refers to iterations. $\gamma$ is a parameter controlling the step
+size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. 
+The algorithm stops when the gradient is
+arbitarily close to zero; or, put differently, when the difference between this
+iteration's estimate and the last iteration's estimate is arbitrarily small.
+
+\begin{equation}
+	\bm{w}_t - \bm{w}_{t-1} \leq \epsilon
+\end{equation}
+
+Note that the batch GD algorithm presented in (2) requires evaluating the
+gradient for every single observation $i$. In a large-$n$ dataset, this can be
+computationally infeasible. SGD's innovation is to instead evaluate only a
+single randomly chosen observation $i$ at each iteration $t$, greatly lightening
+the computational load.
+
+\begin{equation}
+	\bm{w}_{t+1} = \bm{w}_t - \gamma 
+	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+\end{equation}
+
+Having outlined the basic form of SGD, we now examine some of its key
+properties.
 
 \subsection{Key Properties}
+
+\subsubsection{Correctness}
+
+\subsubsection{Speed}
 
@@
 the algorithm disallows a comprehensive or detailed treatment of all
-development. This sub-section covers some of the more interesting developments.
+development. This sub-section covers some of the more interesting extensions.
 
+\cite{bottou2012stochastic}
+\cite{boyd2004convex}
+\cite{dal2015calibrating}
 \subsubsection{Step Size (Learning Rate)}
@@
 \subsubsection{Momentum}
+
+\cite{polyak1992acceleration}
+\cite{nesterov}
 
@@
 
-Consider the SVM problem posed in Section 2.
+The parallelization technique essentially is averaging. In line with previous
+notation, suppose we have fixed learning rate $\gamma$,
+
+Advantages
 

### 2018-04-15 paper/paper.tex
progress

@@
 
-\subsection{Basic setup}
+\subsection{Basic Form}
 
-Loosely following the set-up of \cite{bottou2010large}, consider a typical supervised classification problem. We have
-feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
-features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
-predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
-We have a hypothesis class $\mathcal{F}$ consisting of
-various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
-vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
-expresses the cost of mis-classification. Assume for now that $L$ is convex. 
-We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
-that which minimizes empirical risk over all observations: $\frac{1}{n}
-\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
-Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
-have:
+This section presents the basic form of GD and SGD to set up a more detailed
+examination in subsequent sections. Suppose we want to minimize a function $f: \mathbb{R}^D \to \mathbb{R}$.
 
 \begin{equation}
-	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))
+	\min_w f(w)
 \end{equation}
 
-Because $L$ is convex, basic calculus tells us that there is one critical point,
-it is the global minimum, and it is located where the gradient of the loss
-function with respect to $\bm{w}$, $\nabla_{\bm{w}}$, is zero. GD is an iterative
-algorithm to numerically approximate this point:
+$f$ takes as input $w \in \mathbb{R}^D$. Assume
+that $f$ is convex and differentiable, i.e. we can compute its gradient with
+respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (1),
+is denoted $\hat{w}$. The iterative GD algorithm for finding $\hat{w}$ is:
 
 \begin{equation}
-	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
-	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+	w^{(t+1)} = w^{(t)} - \gamma \nabla f(w^{(t)})
 \end{equation}
 
-$t$ refers to iterations. $\gamma$ is a parameter controlling the step
-size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. 
-The algorithm stops when the gradient is
-arbitarily close to zero; or, put differently, when the difference between this
-iteration's estimate and the last iteration's estimate is arbitrarily small.
+$t$ refers to a particular iteration. Assume that we have supplied an initial
+starting guess for $w$ for $t=0$, $w^{(0)}$. $\gamma$ refers to step size (also called
+learning rate). Assume for now that $\gamma$ is fixed. The GD algorithm iterates
+until some stop condition is met. This may be a fixed number of iterations, or
+that the quality of approximation meets some predefined threshold. 
+A common stopping condition is when the L2 norm of the gradient
+is less than some arbitrarily small constant.
 
 \begin{equation}
-	\bm{w}_t - \bm{w}_{t-1} \leq \epsilon
+	||\nabla f(w^{(t)})||_2 \leq \epsilon
 \end{equation}
 
-Note that the batch GD algorithm presented in (2) requires evaluating the
-gradient for every single observation $i$. In a large-$n$ dataset, this can be
-computationally infeasible. SGD's innovation is to instead evaluate only a
-single randomly chosen observation $i$ at each iteration $t$, greatly lightening
-the computational load.
+Consider a modified situation. Suppose $f$ now takes two arguments, $w$ and $X
+\in \mathbb{R}^D$. $f : \mathbb{R}^D \times \mathbb{R}^D \rightarrow
+\mathbb{R}$. We have $n$ observations of $X$, and denote by
+$\bm{X}_{n \times D}$ the matrix of these observations, with $X_i$ being a
+particular observation or row in that matrix. We apply
+$f$ over all observations, i.e. $\frac{1}{n} \sum_{i=1}^n f(w, X_i)$. Our new
+problem is:
 
 \begin{equation}
-	\bm{w}_{t+1} = \bm{w}_t - \gamma 
-	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+	\min_w  \frac{1}{n} \sum_{i=1}^n f(w, X_i)
 \end{equation}
 
-Having outlined the basic form of SGD, we now examine some of its key
-properties.
+We could apply the GD algorithm above to find the optimal $w$.
+
+\begin{equation}
+	w^{(t+1)} = w^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n \nabla f(w^{(t)},
+	X_i)
+\end{equation}
+
+But, note that doing so requires evaluating the gradient at every single
+observation $i \leq n$. This may be computationally intractable for large $n$
+and high-dimensional $D$. The innovation of SGD is to instead evaluate only a
+single randomly-chosen $i$ at each iteration $t$.
+
+\begin{equation}
+	w^{(t+1)} = w^{(t)} - \gamma \nabla f(w^{(t)}, X_i)
+\end{equation}
+
+As it turns out, SGD converges to $\hat{w}$ and does so in a computationally
+advantageous way compared to GD.
 
@@
 
+This section gives both intuition and (partial) proof for why (S)GD converges to
+$\hat{w}$.
+
+Intuitively, our assumption that $f$ is convex means that there is one critical
+point and it is the global minimum. Basic calculus tells us that this critical
+point is located where $f'=0$ (in one dimension) or $\nabla f = 0$ (in higher
+dimensions). So our task is to search in the space of $f$ for that point. We
+start with some initial guess $w^{(0)}$, and every iteration, move ``down'' the
+gradient in search of zero. Every iteration, we check if the gradient is equal
+to 0; if not, we keep moving ``down.'' If the gradient is arbitrarily close to
+zero, then we have found the critical point, which must be the value of $w$ that
+minimizes $f$ and hence is $\hat{w}$ (or at least a close approximation to it).
+
+For a more formal proof---following that of \cite{tibs_notes}---assume (in addition to previously stated assumptions
+that $f$ is convex and differentiable) that the gradient of $f$ is Lipschitz-continuous
+with constant $L$, i.e. that $||\nabla f(x) - \nabla f(y)||_2 \leq L||x-y||_2$
+for any $x, y$. This implies that $\nabla^2 f(w) - LI$ is a negative
+semi-definite matrix. We perform a quadratic expansion of $f$ around $f(x)$ and
+obtain the following inequality:
+$$
+f(y) \leq f(x) + \nabla f(x)^T(y-x) + \frac{1}{2}\nabla^2 f(x)||y-x||_2^2
+$$
+\begin{equation}
+	\leq f(x) + \nabla f(x)^T(y-x) + \frac{1}{2}L||y-x||_2^2
+\end{equation}
+
+Now, suppose we use the GD algorithm presented in (2) with $\gamma \leq
+\frac{1}{L}$. Denote $x^+ = x - \gamma \nabla
+f(x)$ and substitute $x^+$ in for $y$:
+
+$$ f(x^+) \leq f(x) + \nabla f(x)^T(x^+ - x) + \frac{1}{2}L||x^+ - x||_2^2 $$
+$$ = f(x) + \nabla f(x)^T(x - \gamma \nabla f(x) - x) + \frac{1}{2}L||x - \gamma \nabla f(x) - x||_2^2 $$
+$$ = f(x) - \nabla f(x)^T \gamma \nabla f(x)  + \frac{1}{2}L||\gamma \nabla f(x)||_2^2 $$
+$$ = f(x) - \gamma||\nabla f(x)||_2^2 + \frac{1}{2} L \gamma^2||\nabla f(x)||_2^2$$
+\begin{equation}
+= f(x) - (1 - \frac{1}{2}L\gamma)\gamma||\nabla f(x)||_2^2 
+\end{equation}
+
+We have defined $\gamma \leq \frac{1}{L}$. This implies:
+$$
+-(1 - \frac{1}{2}L \gamma) = \frac{1}{2}L \gamma - 1
+\leq \frac{1}{2}L\frac{1}{L} - 1
+= \frac{1}{2} - 1
+= - \frac{1}{2}
+$$
+
+Returning to (8), we obtain:
+
+\begin{equation}
+f(x^+) \leq f(x) - \frac{1}{2}\gamma||\nabla f(x)||_2^2
+\end{equation}
+
+By definition, a squared L2 norm will always be positive unless its content is
+0, and we have defined $\gamma$ to be positive, so $\frac{1}{2} \gamma||\nabla f(x)||_2^2$ will always be positive unless the
+gradient is equal to zero. So, (9) implies that GD results in a strictly
+decreasing objective function value until it reaches the point where the
+gradient equals zero, which is the optimal value. A key caveat is an
+appropriately chosen $\gamma$, a point covered in more detail later.
+
+The above proof is for GD. A rigorous proof is not given for why SGD also
+converges to the optimal value. Informally, we can note that the above proof
+says given infinite $t$ and appropriate $\gamma$, the iterative algorithm will
+always converge to the optimal value. SGD is doing the same thing as GD, 
+just with a single observation per iteration 
+rather than the entire dataset per iteration. Since SGD is using less information
+per iteration, we would expect the convergence to require more iterations. But
+per (9) it is guaranteed to \textit{eventually} arrive at the optimal solution.
+Put differently, the difference between GD and SGD must then primarily revolve
+around convergence speed, not the final value that is converged to.
+
 \subsubsection{Speed}
+
+Table 1---a simplified version of that in \cite{bottou2010large}---illustrates
+the computational advantages of SGD over GD. As shown in row 1, because SGD uses
+less information per iteration than GD, it requires more iterations to achieve
+some fixed degree of accuracy $\rho$.
+However, row 2 shows that because GD computes the gradient for every observation every iteration,
+while SGD only does so for a single randomly chosen iteration, GD's time per
+iteration scales linearly with $n$ while SGD's is a constant. Thus, we conclude
+that SGD's time to
+reach accuracy $\rho$ scales only with $\rho$, while GD's scales with both
+$\rho$ and---crucially---linearly with $n$. So, the larger $n$ grows, the larger
+SGD's computational advantage over GD.
+
+\begin{table}[h!]
+	\centering
+	\caption{Asymptotic comparison of GD and SGD}
+	\begin{tabular}{|l l l|}
+		\hline
+		& \textbf{GD} & \textbf{SGD} \\
+		Iterations to accuracy $\rho$ & $\log \frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
+		Time per iteration & $n$ & 1 \\
+		Time to accuracy $\rho$ & $n \log \frac{1}{p}$ & $\frac{1}{\rho}$ \\
+		\hline
+	\end{tabular}
+\end{table}
+
+We can visually depict SGD's ``noisier'' convergence in comparison to GD.
 
@@
 
-The basic SGD algorithm has been extended in different ways. The popularity of
+The basic SGD algorithm has been extended in many different ways. The popularity of
 the algorithm disallows a comprehensive or detailed treatment of all
-development. This sub-section covers some of the more interesting extensions.
+development. This sub-section covers some of the more important and interesting extensions.
 
-\cite{bottou2012stochastic}
-\cite{boyd2004convex}
-\cite{dal2015calibrating}
-\subsubsection{Step Size (Learning Rate)}
+\subsubsection{Step Size}
 
-\cite{shalev2011pegasos}
+As hinted at in our proof of convergence to $\hat{w}$, much depends on
+an appropriately chosen step size $\gamma$. If $\gamma$ is too small, (S)GD may
+take an excessively long time to run (since we are only moving a small distance
+``down'' the gradient every iteration); if $\gamma$ is too large, then an
+iteration's step may ``overshoot'' the critical point and prevent convergence.
+An entire literature has developed around the best way to select $\gamma$. The
+central insight that is common to most $\gamma$-selection methods is to treat it
+as a time-indexed rather than fixed hyperparameter, i.e., $\gamma^{(t)}$ rather
+than $\gamma$. Ideally we would like $\gamma^{(t)}$ to be large when far away
+from the critical point (so as to increase speed of convergence) and to be small
+when close to the critical point (so as to avoid overshooting).
 
-\subsubsection{Momentum}
+Line search is one method for computing step size. As described in
+\cite{boyd2004convex}, there are two main variants: \textit{exact} and
+\textit{backtracking}. In exact line search, $\gamma$ is chosen to minimize $f$
+along the ray $\{x + \gamma \Delta x\}$, where $\gamma \in \mathbb{R}^+)$ and
+$\Delta x$ is the descent direction determined by SGD:
+\begin{equation}
+	\gamma = \argmin_{s \geq 0} f(x + s\Delta x)
+\end{equation}
+This obviously is an optimization problem in and of itself, and
+so it can be computationally impractical to add this burden in addition to the
+``main'' optimization problem we are trying to solve using SGD. For this reason,
+\textit{backtracking} is an iterative approximation method that is
+computationally lighter. The backtracking algorithm takes two hyper-parameters
+$\alpha \in (0, 0.5)$ and $\beta \in (0, 1)$. It starts with an initial guess of
+$s^{(0)} = 1$, and then for every iteration $t$, updates $s := \beta t$. The
+algorithm stops when $f(x + s \Delta x) \leq f(x) + \alpha s \Delta f(x)^T
+\Delta x$, and the final value of $s$ is taken as $\gamma$.
 
+\cite{bottou2012stochastic} advocates using learning rates of the form
+$\gamma^{(t)} = \gamma^{(0)}(1 + \gamma^{(0)}\lambda t)^{-1}$. When the Hessian
+$H$ of $f$ is strictly positive, using $\gamma^{(t)} =
+(\lambda_{\mathrm{min}}t)^{-1}$, where $\lambda_{\mathrm{min}}$ is the smallest
+eigenvalue of $H$, produces the best convergence speed. Note that $
+(\lambda_{\mathrm{min}}t)^{-1}$ decreases asymptotically with $t$, matching our intuition
+about larger step sizes in early iterations and smaller step sizes in later
+iterations. However, simply using $\gamma^{(t)} =
+(\lambda_{\mathrm{min}}t)^{-1}$ can produce \textit{too} large of steps in early
+iterations. Hence, it often works better to start with some reasonable initial
+estimate $\gamma^{(0)}$ that then decays in the fashion of
+$(\lambda_{\mathrm{min}}t)^{-1}$, leading to the expression in the first
+sentence of this paragraph. By definition, ``reasonable'' is more of a judgment
+call than a mathematical statement: Bottou provides some heuristics for creating
+such an estimate of $\gamma^{(0)}$.
+
+\subsubsection{Momentum and Acceleration}
+
+In our set-up of SGD, we stipulated that the function to be minimized, $f$, is
+convex. Although this is a necessary condition for some proofs of SGD's
+correctness, SGD is widely used in applications---most prominently, neural
+networks and deep
+learning---where this assumption is not
+always true (as covered in more detail in CHAPTER X). In such settings, it is
+possible that there are multiple local minima. The challenge for SGD is
+how to avoid getting ``stuck'' in these local ravines and continue iterating
+until the true global minimum is achieved.
+
+Momentum---also called acceleration---is a common technique for avoiding this. As outlined in
+\cite{rumelhart1986general} and 
+\cite{qian1999momentum}, we add a new acceleration term to the familiar GD
+equation. Let $z^{(t+1)} = \alpha z^{(t)} + \nabla f(w^{(t)})$. GD with
+momentum then is:
+
+\begin{equation}
+	w^{(t+1)} = w^{(t)} - \gamma z^{(t+1)}
+\end{equation}
+
+Now, the weight vector obtained for the current timestep depends both on the
+current gradient \textit{and the gradient of the previous time-step}.
+Intuitively, without acceleration, SGD tends to oscillate within a ravine
+because the gradient is dominated by the steep curves on either side of the
+ravine; acceleration helps point the descent along the shallower curve leading
+out of the ravine. \cite{nesterov} provides another popular implementation of
+acceleration.
 \cite{polyak1992acceleration}
-\cite{nesterov}
 
@@
 
+\cite{shalev2011pegasos}
 \subsection{SGD and Statistical Learning}
+\cite{dal2015calibrating}
 
+Loosely following the set-up of , consider a typical supervised classification problem. We have
+feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
+features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
+predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
+We have a hypothesis class $\mathcal{F}$ consisting of
+various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
+vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
+expresses the cost of mis-classification. Assume for now that $L$ is convex. 
+We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
+that which minimizes empirical risk over all observations: $\frac{1}{n}
+\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
+Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
+have:
+
+\begin{equation}
+	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))
+\end{equation}
+
+Because $L$ is convex, there is one critical point,
+it is the global minimum, and it is located where the gradient of the loss
+function with respect to $\bm{w}$ is zero. GD is an iterative
+algorithm that can be used to numerically approximate this point. We first state
+it in a general form, and then adapt it to the specific problem outlined above.
+
+\begin{equation}
+	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
+	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+\end{equation}
+
+$t$ refers to iterations. $\gamma$ is a parameter controlling the step
+size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. For
+$t=0$ we must supply an initial starting guess, $\bm{w}_0$.
+
+\begin{equation}
+	\left|\frac{1}{n}\sum_{i=1}^n \nabla L(\bm{Y}_i,
+	f_{\bm{w}_t}(\bm{X}_i))\right| \leq \epsilon
+\end{equation}
+
+Note that the batch GD algorithm presented in (2) requires evaluating the
+gradient for every single observation $i$. In a large-$n$ dataset, this can be
+computationally infeasible. SGD's innovation is to instead evaluate only a
+single randomly chosen observation $i$ at each iteration $t$.
+
+\begin{equation}
+	\bm{w}_{t+1} = \bm{w}_t - \gamma 
+	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+\end{equation}
 \section{Conclusion}

### 2018-04-16 paper/paper.tex
more work

@@
 \usepackage{bm}
+\usepackage{graphicx}
+\usepackage[linesnumbered, lined, boxed]{algorithm2e}
+\RestyleAlgo{boxruled}
 
@@
 
-\title{STAT 672 Final Project: Stochastic Gradient Descent}
+\title{STAT 672 Final Project:\\ Stochastic Gradient Descent}
 
@@
 per iteration, we would expect the convergence to require more iterations. But
-per (9) it is guaranteed to \textit{eventually} arrive at the optimal solution.
-Put differently, the difference between GD and SGD must then primarily revolve
+per (9) we expect it to \textit{eventually} arrive at the optimal solution.
+If so, the difference between GD and SGD must then primarily revolve
 around convergence speed, not the final value that is converged to.
@@
 
-We can visually depict SGD's ``noisier'' convergence in comparison to GD.
+Figure 1 visually compares SGD's convergence to that of GD (and
+mini-batch gradient descent, a variation addressed later in this
+paper).\footnote{Image taken from www.towardsdatascience.com} The
+path of SGD's convergence is very squiggly, reflecting the extra variation
+introduced by only using a single observation per iteration. The path of GD's
+convergence is much smoother, reflecting the use of all observations per
+iteration.
+
+\begin{figure}[h!]
+	\centering
+	\caption{Convergence}
+\includegraphics[scale=0.3]{img}
+\end{figure}
 
@@
 the algorithm disallows a comprehensive or detailed treatment of all
-development. This sub-section covers some of the more important and interesting extensions.
+developments. This sub-section covers some of the more important and interesting
+ones.
 
@@
 As hinted at in our proof of convergence to $\hat{w}$, much depends on
-an appropriately chosen step size $\gamma$. If $\gamma$ is too small, (S)GD may
+an appropriately chosen step size $\gamma$. If $\gamma$ is too small, gradient
+descent may
 take an excessively long time to run (since we are only moving a small distance
 ``down'' the gradient every iteration); if $\gamma$ is too large, then an
-iteration's step may ``overshoot'' the critical point and prevent convergence.
+iteration's step may ``overshoot'' the critical point.
 An entire literature has developed around the best way to select $\gamma$. The
@@
 \cite{boyd2004convex}, there are two main variants: \textit{exact} and
-\textit{backtracking}. In exact line search, $\gamma$ is chosen to minimize $f$
-along the ray $\{x + \gamma \Delta x\}$, where $\gamma \in \mathbb{R}^+)$ and
+\textit{backtracking}. In exact line search, $\gamma^{(t)}$ is chosen to minimize $f$
+along the ray $\{x + \gamma^{(t)} \Delta x\}$, where $\gamma^{(t)} \in \mathbb{R}^+)$ and
 $\Delta x$ is the descent direction determined by SGD:
 \begin{equation}
-	\gamma = \argmin_{s \geq 0} f(x + s\Delta x)
+	\gamma^{(t)} = \argmin_{s \geq 0} f(x + s\Delta x)
 \end{equation}
@@
 algorithm stops when $f(x + s \Delta x) \leq f(x) + \alpha s \Delta f(x)^T
-\Delta x$, and the final value of $s$ is taken as $\gamma$.
+\Delta x$, and the final value of $s$ is taken as $\gamma^{(t)}$.
 
@@
 learning---where this assumption is not
-always true (as covered in more detail in CHAPTER X). In such settings, it is
-possible that there are multiple local minima. The challenge for SGD is
+always true. In such settings, it is
+possible that there are multiple local minima, and/or that there are areas where
+the surface curves much more steeply in one dimension than another. The challenge for SGD is
 how to avoid getting ``stuck'' in these local ravines and continue iterating
-until the true global minimum is achieved.
+until the true global minimum is achieved. Even if there is no pathological
+curvature, incorporating information about the local topology may result in more
+efficient algorithms and faster convergence.
 
-Momentum---also called acceleration---is a common technique for avoiding this. As outlined in
+Momentum---also called acceleration---is a common modification of GD. As outlined in
 \cite{rumelhart1986general} and 
@@
 Now, the weight vector obtained for the current timestep depends both on the
-current gradient \textit{and the gradient of the previous time-step}.
-Intuitively, without acceleration, SGD tends to oscillate within a ravine
-because the gradient is dominated by the steep curves on either side of the
-ravine; acceleration helps point the descent along the shallower curve leading
-out of the ravine. \cite{nesterov} provides another popular implementation of
-acceleration.
-\cite{polyak1992acceleration}
+current gradient \textit{and the update of the previous time-step}, with the
+acceleration parameter $\alpha$ governing how much weight is given to the
+previous time-step. 
+This formulation leads to momentum: dimensions whose
+gradients point in the same direction have proportionally greater updates, while
+dimensions whose gradients exhibit strong curvature have proportionally
+lesser updates. The effect is to help stop (S)GD from oscillating in ravines and
+speed up convergence. There are many variations on acceleration, including
+Nesterov accelerated gradient (NAG) (\cite{nesterov}), Adam
+(\cite{kingma2014adam}), AdaGrad
+(\cite{duchi2011adaptive}), and AdaDelta (\cite{zeiler2012adadelta}).
 
@@
 
-\subsubsection{Predictive Variance Reduction}
+Averaged SGD (ASGD)---to the author's knowledge first proposed in 
+\cite{polyak1992acceleration}---is another variation. Typically, we consider
+$\hat{w}$ (the optimal solution) to be that associated with the final iteration
+of SGD, when some stop condition is met. For example, in a situation where we
+run $N$ iterations before stopping, $\hat{w} = w^{(N)}$. In ASGD, we consider $\hat{w}$ 
+the average of $w$ across all iterations, i.e.
+\begin{equation}
+	\hat{w} = \bar{w} = \frac{1}{N} \sum_{t=1}^N w^{(t)}
+\end{equation}
+
+The value proposition of ASGD is that sometimes SGD will oscillate around the
+optimal solution. By taking the average across all updates, ASGD reduces this
+noise and is likely to give a solution closer to the optimum. More sophisticated
+versions of ASGD have been proposed in 
+\cite{zhang2004solving}
+and 
+\cite{xu2011towards}.
+
+\subsubsection{Mini-batch}
+
+Mini-batch gradient descent (MGD) can be regarded as a mid-point between batch
+and stochastic gradient descent in that it uses more than one but less than all of
+the available observations. Suppose we have $n$ observations and divide them
+into $m$ groups called mini-batches, each of equal size $b=\frac{n}{m}$. Now, at
+every iteration $t$, randomly pick one of the mini-batches (all are equally
+likely to be chosen). For that mini-batch:
+
+\begin{equation}
+	w^{(t+1)} = w^{(t)} - \gamma \frac{1}{b} \sum_{i=1}^b \nabla f(w^{(t)})
+\end{equation}
+
+MGD iterates with $t$ and converges to $\hat{w}$ in the normal manner. Every
+update, a new $b$ is randomly chosen and the above algorithm iterated. 
+\cite{dekel2012optimal} and \cite{li2014efficient} analyze MGD's speed of
+convergence and suggest that has some advantageous qualities for
+parallelization, a topic that is explained in more depth in the following
+subsection.
 
@@
 
-SGD is commonly used in large-$n$, computationally demanding applications. Thus,
+SGD is commonly used in large-$n$ applications. Thus,
 even though SGD is a computational improvement over batch GD, there has been
-interest in whether SGD can be made even faster by parallelizing it.
-\cite{zinkevich2010parallelized} present novel algorithms for doing so. The
-actual algorithms are strikingly simple; their proof is highly technical and
-omitted here.
+interest in whether SGD can be made even faster by parallelizing it. As
+background, typical computer applications are serial: problems are broken
+down into a discrete series of instructions that are executed one after another
+on a single processor. Parallel computing calls for breaking down problems into discrete parts that can
+be solved concurrently and are executed simultaneously on multiple processors
+(\cite{barney2010introduction}).
+There are many paradigms of parallel computing but a useful high-level
+distinction is between shared memory, in which all processors share a common
+address space, and distributed memory, in which each processor or group of processors
+has its own address space. The latter has become increasingly popular in
+industry via Apache Hadoop and associated projects (e.g. Spark). Distributed
+memory applications often require very complex computing strategies for how to combine the
+work of many processors without a shared memory space into a single coherent answer.
 
-The parallelization technique essentially is averaging. In line with previous
-notation, suppose we have fixed learning rate $\gamma$,
+This background helps to emphasize the novelty of the parallel
+implementation of SGD proposed in \cite{zinkevich2010parallelized}. Though backed
+by highly technical proofs, the actual algorithm is strikingly simple. Suppose we have $N$ processors. We seek to
+conduct SGD to estimate $w$ across $n$ observations with fixed learning rate
+$\gamma$ for some fixed number of steps $T$.
 
-Advantages
+\medskip
+
+\begin{algorithm}[H]
+	Define $T = \frac{n}{N}$\;
+
+	Randomly partition examples, giving $T$ to each machine;\
+
+	\ForAll{$i \in \{1 \ldots N\}$}{
+		Randomly shuffle data on machine $i$\;
+
+		Initialize $w^{i, (0)}=0$\;
+
+		\ForAll{$t \in \{1 \ldots T\}$}{
+
+	$w^{i, (t+1)} = w^{i, (t)} - \gamma \nabla f(w^{i, (t)})$ \;
+			}
+		}
+	Aggregate from all computers: $\bar{w} = \frac{1}{N} \sum_{i=1}^{N}
+	w^{i, (T)}$ \;
+
+	\Return{$\bar{w}$}
+
+	\caption{Parallel SGD}
+\end{algorithm}
+
+\medskip
+
+Zinkevich et al. show that this algorithm has a number of desirable properties.
+It requires no communication between machines until the end; asymptotically, the
+error approaches zero; and, the amount of time required is independent of the number of
+examples (though this is an inherent property of SGD and is not unique to their
+algorithm). Further improvements have been proposed: e.g., the Hogwild algorithm
+presented in \cite{recht2011hogwild} deals with locking issues in parallel
+computation of SGD.
 
@@
 
+In theory, SGD is agnostic to $f$ and $w$ as long as they meet required
+assumptions, and so can be used in many different areas of optimization. In
+practice, SGD is overwhelmingly popular in the statistical learning
+field for estimating weight coefficients in predictive models. 
+This section lays out some reasons for SGD's popularity; conducts a case study
+using SGD; and identifies notable real-world applications of SGD.
+
+\subsection{SGD and Statistical Learning}
 \cite{shalev2011pegasos}
-\subsection{SGD and Statistical Learning}
-\cite{dal2015calibrating}
 
@@
 \end{equation}
+
+\subsection{Case Study}
+
+\cite{dal2015calibrating}
+
+\subsection{Real-World Applications}
+
+If a Silicon Valley press release uses words like ``artificial intelligence'',
+``machine learning'', and ``big data'' to describe a new product or service, it very likely
+uses SGD in some way. Such is the extent of SGD's popularity. 
+Below is a list of particularly prominent examples of SGD's practical
+applications.
+
+\begin{itemize}
+	\item \textbf{ImageNet} is a dataset consisting of millions
+		of high-resolution, labeled images. The
+		ImageNet Large Scale Visual Recognition Competition (ILSVRC) is
+		an annual contest in which competing teams train models for
+		automatic image classification. The submission of \cite{krizhevsky2012imagenet} 
+		for the 2012 ILSVRC used convolutional neural
+		nets to achieve dramatic reduction in error rates and is
+		one of the most cited machine learning papers of all time. Per the aforementioned paper: ``We
+		trained our models using stochastic gradient descent with a
+		batch size of 128 examples, momentum of 0.9, and weight decay of
+		0.00005.''
+	\item Google's \textbf{AlphaGo} software. AlphoGo applies deep learning to the
+		board game of Go and has received extensive attention for its
+		ability to defeat even top human players (e.g., twice making the
+		cover of \textit{Nature} with \cite{silver2016mastering} and
+		\cite{silver2017mastering}). Google's slides from the 2016
+		International Conference on Machine Learning (ICML) indicate
+		that SGD is used in every stage of the
+		model: supervised learning of policy networks, reinforcement
+		learning of policy networks, and reinforcement learning of value
+		networks.
+	\item The \textbf{Netflix Grand Prize} was a competition hosted by
+		Netflix in which teams competed to develop models for
+		using Netflix data to make better movie recommendations
+		(\cite{bennett2007netflix}). Most of the top entries used
+		SGD for estimating model coefficients; for example,
+		\cite{koren2009bellkor} states that ``in order to learn the
+		involved parameters$\ldots$ learning is done by a stochastic
+		gradient descent algorithm running for 30 iterations.'' 
+	\item Google's \textbf{word2vec} natural language processing (NLP)
+		software powers Google Translate, among other applications.
+		\cite{mikolov2013distributed} and \cite{rong2014word2vec}
+		discuss the continuous bag of words (CBOW) model underlying
+		word2vec and how SGD is used to estimate coefficients
+		in the model.
+\end{itemize}
+
 \section{Conclusion}

### 2018-04-17 paper/paper.tex
svm

@@
 field for estimating weight coefficients in predictive models.
-This section lays out some reasons for SGD's popularity; conducts a case study
-using SGD; and identifies notable real-world applications of SGD.
 
 \subsection{SGD and Statistical Learning}
-\cite{shalev2011pegasos}
 
-Loosely following the set-up of , consider a typical supervised classification problem. We have
-feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
+Consider a typical supervised classification problem. We have
+feature matrix $\bm{X}_{n \times D}$ (corresponding to $n$ observations and $D$
 features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
-predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
+predict a particular observation's label $Y_i$ using that observation's features
+$\bm{X}_i$. To emphasize the utility of SGD, suppose that both $n$ and $D$ are
+very large.
 We have a hypothesis class $\mathcal{F}$ consisting of
 various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
-vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
-expresses the cost of mis-classification. Assume for now that $L$ is convex. 
+vector $\bm{w}$. We have convex loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
+expresses the cost of misclassification. 
 We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
@@
 
-Because $L$ is convex, there is one critical point,
-it is the global minimum, and it is located where the gradient of the loss
-function with respect to $\bm{w}$ is zero. GD is an iterative
-algorithm that can be used to numerically approximate this point. We first state
-it in a general form, and then adapt it to the specific problem outlined above.
+How shall we solve this optimization problem? Newton's method requires
+calculating the Hessian, which will have $D^2$ entries; since $D$ is large,
+this is inconvenient to compute. We could try batch gradient descent:
 
 \begin{equation}
-	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
-	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
+	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \gamma \frac{1}{n}\sum_{i=1}^n
+	\nabla_{\bm{w}} L(Y_i, f_{\bm{w}^{(t)}}(\bm{X}_i))
 \end{equation}
 
-$t$ refers to iterations. $\gamma$ is a parameter controlling the step
-size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. For
-$t=0$ we must supply an initial starting guess, $\bm{w}_0$.
+But, in line with our analysis in section 2.2.2, batch GD requires evaluating
+the gradient over every observation $i \leq n$ for every iteration $t$. Since we have defined $n$ to be
+large, this is computationally inconvenient. Rather, we use SGD to evaluate only
+a single randomly chosen observation per iteration.
 
 \begin{equation}
-	\left|\frac{1}{n}\sum_{i=1}^n \nabla L(\bm{Y}_i,
-	f_{\bm{w}_t}(\bm{X}_i))\right| \leq \epsilon
+	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \nabla_{\bm{w}} L(Y_i, f_{\bm{w}^{(t)}}(\bm{X}_i))
 \end{equation}
 
-Note that the batch GD algorithm presented in (2) requires evaluating the
-gradient for every single observation $i$. In a large-$n$ dataset, this can be
-computationally infeasible. SGD's innovation is to instead evaluate only a
-single randomly chosen observation $i$ at each iteration $t$.
+Per previous analyses we know that this approach will converge to an arbitrarily
+good answer (if certain assumptions are met); will converge to that arbitrary
+degree of precision in less time than plain gradient descent; does not
+necessarily require computing the Hessian; and can be parallelized across many
+machines. These attractive qualities make SGD well-represented in almost all areas of
+machine learning. In the following subsection, we take a closer look at one
+particularly prominent applications, support vector machines.
 
-\begin{equation}
-	\bm{w}_{t+1} = \bm{w}_t - \gamma 
-	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
-\end{equation}
+\subsubsection{Support vector machines}
+
+Originally proposed by \cite{boser1992training} and \cite{cortes1995support}, support vector machines (SVM)
+are among the best supervised learning algorithms. Here we set up the SVM
+problem and then show how SGD can be used to solve it.
+
+\cite{shalev2011pegasos}
 

### 2018-04-17 paper/paper.tex
close to done

@@
 We have a hypothesis class $\mathcal{F}$ consisting of
-various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
-vector $\bm{w}$. We have convex loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
+various functions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
+vector $\bm{w}$. We have convex loss function $L(Y_i, f(\bm{w}, \bm{X}_i))$ that
 expresses the cost of misclassification. 
-We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
+We will consider the optimal function
 that which minimizes empirical risk over all observations: $\frac{1}{n}
-\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
+\sum_{i=1}^n L(Y_i, f(\bm{w}, \bm{X}_i))$.
 Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
@@
 \begin{equation}
-	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))
+	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f(\bm{w}, \bm{X}_i))
 \end{equation}
@@
 	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \gamma \frac{1}{n}\sum_{i=1}^n
-	\nabla_{\bm{w}} L(Y_i, f_{\bm{w}^{(t)}}(\bm{X}_i))
+	\nabla_{\bm{w}} L(Y_i, f(\bm{w}^{(t)}, \bm{X}_i))
 \end{equation}
@@
 \begin{equation}
-	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \nabla_{\bm{w}} L(Y_i, f_{\bm{w}^{(t)}}(\bm{X}_i))
+	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \nabla_{\bm{w}} L(Y_i, f(\bm{w}^{(t)}, \bm{X}_i))
 \end{equation}
@@
 machine learning. In the following subsection, we take a closer look at one
-particularly prominent applications, support vector machines.
+particularly prominent application, support vector machines.
 
@@
 are among the best supervised learning algorithms. Here we set up the SVM
+optimization
 problem and then show how SGD can be used to solve it.
 
-\cite{shalev2011pegasos}
+Suppose we have labels $\bm{Y}$ with $Y_i \in \{-1,1\}$ and features $\bm{X}_{n
+\times D}$, corresponding to $n$ observations and $D$ features.
+We want to construct a decision hyperplane 
+$\langle \bm{w}, \bm{X} \rangle + w_0= 0$. $\bm{w}$ is
+a weight vector that we apply to $\bm{X}$. If for a particular observation $i$
+$\langle \bm{w}, \bm{X}_i \rangle + w_0 < 0$, we
+classify $Y_i=-1$. Similarly, if $\langle \bm{w}, \bm{X}_i \rangle + w_0 > 0$, we
+classify $Y_i = 1$. We introduce slack variables $\xi_i$ to allow some
+violation of the margin, as well as a cost parameter $C$ to control them. The
+primal soft-margin SVM problem thus is:
 
-\subsection{Case Study}
+\begin{equation}
+	\min_{\bm{w} \in \mathbb{R}^D, w_0 \in \mathbb{R}, \bm{\xi} \in \mathbb{R}^n} 
+	\frac{1}{2} ||\bm{w}||^2 + C \frac{1}{n} \sum_{i=1}^n \xi_i
+\end{equation}
+$$
+s.t. \quad Y_i(\langle \bm{w}, \bm{X}_i \rangle + w_0) \geq 1 - \xi_i 
+$$
+$$
+\quad \xi_i \geq 0 
+$$
 
-\cite{dal2015calibrating}
+This problem can be recharacterized through the lens of empirical risk
+minimization. Suppose we obtain an optimal answer $\bm{w}^*, w_0^*, \bm{\xi}^*$. 
+Then, it makes sense that:
 
+$$
+\xi_i^* = \max(0, 1 - Y_i(\langle \bm{w}, \bm{X}_i \rangle + w_0)) 
+$$
+
+We now can recharacterize the problem as:
+
+$$
+\min_{\bm{w}, w_0} C \frac{1}{n} \sum_{i=1}^n \left[ \max(0, 1 - Y_i(\langle \bm{w},
+\bm{X}_i \rangle +
+w_0)) + \frac{1}{2}||\bm{w}||^2 \right]
+$$
+
+This is equivalent to:
+\begin{equation}
+\min_{f \in \mathcal{F}} \frac{1}{n} \sum_{i=1}^n \left[ L(Y_i, \bm{w},
+	\bm{X}_i) + \frac{\lambda}{2}
+||\bm{w}||^2 \right]
+\end{equation}
+
+Where $\mathcal{F}$ is the linear hypothesis class, $L$ is the hinge loss
+
+$$
+L_{\mathrm{hinge}}(Y_i, \bm{w}, \bm{X}_i) = \max(0, 1 - Y_i
+f(\bm{w}, \bm{X}_i))
+$$
+$$
+f(\bm{w}, \bm{X}_i) = \bm{w}^T\bm{X}_i + w_0
+$$
+
+and
+$\lambda=\frac{1}{2C}$. We use $\frac{\lambda}{2}$ for reasons of
+convenience. This problem can be approached in different ways;
+for example, it can be shown that the Lagrangian dual depends only on
+``support vectors'' (observations violating the margin), leading to useful
+data-reduction properties. Gradient descent is a
+popular method to solve the primal formulation given here. First note
+that the hinge loss function is not differentiable. We must use a sub-gradient:
+
+\begin{equation}
+	\frac{\partial L}{\partial \bm{w}} = 
+	\begin{cases}
+		-Y_i \bm{X}_i \quad \mathrm{if} \hspace{0.2em} Y_if(\bm{w}, \bm{X}_i) < 1\\
+		0 \quad \mathrm{otherwise} \\
+	\end{cases}
+\end{equation}
+
+The basic batch GD approach
+
+$$
+	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
+	\nabla_{\bm{w}} \left[ L(Y_i, \bm{w},
+	\bm{X}_i) + \frac{\lambda}{2}
+||\bm{w}||^2\right]
+$$
+
+$$
+	= \bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
+	 \left[ L\nabla_{\bm{w}} (Y_i, \bm{w},
+	\bm{X}_i) + \lambda \bm{w} \right]
+$$
+thus becomes
+\begin{equation}
+	\bm{w}^{(t+1)} = 
+	\begin{cases}
+		\bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
+		\left[
+			\lambda \bm{w}^{(t)} - Y_i\bm{X}_i
+		\right]
+		\quad \mathrm{if} \hspace{0.2em} Y_if(\bm{w}^{(t)}, \bm{X}_i) < 1 \\
+
+		\bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
+		\lambda \bm{w}^{(t)} 
+		\quad \textrm{otherwise} \\
+
+	\end{cases}
+\end{equation}
+
+As previously discussed, batch GD is computationally expensive and
+so for large $n$ we prefer SGD, selecting only a single randomly chosen
+observation per iteration.
+
+\begin{equation}
+	\bm{w}^{(t+1)} = 
+	\begin{cases}
+		\bm{w}^{(t)} - \gamma 
+		\left(
+			\lambda \bm{w}^{(t)} - Y_i\bm{X}_i
+		\right)
+		\quad \mathrm{if} \hspace{0.2em} Y_if(\bm{w}^{(t)}, \bm{X}_i) < 1 \\
+
+		\bm{w}^{(t)} - \gamma 
+		\lambda \bm{w}^{(t)} 
+		\quad \textrm{otherwise} \\
+
+	\end{cases}
+\end{equation}
+
+There have been further efforts to improve SGD for the specific application of
+SVM, most notably \cite{shalev2011pegasos}.
+
+\subsection{Simulation Case Study}
+
+Simulation provides a useful example of the computational advantages of SGD for
+estimating coefficients in a predictive model. This section outlines such a
+brief simulation, annotated with Python code.
+
+Suppose we have feature matrix $\bm{X}_{n \times D}$ with $n=10^6$ and $D=1$
+and outcome variable $\bm{Y}_i \sim N(\bm{X}_i,1)$.
+
+We seek to estimate coefficients in a
+regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
+squares. We do so via two
+competing methods: solving directly via $\hat{\bm{\beta}} =
+(\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ and iterative estimation via SGD.
+
+We compare the estimated coefficients and run-time.
 \subsection{Real-World Applications}

### 2018-04-18 paper/paper.tex
case study

@@
 \usepackage{graphicx}
+
 \usepackage[linesnumbered, lined, boxed]{algorithm2e}
@@
 \addbibresource{paper.bib}
+
+\usepackage{listings}
+\lstset{frame = single,
+breakatwhitespace=false,
+breaklines=true,
+numbersep=5pt,
+showspaces=false,
+showtabs=false,
+tabsize=2,
+basicstyle=\footnotesize}
 
@@
 \textbf{Introduction} section gives intuitive motivation for stochastic gradient
-descent (SGD). The \textbf{Method and Theory} section more rigorously presents the
-mathematics of SGD and some of its notable properties. The
-\textbf{Applications} sections highlights the real-world settings and uses of SGD,
-including a case study data analysis. The \textbf{Conclusion} section summarizes
+descent (SGD). The \textbf{Method and Theory} section more rigorously derives
+SGD, outlines convergence properties, and reviews extensions to the basic
+algorithm. The \textbf{Applications} sections highlights SGD's role in
+statistical learning, including a simulation analysis to demonstrate SGD's utility
+in large-$n$ settings. The \textbf{Conclusion} section summarizes
 overall findings.
@@
 even though SGD is a computational improvement over batch GD, there has been
-interest in whether SGD can be made even faster by parallelizing it. As
-background, typical computer applications are serial: problems are broken
-down into a discrete series of instructions that are executed one after another
-on a single processor. Parallel computing calls for breaking down problems into discrete parts that can
-be solved concurrently and are executed simultaneously on multiple processors
-(\cite{barney2010introduction}).
-There are many paradigms of parallel computing but a useful high-level
-distinction is between shared memory, in which all processors share a common
-address space, and distributed memory, in which each processor or group of processors
-has its own address space. The latter has become increasingly popular in
-industry via Apache Hadoop and associated projects (e.g. Spark). Distributed
-memory applications often require very complex computing strategies for how to combine the
-work of many processors without a shared memory space into a single coherent answer.
-
-This background helps to emphasize the novelty of the parallel
-implementation of SGD proposed in \cite{zinkevich2010parallelized}. Though backed
+interest in whether SGD can be made even faster by parallelizing it.
+\cite{zinkevich2010parallelized} present a novel technique for doing so. Though backed
 by highly technical proofs, the actual algorithm is strikingly simple. Suppose we have $N$ processors. We seek to
@@
 
-Simulation provides a useful example of the computational advantages of SGD for
-estimating coefficients in a predictive model. This section outlines such a
-brief simulation, annotated with Python code.
+Simulation provides a useful example of the computational advantages of SGD. 
+Suppose we have one-dimensional feature matrix $\bm{X}$ with $X_i \sim N(0, 1)$ 
+and outcome variable $\bm{Y}$ with the
+relationship $Y_i \sim (0.5 X_i) + \epsilon$. There are 10 million observations.
 
-Suppose we have feature matrix $\bm{X}_{n \times D}$ with $n=10^6$ and $D=1$
-and outcome variable $\bm{Y}_i \sim N(\bm{X}_i,1)$.
+\begin{lstlisting}[language=Python]
+import numpy as np
+X = np.random.randn(int(10e6))
+Y = (0.5 * X) + np.randn(1)
+\end{lstlisting}
 
-We seek to estimate coefficients in a
-regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
-squares. We do so via two
-competing methods: solving directly via $\hat{\bm{\beta}} =
-(\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ and iterative estimation via SGD.
-
+We seek to estimate coefficients in a regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
+squares. We do so via two competing methods: batch GD and SGD.
+\begin{lstlisting}[language=Python]
+def GD(X, Y, gamma, epsilon, maxiter):
+	n = Y.size
+	w = 1 #initial guess
+	for i in range(maxiter):
+		Yhat = np.dot(X, w)
+		update = w - gamma * (np.dot(X.T, Yhat - Y) / n)
+		if (abs(w - update) < epsilon): 
+			break
+		else:
+			w = update
+	return(w)
+\end{lstlisting}
+\begin{lstlisting}[language=Python]
+def SGD(X, Y, gamma, epsilon, maxiter):
+	n = Y.size
+	w = 1 #initial guess
+	for i in range(maxiter):
+		j = np.random.randint(0, n)
+		Yhat = w * X[j]
+		update = w - gamma * ((Yhat - Y[j])*X[j])
+		if (abs(w - update) < epsilon):
+			break
+		else:
+			w = update
+	return(w)
+\end{lstlisting}
 We compare the estimated coefficients and run-time.

### 2018-04-18 paper/paper.py

@@ -1,0 +1,33 @@
+import numpy as np
+
+X = np.random.randn(int(10e6))
+Y = (0.5 * X) + np.random.randn(1)
+
+def GD(X, Y, gamma, epsilon, maxiter):
+	n = Y.size
+	w = 1 #initial guess
+	for i in range(maxiter):
+		Yhat = np.dot(X, w)
+		update = w - gamma * (np.dot(X.T, Yhat - Y) / n)
+		if (abs(w - update) < epsilon): 
+			break
+		else:
+			w = update
+	return(w)
+
+def SGD(X, Y, gamma, epsilon, maxiter):
+	n = Y.size
+	w = 1 #initial guess
+	for i in range(maxiter):
+		j = np.random.randint(0, n)
+		Yhat = w * X[j]
+		update = w - gamma * ((Yhat - Y[j])*X[j])
+		if (abs(w - update) < epsilon):
+			break
+		else:
+			w = update
+	return(w)
+#print(GD(X, Y, 0.05, 0.0001, 10000000))
+#print(SGD(X, Y, 0.05, 0.00001, 10000000))
+
+

### 2018-04-18 paper/paper.tex
update

@@
 
-Optimization is fundamental to statistical modeling. The chief task of
-statistical modeling is to characterize the relationship between 
-explanatory variables and an outcome variable, and the chief method for doing so
-is to estimate values for coefficients that best relate
-each explanatory variable to the outcome variable. The term ``best''
-implies picking coefficient values that
-maximize some measure of goodness (e.g. likelihood) or minimize some measure of
-badness (e.g. loss function). 
-Mathematical optimization is the typical
-route to achieving such minimization or maximization. Two important
-considerations for optimization are parametric assumptions and computational
-complexity. SGD, an optimization technique, is
-particularly motivated by these considerations.
+SGD is an optimization algorithm commonly used for fitting coefficients
+in statistical models, particularly in non-parametric and large-$n$ settings.
+Before covering exactly how SGD works, we contextualize its advantages in these areas.
 
@@
 
-Assuming that the outcome variable follows a
-particular statistical distribution aids the computation of optimal coefficients. 
+Consider a feature matrix $\bm{X}$ and outcome variable $\bm{Y}$, and the task
+of fitting coefficients in a model to predict $Y_i$ based on $\bm{X}_i$. 
+Assuming that particular model elements follow a
+particular distribution aids the estimation of coefficients. 
 For example, assumptions in ordinary least squares (OLS)
 regression---assumptions that readers almost certainly are familiar with and so will not be
-repeated here---allow a closed form solution. The optimization problem of
-choosing coefficient $\hat{\bm{\beta}}$ that minimize squared error is solved
-by $\hat{\bm{\beta}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ (where $\bm{X}$ is the
-feature matrix and $\bm{Y}$ the outcome variable).
+repeated here---allow a closed form solution, $\hat{\bm{\beta}} = (\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$.
 
@@
 $$
-\hat{\bm{\beta}}_{n+1} = \hat{\bm{\beta}}_{n} -
-\bm{H}^{-1}(\hat{\bm{\beta}}_n)\bm{J}(\hat{\bm{\beta}}_n)
+\hat{\bm{\beta}}^{(t+1)} = \hat{\bm{\beta}}^{(t)} -
+\bm{H}^{-1}(\hat{\bm{\beta}}^{(t)})\bm{J}(\hat{\bm{\beta}}^{(t)})
 $$
 
-$\bm{J}$ is the Jacobian (the
-first derivative of the log-likelihood function $l$ with respect to each $w_j$)
-and $\bm{H}$ is the Hessian (the second derivative of $l$ with respect to $w_j,
-w_{j'}$). The practicality of Newton-Raphson thus depends on whether it is convenient to
-find $\bm{J}$ and $\bm{H}$. It is convenient for logistic regression
+$(t)$ refers to a particular iteration, $\bm{J}$ is the Jacobian,
+and $\bm{H}$ is the Hessian. The practicality of Newton-Raphson thus depends on whether it is convenient to
+find $\bm{J}$ and in particular $\bm{H}$. It is convenient for logistic regression
 because parametric and independent-and-identically-distributed (IID) assumptions
 mean $l$ is a simple sum of the log probability distribution
-function (PDF, in this case binomial) for each observation. We ``know'' (assume)
-the form of this PDF and so are
-confident that the second derivative exists and is not too onerous to calculate.
+function for each observation. We ``know'' (assume) the form of this PDF and so are
+confident that the second derivative exists and is not onerous.
 In non-parametric settings, we often cannot be so certain and face the possibility of
@@
 The need to conduct optimization in non-parametric settings is a chief
-motivation for gradient descent (GD), of which SGD is a variant. In
-non-parametric settings---most notably supervised and unsupervised statistical
-learning, in which we again seek to find optimal coefficients to relate input
-variables to output variables for the purposes of classification or regression---there 
-typically is no closed form solution for the coefficients. It
-also may not be convenient to find and evaluate the Hessian, making
-Newton-Raphson undesirable. SGD does not require any parametric assumptions. 
+motivation for gradient descent (GD), of which SGD is a variant. SGD does not require any parametric assumptions. 
 In its most basic form, SGD only requires finding the gradient (though some extensions
 do need the Hessian or an approximation to it). 
-SGD thus is well-suited for non-parametric settings.
+SGD thus is well-suited for supervised and unsupervised statistical
+learning, where parametric assumptions often are not made and 
+Newton-Raphson and other second-order methods often are undesirable.
 
@@
 requires an excessive amount of time to do so. ``Plain'' or
-``batch'' GD requires evaluating the gradient for every single observation,
-every single iteration, until the algorithm converges. For example, for a
+``batch'' GD requires evaluating the gradient for every observation,
+every iteration, until the algorithm converges. For example, for a
 dataset of $n=10^6$ that required 25 iterations to converge, batch GD would require 
 evaluating the gradient $25 \times 10^6$ times. This scaling with
-$n$ can cause untenably long computation time.
+$n$ can cause untenably long computation time. Similarly, note that if the
+Jacobian of some function has $D$ elements, the Hessian will have $D^2$. In
+high-dimensional settings, this too can be computationally intractable.
 
-SGD alleviates these computational difficulties by requiring the gradient to be
-evaluated for only a single randomly chosen observation per iteration. This
+In contrast, SGD evaluates the gradient for only a single randomly chosen observation per iteration. This
 approach means convergence is ``noisier'' and hence requires more iterations to
-converge, but each iteration is less complex to compute and so can be done
-faster. SGD thus scales much more favorably with $n$ than GD, and so is
-particularly useful for large-$n$ applications such as machine learning
-and big data problems.
+converge, but each iteration is less compute-intensive and so can be done
+faster. SGD thus scales more favorably with $n$ than does GD and so is
+useful for large-$n$ applications.
 
@@
 
-This section presents the basic form of GD and SGD to set up a more detailed
-examination in subsequent sections. Suppose we want to minimize a function $f: \mathbb{R}^D \to \mathbb{R}$.
+Suppose we want to minimize a function $f: \mathbb{R}^D \to \mathbb{R}$.
 
@@
 respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (1),
-is denoted $\hat{w}$. The iterative GD algorithm for finding $\hat{w}$ is:
+is denoted $\tilde{w}$. We are willing to settle for a close approximation
+$\hat{w}$. The iterative GD algorithm for finding $\hat{w}$ is:
 
@@
 $t$ refers to a particular iteration. Assume that we have supplied an initial
-starting guess for $w$ for $t=0$, $w^{(0)}$. $\gamma$ refers to step size (also called
+starting guess $w^{(0)}$. $\gamma$ refers to step size (also called
 learning rate). Assume for now that $\gamma$ is fixed. The GD algorithm iterates
@@
 A common stopping condition is when the L2 norm of the gradient
-is less than some arbitrarily small constant.
+is less than some arbitrarily small constant. Obviously, the smaller the
+constant, the closer $\hat{w}$ will be to $\tilde{w}$.
 
@@
 $\bm{X}_{n \times D}$ the matrix of these observations, with $X_i$ being a
-particular observation or row in that matrix. We apply
-$f$ over all observations, i.e. $\frac{1}{n} \sum_{i=1}^n f(w, X_i)$. Our new
+row in that matrix. We apply $f$ over all observations, i.e. $\frac{1}{n} \sum_{i=1}^n f(w, X_i)$. Our new
 problem is:
@@
 
-As it turns out, SGD converges to $\hat{w}$ and does so in a computationally
+As it turns out, SGD asymptotically converges to $\tilde{w}$ and does so in a computationally
 advantageous way compared to GD.
@@
 
-This section gives both intuition and (partial) proof for why (S)GD converges to
-$\hat{w}$.
+This section gives intuition and partial proof for why SGD converges to
+$\tilde{w}$.
 
@@
 zero, then we have found the critical point, which must be the value of $w$ that
-minimizes $f$ and hence is $\hat{w}$ (or at least a close approximation to it).
+minimizes $f$ and hence is $\tilde{w}$ (or at least a close approximation to it).
 
@@
 less information per iteration than GD, it requires more iterations to achieve
-some fixed degree of accuracy $\rho$.
+fixed degree of accuracy $\rho$.
 However, row 2 shows that because GD computes the gradient for every observation every iteration,
-while SGD only does so for a single randomly chosen iteration, GD's time per
+while SGD only does so for a single randomly observation, GD's time per
 iteration scales linearly with $n$ while SGD's is a constant. Thus, we conclude
@@
 The basic SGD algorithm has been extended in many different ways. The popularity of
-the algorithm disallows a comprehensive or detailed treatment of all
-developments. This sub-section covers some of the more important and interesting
+the algorithm disallows a comprehensive treatment of all
+developments. This sub-section covers some of the more important
 ones.
@@
 
-As hinted at in our proof of convergence to $\hat{w}$, much depends on
+Much depends on
 an appropriately chosen step size $\gamma$. If $\gamma$ is too small, gradient
@@
 \end{equation}
-This obviously is an optimization problem in and of itself, and
+This is an optimization problem in and of itself, and
 so it can be computationally impractical to add this burden in addition to the
-``main'' optimization problem we are trying to solve using SGD. For this reason,
-\textit{backtracking} is an iterative approximation method that is
+``main'' problem we are trying to solve using SGD. \textit{Backtracking} is an iterative approximation method that is
 computationally lighter. The backtracking algorithm takes two hyper-parameters
@@
 estimate $\gamma^{(0)}$ that then decays in the fashion of
-$(\lambda_{\mathrm{min}}t)^{-1}$, leading to the expression in the first
-sentence of this paragraph. By definition, ``reasonable'' is more of a judgment
+$(\lambda_{\mathrm{min}}t)^{-1}$. By definition, ``reasonable'' is more of a judgment
 call than a mathematical statement: Bottou provides some heuristics for creating
@@
 
-In our set-up of SGD, we stipulated that the function to be minimized, $f$, is
-convex. Although this is a necessary condition for some proofs of SGD's
-correctness, SGD is widely used in applications---most prominently, neural
+In our set-up of SGD, we stipulated that the objective function is
+convex. However, SGD is widely used in applications---most prominently, neural
 networks and deep
@@
 dimensions whose gradients exhibit strong curvature have proportionally
-lesser updates. The effect is to help stop (S)GD from oscillating in ravines and
+lesser updates. The effect is to help stop SGD from oscillating in ravines and
 speed up convergence. There are many variations on acceleration, including
@@
 
-Averaged SGD (ASGD)---to the author's knowledge first proposed in 
+Averaged SGD (ASGD)---first proposed in 
 \cite{polyak1992acceleration}---is another variation. Typically, we consider
-$\hat{w}$ (the optimal solution) to be that associated with the final iteration
+$\hat{w}$ to be the final iteration
 of SGD, when some stop condition is met. For example, in a situation where we
-run $N$ iterations before stopping, $\hat{w} = w^{(N)}$. In ASGD, we consider $\hat{w}$ 
-the average of $w$ across all iterations, i.e.
+run $N$ iterations, $\hat{w} = w^{(N)}$. In ASGD, we consider $\hat{w}$ 
+the average across all iterations.
+
 \begin{equation}
@@
 
-Mini-batch gradient descent (MGD) can be regarded as a mid-point between batch
-and stochastic gradient descent in that it uses more than one but less than all of
-the available observations. Suppose we have $n$ observations and divide them
+Mini-batch gradient descent (MGD) is a mid-point between batch
+and stochastic gradient descent: it uses more than one but less than all
+observations per iteration. Suppose we have $n$ observations and divide them
 into $m$ groups called mini-batches, each of equal size $b=\frac{n}{m}$. Now, at
@@
 
-SGD is commonly used in large-$n$ applications. Thus,
+SGD is common in large-$n$ applications. Thus,
 even though SGD is a computational improvement over batch GD, there has been
-interest in whether SGD can be made even faster by parallelizing it.
-\cite{zinkevich2010parallelized} present a novel technique for doing so. Though backed
-by highly technical proofs, the actual algorithm is strikingly simple. Suppose we have $N$ processors. We seek to
-conduct SGD to estimate $w$ across $n$ observations with fixed learning rate
-$\gamma$ for some fixed number of steps $T$.
+interest in whether SGD can be made even faster via parallelization.
+\cite{zinkevich2010parallelized} present a technique for doing so. Though backed
+by very technical proofs, the actual algorithm is simple. Suppose we have $N$ processors. We seek to
+use SGD to estimate $w$ across $n$ observations with fixed learning rate
+$\gamma$ for fixed number of steps $T$.
 
@@
 error approaches zero; and, the amount of time required is independent of the number of
-examples (though this is an inherent property of SGD and is not unique to their
-algorithm). Further improvements have been proposed: e.g., the Hogwild algorithm
-presented in \cite{recht2011hogwild} deals with locking issues in parallel
-computation of SGD.
+examples. Further improvements have been proposed e.g., the Hogwild algorithm
+presented in \cite{recht2011hogwild}.
 
@@
 
-In theory, SGD is agnostic to $f$ and $w$ as long as they meet required
-assumptions, and so can be used in many different areas of optimization. In
+In theory, SGD is topic-agnostic and so can be used in many different areas of optimization. In
 practice, SGD is overwhelmingly popular in the statistical learning
@@
 Consider a typical supervised classification problem. We have
-feature matrix $\bm{X}_{n \times D}$ (corresponding to $n$ observations and $D$
+feature matrix $\bm{X}_{n \times D}$ ($n$ observations and $D$
 features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
-predict a particular observation's label $Y_i$ using that observation's features
+predict an observation's label $Y_i$ using that observation's features
 $\bm{X}_i$. To emphasize the utility of SGD, suppose that both $n$ and $D$ are
 very large.
-We have a hypothesis class $\mathcal{F}$ consisting of
-various functions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
+We have a hypothesis class $\mathcal{F}$ of unctions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
 vector $\bm{w}$. We have convex loss function $L(Y_i, f(\bm{w}, \bm{X}_i))$ that
@@
 minimization. Suppose we obtain an optimal answer $\bm{w}^*, w_0^*, \bm{\xi}^*$. 
-Then, it makes sense that:
+Then:
 
@@
 This is equivalent to:
+
 \begin{equation}
@@
 $\lambda=\frac{1}{2C}$. We use $\frac{\lambda}{2}$ for reasons of
-convenience. This problem can be approached in different ways;
-for example, it can be shown that the Lagrangian dual depends only on
-``support vectors'' (observations violating the margin), leading to useful
-data-reduction properties. Gradient descent is a
+convenience. This problem could solved using the Lagrangian dual, but SGD is a
 popular method to solve the primal formulation given here. First note
@@
 As previously discussed, batch GD is computationally expensive and
-so for large $n$ we prefer SGD, selecting only a single randomly chosen
+so for large $n$ we prefer SGD, selecting only a single random
 observation per iteration.
@@
 
-Simulation provides a useful example of the computational advantages of SGD. 
+Simulation provides an example of the computational advantages of SGD. 
 Suppose we have one-dimensional feature matrix $\bm{X}$ with $X_i \sim N(0, 1)$ 
-and outcome variable $\bm{Y}$ with the
+and outcome $\bm{Y}$ with the
 relationship $Y_i \sim (0.5 X_i) + \epsilon$. There are 10 million observations.
@@
 
-We seek to estimate coefficients in a regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
+We seek to estimate coefficients in a regression $y = \beta_0 + \beta_1 X$ by least
 squares. We do so via two competing methods: batch GD and SGD.
+
 \begin{lstlisting}[language=Python]
@@
 \end{lstlisting}
-We compare the estimated coefficients and run-time.
+
+Both algorithms were run on the $n=10^6$ dataset with the same step size
+$\gamma$ and precision $\epsilon$. As shown in Table 2, both algorithms 
+converge to essentially the correct answer, but SGD does so in dramatically less time than GD.
+
+\begin{table}[h!]
+	\centering
+	\caption{Runtime comparison}
+	\begin{tabular}{|l l l|}
+		\hline
+		& $\hat{\bm{\beta}_1}$ & \textbf{Time (s)} \\
+		\textbf{GD} & 0.499 & 19.851 \\
+		\textbf{SGD} & 0.507 & 1.832 \\
+		\hline
+	\end{tabular}
+\end{table}
+
 \subsection{Real-World Applications}
@@
 uses SGD in some way. Such is the extent of SGD's popularity. 
-Below is a list of particularly prominent examples of SGD's practical
+Below are some prominent examples of SGD's practical
 applications.
 
-\begin{itemize}
-	\item \textbf{ImageNet} is a dataset consisting of millions
-		of high-resolution, labeled images. The
+\textbf{ImageNet} is a dataset consisting of millions
+of labeled images. The
 		ImageNet Large Scale Visual Recognition Competition (ILSVRC) is
 		an annual contest in which competing teams train models for
-		automatic image classification. The submission of \cite{krizhevsky2012imagenet} 
-		for the 2012 ILSVRC used convolutional neural
+		automatic image classification. The submission of \cite{krizhevsky2012imagenet} used convolutional neural
 		nets to achieve dramatic reduction in error rates and is
@@
 		0.00005.''
-	\item Google's \textbf{AlphaGo} software. AlphoGo applies deep learning to the
+
+Google's \textbf{AlphaGo} software applies deep learning to the
 		board game of Go and has received extensive attention for its
@@
 		networks.
-	\item The \textbf{Netflix Grand Prize} was a competition hosted by
-		Netflix in which teams competed to develop models for
-		using Netflix data to make better movie recommendations
+
+The \textbf{Netflix Grand Prize} was a competition hosted by
+		Netflix to create better movie recommendation models
 		(\cite{bennett2007netflix}). Most of the top entries used
@@
 		gradient descent algorithm running for 30 iterations.''
-	\item Google's \textbf{word2vec} natural language processing (NLP)
-		software powers Google Translate, among other applications.
-		\cite{mikolov2013distributed} and \cite{rong2014word2vec}
-		discuss the continuous bag of words (CBOW) model underlying
-		word2vec and how SGD is used to estimate coefficients
-		in the model.
-\end{itemize}
 
@@
 
-\printbibliography
+SGD is an iterative optimization algorithm. SGD does not require parametric
+assumptions and is computationally attractive. These factors make it popular 
+in statistical learning, where parametric assumptions often are not made and $n$
+and $D$ often are large. There have been many extensions to the basic
+SGD algorithm and it remains a topic of active research. Outside of academia,
+SGD plays a central role in many real-world applications of statistical learning
+such as image recognition, artifical intelligence, and online personalization.
+
+\newpage
+
+\printbibliography[heading=bibnumbered]
 

### 2018-04-18 paper/paper.py

@@ -27,7 +27,7 @@ def SGD(X, Y, gamma, epsilon, maxiter):
 		else:
 			w = update
 	return(w)
-#print(GD(X, Y, 0.05, 0.0001, 10000000))
-#print(SGD(X, Y, 0.05, 0.00001, 10000000))
+#print(GD(X, Y, 0.05, 0.00001, 10000000))
+print(SGD(X, Y, 0.05, 0.00001, 10000000))
 
 

### 2018-04-18 paper/paper.tex
minor changes

@@
 Assuming that particular model elements follow a
-particular distribution aids the estimation of coefficients. 
+known statistical distribution aids the estimation of coefficients. 
 For example, assumptions in ordinary least squares (OLS)
@@
 
-$(t)$ refers to a particular iteration, $\bm{J}$ is the Jacobian,
-and $\bm{H}$ is the Hessian. The practicality of Newton-Raphson thus depends on whether it is convenient to
+$(t)$ refers to a particular iteration, $\bm{J}$ is the Jacobian of the
+log likelihood function $l$,
+and $\bm{H}$ is the Hessian of $l$. The practicality of Newton-Raphson thus depends on whether it is convenient to
 find $\bm{J}$ and in particular $\bm{H}$. It is convenient for logistic regression
@@
 mean $l$ is a simple sum of the log probability distribution
-function for each observation. We ``know'' (assume) the form of this PDF and so are
+function (PDF) for each observation. We ``know'' (assume) the form of this PDF and so are
 confident that the second derivative exists and is not onerous.
@@
 
-We could apply the GD algorithm above to find the optimal $w$.
+We could apply the GD algorithm above to find $\hat{w}$.
 
@@
 
-The above proof is for GD. A rigorous proof is not given for why SGD also
+The above proof is for GD. Proof is not given for why SGD also
 converges to the optimal value. Informally, we can note that the above proof

### 2018-04-25 sgd/sgd.tex
final presentation

@@
 \begin{frame}
-	\frametitle{Optimization is fundamental to statistical modeling}
+	\frametitle{Estimating model coefficients requires optimization}
+	We typically fit statistical models to maximize some measure of goodness
+	(e.g., likelihood) or minimize some measure of badness (e.g., risk)
+	\\~\\
+
+	Parametric assumptions can make this optimization very nice \\~\\
+
+	OLS has a closed form solution: $\hat{\bm{\beta}} =
+	(\bm{X}'\bm{X})\bm{X}'\bm{Y}$ \\~\\
+
+	GLM often uses Newton's method
+	\begin{itemize}
+		\item \small Knowing the PDF makes finding the
+			Hessian (2nd order derivatives) not too onerous \\~\\
+	\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+	\frametitle{Non-parametric \& high-dimensional settings}
 	Suppose that we have a typical supervised classification problem
@@
 		\item Feature vector $\mathbf{X}_i$, label $Y_i$
-		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
+		\item Want to find best prediction function $f^*(\bm{w}; \bm{X})$ from
+			hypothesis class $\mathcal{F}$
 		\item Optimization: pick weights $\hat{\bm{w}}$ that minimize
 			empirical risk according to some convex
-			loss function $L(f_w(x), y)$ \\~\\
+			loss function \\~\\
 	\end{itemize}
 
-	Our lack of assumptions takes away some familiar tools for finding $\hat{\mathbf{w}}$
-	\begin{itemize}
-		\item \small Cannot analytically identify $\hat{\mathbf{w}}$
-			(e.g. in OLS $=(X'X)^{-1}X'Y$))
-		\item \small Newton-Raphson requires 2nd derivative of $L$,
-			which might be a pain (unlike in, for example, GLM)\\~\\
-	\end{itemize}
+	Some familiar tools for finding $\hat{\mathbf{w}}$ no longer work well
+	\\~\\
 
 	But we still know that if $L$ is convex, there is a unique global
-	minimum, and the gradient at that point must be 0
+	minimum, and the gradient at that point must be 0 \\~\\
+
+	Gradient descent is a numerical strategy to search for that point
 \end{frame}
@@
 
-	$\mathbf{w}_{t+1} = \mathbf{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
-	\nabla_w L(f_w(\mathbf{X}_i), y_i)$ \\~\\
+	$\mathbf{w}^{(t+1)} = \mathbf{w}^{(t)} - \gamma \frac{1}{n}\sum_{i=1}^n
+	\nabla_w L(f(\bm{w}^{(t)}; \mathbf{X}_i), Y_i)$ \\~\\
 
@@
 	Step size $\gamma$ can vary over time \\~\\
-
-	Theoretical guarantees on speed of convergence ($\rho :=$ size of error):
-	\begin{itemize}
-		\item Version presented here: $ - \log \rho \sim t$
-		\item More optimized version: $ - \log \log \rho \sim t$ \\~\\
-	\end{itemize}
 
@@
 	$$
-	\textcolor{red}{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \frac{1}{n}
+	\textcolor{red}{\mathbf{w}^{(t+1)}} = \mathbf{w}^{(t)} - \gamma \frac{1}{n}
 	\textcolor{blue}{\sum_{i=1}^n}
-	\nabla_w L(f_w(\mathbf{X}_i), y_i)
+	\nabla_w L(f(\bm{w}^{(t)}; \mathbf{X}_i), Y_i)
 	$$\\~\\
@@
 	\begin{itemize}
-		\item \small If we have $n$ = 10 million, we have to evaluate the
-			gradient 10 million times for every step (and the
-			algorithm may take many steps to converge)\\~\\
-		\item \small May take hours or days to converge \\~\\
+		\item \small If $n$ = 10 million, have to evaluate 
+			gradient 10 million times every step
 	\end{itemize}
@@
 	$$
-	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w
-	L(f_w(\mathbf{X}_{\textcolor{red}{i}}), y_{\textcolor{red}{i})}
+	{\mathbf{w}^{(t+1)}} = \mathbf{w}^{(t)} - \gamma \nabla_w
+	L(f(\bm{w}^{(w)}; \mathbf{X}_{\textcolor{red}{i}}), Y_{\textcolor{red}{i})}
 	$$ \\~\\
@@
 	\begin{figure}
+		\caption{Fitting OLS coefficients, $n=10$ million}
 	\centering
-	\includegraphics[scale=0.55]{comparison}
+	\includegraphics[scale=0.35]{speed}
 	\end{figure}
@@
 \begin{frame}
-	\frametitle{SGD is widely used in industry}
+	\frametitle{SGD is very popular in academia and industry}
+	Many, many statistical learning models use SGD to fit weights \\~\\
+
 	If a Silicon Valley press release uses any of the following phrases...

### 2018-04-29 paper/paper.tex
updated paper
wek

@@
-\documentclass{article}
+\documentclass{report}
 
@@
 
-\usepackage[citestyle=authoryear]{biblatex}
+\usepackage[citestyle=authoryear, style=authoryear]{biblatex}
 \addbibresource{paper.bib}
@@
 
-\title{STAT 672 Final Project:\\ Stochastic Gradient Descent}
+\title{STAT 672 final project:\\ Stochastic gradient descent}
 
@@
 
-\section{Introduction}
+\tableofcontents
 
-\subsection{Organization}
+\newpage
+
+\chapter{Introduction}
+
+\section{Organization}
 
@@
 
-\subsection{Motivation}
+\section{Motivation}
 
@@
 
-\subsubsection{Parametric vs. non-parametric}
+\subsection{Parametric vs. non-parametric}
 
@@
 
-\subsubsection{Computational Complexity}
+\subsection{Computational complexity}
 
@@
 
-\section{Method and Theory}
+\chapter{Method and theory}
 
-\subsection{Basic Form}
+\section{Basic form}
 
@@
 that $f$ is convex and differentiable, i.e. we can compute its gradient with
-respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (1),
+respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (2.1),
 is denoted $\tilde{w}$. We are willing to settle for a close approximation
@@
 
-\subsection{Key Properties}
+\section{Key properties}
 
-\subsubsection{Correctness}
+\subsection{Correctness}
 
@@
 
-Now, suppose we use the GD algorithm presented in (2) with $\gamma \leq
+Now, suppose we use the GD algorithm presented in (2.2) with $\gamma \leq
 \frac{1}{L}$. Denote $x^+ = x - \gamma \nabla
@@
 
-Returning to (8), we obtain:
+Returning to (2.8), we obtain:
 
@@
 
-By definition, a squared L2 norm will always be positive unless its content is
+A squared L2 norm will always be positive unless its content is
 0, and we have defined $\gamma$ to be positive, so $\frac{1}{2} \gamma||\nabla f(x)||_2^2$ will always be positive unless the
-gradient is equal to zero. So, (9) implies that GD results in a strictly
+gradient is equal to zero. So, (2.9) implies that GD results in a strictly
 decreasing objective function value until it reaches the point where the
@@
 per iteration, we would expect the convergence to require more iterations. But
-per (9) we expect it to \textit{eventually} arrive at the optimal solution.
+per (2.9) we expect it to \textit{eventually} arrive at the optimal solution.
 If so, the difference between GD and SGD must then primarily revolve
@@
 
-\subsubsection{Speed}
+\subsection{Speed}
 
-Table 1---a simplified version of that in \cite{bottou2010large}---illustrates
+Table 2.1---a simplified version of that in \cite{bottou2010large}---illustrates
 the computational advantages of SGD over GD. As shown in row 1, because SGD uses
@@
 
-\begin{table}[h!]
+\begin{table}[t]
 	\centering
@@
 
-Figure 1 visually compares SGD's convergence to that of GD (and
+Figure 2.1 visually compares SGD's convergence to that of GD (and
 mini-batch gradient descent, a variation addressed later in this
@@
 
-\begin{figure}[h!]
+\begin{figure}[t]
 	\centering
@@
 
-\subsection{Extensions}
+\section{Extensions}
 
@@
 
-\subsubsection{Step Size}
+\subsection{Step size}
 
@@
 
-\subsubsection{Momentum and Acceleration}
+\subsection{Momentum and acceleration}
 
@@
 
-\subsubsection{Averaging}
+\subsection{Averaging}
 
@@
 
-\subsubsection{Mini-batch}
+\subsection{Mini-batch}
 
@@
 
-\subsubsection{Parallelization}
+\subsection{Parallelization}
 
@@
 
-\section{Applications}
+\chapter{Applications}
 
@@
 
-\subsection{SGD and Statistical Learning}
+\section{SGD and statistical learning}
 
@@
 very large.
-We have a hypothesis class $\mathcal{F}$ of unctions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
+We have a hypothesis class $\mathcal{F}$ of functions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
 vector $\bm{w}$. We have convex loss function $L(Y_i, f(\bm{w}, \bm{X}_i))$ that
@@
 
-\subsubsection{Support vector machines}
+\subsection{Support vector machines}
 
@@
 There have been further efforts to improve SGD for the specific application of
-SVM, most notably \cite{shalev2011pegasos}.
+SVM, most notably the Pegasos algorithm of \cite{shalev2011pegasos}.
 
-\subsection{Simulation Case Study}
+\section{Simulation case study}
 
@@
 relationship $Y_i \sim (0.5 X_i) + \epsilon$. There are 10 million observations.
+We seek to estimate coefficients in a regression $y = \beta_0 + \beta_1 X$ by least
+squares. We do so via two competing methods: batch GD and SGD.
+Both algorithms were run on the $n=10^6$ dataset with the same step size
+$\gamma$ and precision $\epsilon$. As shown in Table 3.1, both algorithms 
+converge to essentially the correct answer, but SGD does so in dramatically less time than GD.
 
+\newpage
+
+\begin{figure}[h!]
+	\caption{Generating data}
 \begin{lstlisting}[language=Python]
@@
 \end{lstlisting}
+\end{figure}
 
-We seek to estimate coefficients in a regression $y = \beta_0 + \beta_1 X$ by least
-squares. We do so via two competing methods: batch GD and SGD.
-
+\begin{figure}[h!]
+	\caption{Batch GD algorithm}
 	\begin{lstlisting}[language=Python]
@@
 	\end{lstlisting}
+\end{figure}
+
+\begin{figure}[h!]
+	\caption{SGD algorithm}
 \begin{lstlisting}[language=Python]
@@
 \end{lstlisting}
-
-Both algorithms were run on the $n=10^6$ dataset with the same step size
-$\gamma$ and precision $\epsilon$. As shown in Table 2, both algorithms 
-converge to essentially the correct answer, but SGD does so in dramatically less time than GD.
+\end{figure}
 
@@
 
-\subsection{Real-World Applications}
+\newpage
+
+\section{Real-world applications}
 
@@
 
-\textbf{ImageNet} is a dataset consisting of millions
+\subsection{ImageNet}
+ImageNet is a dataset consisting of millions
 of labeled images. The
@@
 		0.00005.''
-
-Google's \textbf{AlphaGo} software applies deep learning to the
+\subsection{AlphaGo}
+Google's AlphaGo software applies deep learning to the
 		board game of Go and has received extensive attention for its
@@
 
-The \textbf{Netflix Grand Prize} was a competition hosted by
+\subsection{Netflix grand prize}
+The Netflix Grand Prize was a competition hosted by
 		Netflix to create better movie recommendation models
@@
 
-\section{Conclusion}
+\chapter{Conclusion}
 

### 2018-04-29 sgd/sgd.tex
wek
werk

@@
 
-	Parametric assumptions can make this optimization very nice \\~\\
+	Parametric assumptions can make this optimization ``nice'' \\~\\
 
@@
 	\frametitle{SGD is very popular in academia and industry}
-	Many, many statistical learning models use SGD to fit weights \\~\\
+	Many, many, \textit{many} statistical learning models use SGD to fit weights \\~\\
 
@@
 
-	Example: Google's \textbf{AlphaGo} program.
 \end{frame}
@@
 \begin{frame}
+	\frametitle{Example: Google's \textbf{AlphaGo} program}
 	\begin{figure}[b]

### 2018-05-04 paper/paper.tex
final

@@
 
+\usepackage{geometry}
+\geometry{margin=1.5in}
+
 \author{Tom Wallace}
 
-\title{STAT 672 final project:\\ Stochastic gradient descent}
+\title{STAT 672 Final Project: Stochastic Gradient Descent}
 
@@
 
-\subsection{Parametric vs. non-parametric}
+\subsection{Parametric vs. non-parametric approaches}
 
@@
 
-\begin{table}[t]
+\begin{table}[h]
 	\centering
@@
 
-Averaged SGD (ASGD)---first proposed in 
+Averaged SGD (ASGD)---proposed in
 \cite{polyak1992acceleration}---is another variation. Typically, we consider
@@
 
+\subsection{Neural networks and back-propagation}
+
+Neural networks are another class of statistical learning
+models for which SGD plays a fundamental role in estimating weights, in
+particular through a process called back-propagation. Space
+constraints disallow a detailed examination along the lines of that conducted
+for SVM. This subsection aims only to note that SGD is widely used with neural
+networks, and that this usage provides the practical motivation behind many of
+the extensions to the SGD algorithm covered in Section 2.3. This is not to say
+that these techniques are \textit{solely} intended for neural networks---indeed,
+many of them precede neural networks' current popularity---but
+their utility for neural networks is part of why the relevant papers have
+thousands of citations.
+
+Neural networks
+often use non-convex loss functions that have multiple local minima in
+which SGD can get ``lost'' while searching for global minimum.
+Acceleration and momentum (2.3.2) thus are required to achieve effective
+training, particularly on ``deep'' networks with many hidden layers
+(\cite{sutskever2013importance}). Neural networks are
+often extremely complex and high-dimensional and hence require enormous training
+data (\cite{lecun2015deep}); hence, speeding up computation via parallelization
+(2.3.5) or GPU computing is common.
+
 \section{Simulation case study}
@@
 
-If a Silicon Valley press release uses words like ``artificial intelligence'',
+If a Silicon Valley press release uses phrases like ``artificial intelligence'',
 ``machine learning'', and ``big data'' to describe a new product or service, it very likely
-uses SGD in some way. Such is the extent of SGD's popularity. 
+relies upon SGD in some way. Such is the extent of SGD's popularity. 
 Below are some prominent examples of SGD's practical

### 2018-05-11 sgd/sgd.tex
o

@@
 	OLS has a closed form solution: $\hat{\bm{\beta}} =
-	(\bm{X}'\bm{X})\bm{X}'\bm{Y}$ \\~\\
+	(\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ \\~\\
 
@@
 
-	Some familiar tools for finding $\hat{\mathbf{w}}$ no longer work well
+	Some familiar tools for finding $\hat{\mathbf{w}}$ may no longer work well
 	\\~\\
@@
 \begin{frame}
-	\frametitle{Gradient descent is an iterative search procedure}
+	\frametitle{Gradient descent visualized}
 	\begin{figure}[h]
@@
 
-	Stop if gradient $\leq \epsilon$ \\~\\
+	Stop if $||\mathrm{gradient}||_2$ $\leq \epsilon$ \\~\\
 
@@
 	{\mathbf{w}^{(t+1)}} = \mathbf{w}^{(t)} - \gamma \nabla_w
-	L(f(\bm{w}^{(w)}; \mathbf{X}_{\textcolor{red}{i}}), Y_{\textcolor{red}{i})}
+	L(f(\bm{w}^{(t)}; \mathbf{X}_{\textcolor{red}{i}}), Y_{\textcolor{red}{i})}
 	$$ \\~\\
@@
 	\frametitle{SGD is very popular in academia and industry}
-	Many, many, \textit{many} statistical learning models use SGD to fit weights \\~\\
+	Many statistical learning models use SGD to fit weights \\~\\
+
+	Particularly popular for neural networks (backpropagation) \\~\\
 
@@
 	\begin{itemize}
-		\item \small ``Neural networks''
-		\item ``Machine learning''
-		\item ``AI'' 
+		\item \small ``Machine learning''
+		\item \small ``AI'' 
+		\item \small ``Cognitive computing''
 	\end{itemize}
